### 2024-02-01 products/manuscript/Manuscript.qmd
Project Part 1
Worked on the manuscript .qmd file to write the project proposal. The file was renamed to 'Erick_project_Part1.docx'. Edited some README files and deleted some unnecessary content

---
title: "Elemental characterization of personal samples from the HAPIN trial for Source Apportionment"
subtitle: "Project: Part 1"
author: Erick E. Mollinedo
date: '`r format(Sys.Date(), "%B %d, %Y")`'
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
---

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
library(tidyverse)
library(readxl)
```

# Introduction

## General Background Information

Household air pollution (HAP), due to the combustion of solid fuels for cooking and heating, is one of the most prominent environmental health issues in low- and middle-income countries (LMICs).^1^ Exposure to HAP has been linked to more than 2.31 million deaths in 2019 and it is associated with adverse cardiovascular and respiratory health outcomes.^2,3^ Particulate matter with an aerodynamic diameter less than 2.5 micrometers (PM~2.5~) is one of the major air pollutants chemically and physically composed of crustal material, black carbon (BC), organic compounds, metals and others. Toxicity from PM~2.5~ depends, among many factors, on the chemical composition of the particles which heavily depend on the sources of emissions.^4,5^

The identification of point sources of pollution is important for the development of strategies to improve air quality. Source apportionment (SA) is a methodology to reconstruct the impact of emissions from different sources, a method applied to different pollutants such as particulate matter.^6^ The most used SA methods includes receptor models, an example of this is the positive matrix factorization (PMF). One advantage of PMF is that it does not require the source profiles prior to analysis and there is no limit on the number of sources, however, it is required the knowledge of potential source profiles. Another important piece before conducting source apportionment is to correctly process the data (which includes concentrations and uncertainties), and to explore associations to try to identify the potential sources. Just a few studies about SA in LMICs have been published, therefore it is important to characterize chemical composition and potential sources of air pollutants in these settings.^7-9^

## Description of data and data source

The data comes from filter samples of PM~2.5~ from personal exposure measurements conducted in pregnant women from the Household Air Pollution Intervention Network [(HAPIN)](https://www.hapintrial.org/) trial, in Guatemala. In brief, this is a randomized controlled trial, where a liquefied petroleum gas (LPG) stove was delivered as an intervention to half of the pregnant women participants, meanwhile the other half remained cooking using their biomass stove. The intervention was evaluated in terms of reduction of exposures and specific health outcomes by comparing the two study groups (Control and Intervention). The filter samples represent the exposures to PM~2.5~ in a 24-hour period. In total, six hundred and forty eight sample filters are available, where eighteen correspond to field blanks. For these filters, the elemental composition was determined using X-Ray Fluorescence (XRF) to determine the concentrations of 24 elements.

1.  Concentrations:

As seen in the following glimpse from the one of the spreadsheets that contains the concentrations data, each filter sample has its unique ID in `Filter`. There are other 24 variables that list the concentrations of each element. The concentrations are listed in micrograms per square centimeter (ug/cm^2^). Also, each filter is categorized in either two types (Personal sample or Blank) under `Type`.

```{r}
concentration <- read_excel(here("data", "raw-data", "conc_unc_hapingt_emollinedo.xlsx"), sheet = "Concentration")

#Have a 'glimpse' of the data
glimpse(concentration)

#Check which values are under the `Type` variable
unique(concentration$Type)
```

2.  Uncertainties:

The uncertainties are stored on a different tab from the same excel file, but they match to the concentrations for each filter sample (rows) and each element (columns). These values are also stored in ug/cm^2^.

```{r}
uncertainty <- read_excel(here("data", "raw-data", "conc_unc_hapingt_emollinedo.xlsx"), sheet = "Uncertainty")

#Produce a summary of uncertainties by each element
summary(uncertainty)
```

3.  Other variables from the HAPIN study:

Another source of data is a .csv file that contains over 1000 variables collected from the study, associated to a single filter sample or a single participant (depending on the perspective). I am only interested in exploring some of the variables such as "Treatment" and "Timepoint" to categorize the filters according to study arm or type of fuel. "PM~2.5~", "Black Carbon" and "Carbon monoxide" to have a reference of the concentrations of the three air pollutants measured. And other variables that record exposures such as trash burning, vehicle emissions, tobacco smoke among others.

## Questions/Hypotheses to be addressed

One of the objectives of this project is to create a clean output that consists in a file with adjusted concentrations for all samples and each elemental species, and another file with the adjusted uncertainties for the same samples and elemental species. These outputs will be used as inputs for the EPA PMF 5.0 software to try to determine the potential sources that contributed to these samples. However, there are certain questions that I plan to answer:

1.  What elements are correlated among each other from these filter samples?
2.  Is there a difference in concentrations if we categorize the samples by type of fuel and by study arm?
3.  What exposures could explain the variations in the concentrations of the most significant elemental species?

Answering these questions will also be helpful before performing the PMF analysis using the EPA software, to check patterns and associations between elemental species.

# Methods

## Data processing

First, the concentration values will be cleaned by removing unnecessary character such as brackets, and by standardizing in the same numeric notation. Then, all elements where more than 50% of the samples are below the detection limit (DL) will be filtered out. The detection limit for each elemental species is listed in the same excel spreadsheet that contains the raw concentrations and uncertainties. For the remaining elemental species, the concentrations will be adjusted based on the blank filters, and the values below DL will be replaced as their corresponding DL divided by the square root of 2. Finally, the concentrations will be transformed to micrograms per cubic meter (ug/m^3^) operating the concentration by the filter area, divided by the sample volume. Also, the uncertainties will be transformed to ug/m^3^ to match the units of the concentrations. In addition, the black carbon concentrations will be added to add another species to the analysis.

The uncertainties will be processed by first, removing all the elemental species that had more than 50% of their samples below the detection limit, as seen above. Then, the uncertainties will be adjusted based on equations derived from the law of propagation of uncertainty. In brief, the equations consider the individual uncertainties from each one of the variables, in this case the XRF instrument uncertainty, the area of the filter, volume, and the attenuation (in case of the Black carbon measurements).

## Statistical analysis

The correlation (r^2^) between elemental species will be determined using either individual paired correlations or by a correlation plot matrix. Additional plots, such as boxplots, to visualize and compare how the concentrations differ by type of fuel and by study arm. These differences will be assessed for statistical significance. After exploring the distribution that best fits the dataste, the most appropriate model will be implemented to explore what exposure variables are mostly associated with the concentrations of the most significant elemental species.

# References

1.  Gordon, S. B. et al. Respiratory risks from household air pollution in low and middle income countries. The Lancet Respiratory Medicine 2, 823–860 (2014).
2.  Murray, C. J. L. et al. Global burden of 87 risk factors in 204 countries and territories, 1990–2019: a systematic analysis for the Global Burden of Disease Study 2019. The Lancet 396, 1223–1249 (2020).
3.  Pope, C. A. & Dockery, D. W. Health Effects of Fine Particulate Air Pollution: Lines that Connect. Journal of the Air & Waste Management Association 56, 709–742 (2006).
4.  Clasen, T. et al. Design and Rationale of the HAPIN Study: A Multicountry Randomized Controlled Trial to Assess the Effect of Liquefied Petroleum Gas Stove and Continuous Fuel Distribution. Environ Health Perspect 128, 047008 (2020).
5.  Mohammed, M. O. A. et al. Potential Toxicological and Cardiopulmonary Effects of PM2.5 Exposure and Related Mortality: Findings of Recent Studies Published during 2003-2013. Biomed Environ Sci.
6.  Viana, M. et al. Source apportionment of particulate matter in Europe: A review of methods and results. Journal of Aerosol Science 39, 827–849 (2008).
7.  Zhou, Z. et al. Chemical Characterization and Source Apportionment of Household Fine Particulate Matter in Rural, Peri-urban, and Urban West Africa. Environ. Sci. Technol. 48, 1343–1351 (2014).
8.  Sharma, S. K. et al. Source Apportionment of PM2.5 in Delhi, India Using PMF Model. Bull Environ Contam Toxicol 97, 286–293 (2016).
9.  Martínez-Cinco, M., Santos-Guzmán, J. & Mejía-Velázquez, G. Source apportionment of PM 2.5 for supporting control strategies in the Monterrey Metropolitan Area, Mexico. Journal of the Air & Waste Management Association 66, 631–642 (2016).

### 2024-02-23 code/data_cleaning_code.qmd
Worked on Project Part 2
Created the project part 2 document, and two code files for data cleaning and exploratory analyses. Figures and other outputs are also included

---
title: "Data cleaning"
subtitle: "Project: Part 2"
author: Erick E. Mollinedo
date: '`r format(Sys.Date(), "%B %d, %Y")`'
format: html
editor: visual
---

Packages needed for data cleaning and data exploration

```{r}
library(here)
library(readxl)
library(tidyverse)
```

## HAPIN Guatemala data cleaning

First I will load the HAPIN Guatemala dataset that has some of the variables of interest and assign it to the `hapin` dataframe, and I will also load a file that has all the filter IDs that I need to find in the `hapin` dataframe, I'll assign it to the `xrflist` dataframe.

```{r}
hapin <- read_csv(here("data", "raw-data", "HAPIN_EX17_Guatemala_20230728.csv"))
xrflist <- read_csv(here("data", "raw-data", "sa_filters_emollinedo.csv"))
```

Now I will only select the variables I am interested to work with.

```{r}
hapin <- hapin %>% select(trt_blinded, timepoint, h41_m_ecm_fid, h41_b_ecm_fid, h41b_filter1, h41b_filter2, h42_cook, h42_smoke, h42_coil, h42_trash, h42_kero, h42_incence, h42_generator, h42_smoked, h42_crop, h42_smoky_kitc, ECM_grav_neph_conc_M, ECM_bc_conc_M, CO_avg_ppm_M)
```

And now I will select only the observations from `hapin` that correspond to the filter IDs from the `xrflist` dataframe. These filter IDs could be located in either of the following variables: `h41_m_ecm_fid`, `h41_b_ecm_fid`, `h41b_filter1`, `h41b_filter2`, so I searched in these columns. But first I made sure all the filter IDs were in the same format as the IDs from the `xrflist`.

```{r}
#Make sure all filter IDs are upper case, using the `mutate()` function with `toupper()`
hapin <- hapin %>% 
  mutate(h41_m_ecm_fid = toupper(h41_m_ecm_fid),
         h41b_filter1 = toupper(h41b_filter1),
         h41_b_ecm_fid = toupper(h41_b_ecm_fid),
         h41b_filter2 = toupper(h41b_filter2))

#Select all filters that match with the list of filters from XRF, using the `filter()` function.
hapin <- hapin %>%
  filter(h41_m_ecm_fid %in% xrflist$filter_id |
           h41b_filter1 %in% xrflist$filter_id |
           h41_b_ecm_fid %in% xrflist$filter_id |
           h41b_filter2 %in% xrflist$filter_id)
```

Here I will create a new variable called `fueltype`, this variable will be coded as "LPG" or "Biomass" depending on the type of visit and the treatment or intervention `trt_blinded`.

```{r}
#Concatenate `trt_blinded` and `timepoint` into a new temporary variable
hapin$concatenated <- paste(hapin$trt_blinded, hapin$timepoint)

#Create a new variable `fueltype` that categorizes all the observations by type of fuel ('R' means the visit belongs to the 'Control' treatment and 'Q' for the 'Intervention' treatment)
hapin <- mutate(hapin, fueltype = case_when(
  endsWith(concatenated, "BL") ~ "Biomass", #BL corresponds to the baseline, where all the participants had a biomass type fuel stove
  endsWith(concatenated, "R P1") ~ "Biomass",
  endsWith(concatenated, "R P2") ~ "Biomass",
  endsWith(concatenated, "Q P1") ~ "LPG",
  endsWith(concatenated, "Q P2") ~ "LPG",
  endsWith(concatenated, "R BLP1") ~ "Biomass",
  endsWith(concatenated, "R P1P2") ~ "Biomass",
  endsWith(concatenated, "R P2B1") ~ "Biomass",
  endsWith(concatenated, "Q BLP1") ~ "LPG",
  endsWith(concatenated, "Q P1P2") ~ "LPG",
  endsWith(concatenated, "Q P2B1") ~ "LPG"))
```

And now I will rename some of the variables and delete unnecessary ones. I will also recode the values from the newly named `arm` variable.

```{r}
hapin <- hapin %>% rename(filter_id = "h41_m_ecm_fid", dup1_filterid = "h41b_filter1", dup2_filterid = "h41b_filter2",
                         blank_filterid = "h41_b_ecm_fid", arm = "trt_blinded", pm25 = "ECM_grav_neph_conc_M",
                         bc = "ECM_bc_conc_M", co = "CO_avg_ppm_M", stove = "h42_cook", smoke = "h42_smoke",
                         coil = "h42_coil", trash = "h42_trash", kerosene = "h42_kero", incense = "h42_incence",                                   generator = "h42_generator", smoky = "h42_smoked", crop = "h42_crop", stove_other = "h42_smoky_kitc") %>% 
  select(-c(concatenated, timepoint)) %>% 
  mutate(arm = case_when(arm == 'R' ~ 'Control', arm == 'Q' ~ 'Intervention', TRUE ~ arm))
```

Then, I will create four dataframes, one for the personal filters, other two for the duplicates and another one for the blank filters. I will also remove the unnecessary columns from these dataframes.

```{r}
#Creating the dataframe for personal sample filters, and deleting the other columns with unnecessary filter IDs
hapin_personal <- hapin %>% filter(filter_id %in% xrflist$filter_id) %>% 
  select(-c(blank_filterid, dup1_filterid, dup2_filterid))

#Creating the dataframe for duplicates in column 4, and deleting the other columns with unnecessary filter IDs, and renaming `dup1_filterid` to `filter_id`.
hapin_dup1 <- hapin %>% filter(dup1_filterid %in% xrflist$filter_id) %>% 
  select(-c(blank_filterid, filter_id, dup2_filterid)) %>% 
  rename(filter_id = "dup1_filterid")

#Creating the dataframe for duplicates in column 5, and deleting the other columns with unnecessary filter IDs, and renaming `dup2_filterid` to `filter_id`.
hapin_dup2 <- hapin %>% filter(dup2_filterid %in% xrflist$filter_id) %>% 
  select(-c(blank_filterid, filter_id, dup1_filterid)) %>% 
  rename(filter_id = "dup2_filterid")

#Creating the dataframe for blank filters, and deleting the other columns with unnecessary filter IDs, and renaming `blank_filterid` to `filter_id`.
hapin_blanks <- hapin %>% filter(blank_filterid %in% xrflist$filter_id) %>% 
  select(-c(filter_id, dup1_filterid, dup2_filterid)) %>% 
  rename(filter_id = "blank_filterid")
```

And now I will combine the personal and duplicates dataframes into a single one named `hapin_filters`.

```{r}
#Combining the `personal` and `dup` dataframes into a single one, using `bind_rows()`
hapin_filters <- bind_rows(hapin_personal, hapin_dup1, hapin_dup2)
```

Both the personal exposure sample and the blank dataframes are ready to be used for further analysis, so I saved them in RDS format.

```{r}
saveRDS(hapin_filters, file = here("data", "processed-data", "hapin_samples.rds"))
saveRDS(hapin_blanks, file = here("data", "processed-data", "hapin_blanks.rds"))
```

## Chemical species Concentrations data cleaning

The following sections will be long, but necessary to demonstrate how to clean all the data and make all the adjustments before getting the final data for analysis. First I will load the file with the chemical species concentrations, assigned to the `concentration` dataframe. I will also load the limit of detection values from all chemical species under the `lod` dataframe.

```{r}
#Loading the chemical species concentration data
concentration <- read_excel(here("data", "raw-data", "conc_unc_hapingt_emollinedo.xlsx"), sheet = "Concentration")

#Loading the chemical species limit of detection values
lod <- read_excel(here("data", "raw-data", "conc_unc_hapingt_emollinedo.xlsx"), sheet = "DL")

#having a glimpse of this dataframe
glimpse(concentration)
```

First things first, some of the observations contain brackets, which represent values below the detection limit and they have to be corrected in future steps. So first, here I will delete the brackets, then I will convert these variables to numeric type and round them to 4 decimal digits.

```{r}
#Remove brackets, convert to numeric, and round to 4 decimal places
concentration <- concentration %>%
  mutate(across(3:26, ~ round(as.numeric(str_remove_all(., "\\[|\\]")), 4))) %>% 
  rename(filter_id = "Filter")
```

Now, I will remove the filters that correspond to field blanks, since those will be used to adjust the sample filters. I will create a new data frame `blanks`, that contains those observations.

```{r}
#Create dataframe with the concentrations for the blank filters for each element
blanks <- concentration %>% filter(filter_id %in% hapin_blanks$filter_id) %>% 
  select(-c(filter_id, Type))

#Remove blank filters from the `concentration` dataframe
concentration <- concentration %>%
  anti_join(hapin_blanks, by = "filter_id")

#Calculate the median for each element, using the blank filter values
median_blanks <- blanks %>% select(Mg: Na) %>%
  map_dbl(median) %>% round(4)

median_blanks
```

Now, I will estimate which chemical species are representative, by filtering all the species with less than 50% of their values are below the detection limit.

```{r}
#Round the limit of detection values to 4 decimal places
lod <- lod %>% mutate(`Concentration (ug/cm2)` = round(`Concentration (ug/cm2)`, 4))
print(lod)

#Estimate the percentage of values below detection limit for each element
bdl <- concentration %>%
  summarise(
    percent_bdl_mg = sum(Mg < 0.0051) / n() * 100,
    percent_bdl_al = sum(Al < 0.0059) / n() * 100,
    percent_bdl_si = sum(Si < 0.0068) / n() * 100,
    percent_bdl_s = sum(S < 0.0026) / n() * 100,
    percent_bdl_k = sum(K < 0.0040) / n() * 100,
    percent_bdl_ca = sum(Ca < 0.0040) / n() * 100,
    percent_bdl_ti = sum(Ti < 0.0086) / n() * 100,
    percent_bdl_cr = sum(Cr < 0.0055) / n() * 100,
    percent_bdl_mn = sum(Mn < 0.0030) / n() * 100,
    percent_bdl_fe = sum(Fe < 0.0043) / n() * 100,
    percent_bdl_ni = sum(Ni < 0.0043) / n() * 100,
    percent_bdl_cu = sum(Cu < 0.0080) / n() * 100,
    percent_bdl_zn = sum(Zn < 0.0024) / n() * 100,
    percent_bdl_ga = sum(Ga < 0.0049) / n() * 100,
    percent_bdl_as = sum(As < 0.0062) / n() * 100,
    percent_bdl_se = sum(Se < 0.0181) / n() * 100,
    percent_bdl_cd = sum(Cd < 0.0932) / n() * 100,
    percent_bdl_in = sum(In < 0.1089) / n() * 100,
    percent_bdl_sn = sum(Sn < 0.1844) / n() * 100,
    percent_bdl_te = sum(Te < 0.2166) / n() * 100,
    percent_bdl_i = sum(I < 0.3149) / n() * 100,
    percent_bdl_pb = sum(Pb < 0.0324) / n() * 100,)

print(bdl)
```

Only 10 elements have less than 50% of their observations below the detection limit. A note: Na and Cl were seen in high and very variable concentrations, so those elements will also be removed from the analysis. Now, I will filter the remaining elements (variables).

```{r}
#Select only the filter ID, type and the 10 elements that have more than 50% of their data not below the detection limit
concentration <- concentration %>% select(c(filter_id, Type, Mg, Al, Si, S, K, Ca, Ti, Mn, Fe, Zn)) %>% 
  arrange(filter_id) #Also I decided to arrange them ascending based on the filter_id
```

Now, blank adjusting all the values, by substracting the median value for each element, obtained from the blank filters. Also, replace all values below the detection limit with (LOD/sqrt(2)).

```{r}
#Substract the median blank value for the elements that need blank adjustment
concentration <- concentration %>%
  mutate(Mg = Mg - 0.0112, S = S - 0.0044, K = K - 0.0014, Ca = Ca - 0.0048, Fe = Fe - 0.0038, Zn = Zn - 0.0045)

#Replace all values below the detection limit with DL/sqrt(2)
concentration <- concentration %>% 
    mutate(Mg = round(if_else(Mg < 0.0051, 0.0051 / sqrt(2), Mg), 4), Al = round(if_else(Al < 0.0059, 0.0059 / sqrt(2), Al), 4),
           Si = round(if_else(Si < 0.0068, 0.0068 / sqrt(2), Si), 4), S = round(if_else(S < 0.0026, 0.0026 / sqrt(2), S), 4),
           K = round(if_else(K < 0.0016, 0.0016 / sqrt(2), K), 4), Ca = round(if_else(Ca < 0.0040, 0.0040 / sqrt(2), Ca), 4),
           Ti = round(if_else(Ti < 0.0086, 0.0086 / sqrt(2), Ti), 4), Mn = round(if_else(Mn < 0.0030, 0.0030 / sqrt(2), Mn), 4),
           Fe = round(if_else(Fe < 0.0043, 0.0043 / sqrt(2), Fe), 4), Zn = round(if_else(Zn < 0.0024, 0.0024 / sqrt(2), Zn), 4))

#I will save these concentrations in a new dataframe `concentrations_ugcm2` that will be used for uncertainty estimations in future steps
concentrations_ugcm2 <- concentration %>% select(-Type)
```

And then, converting the concentrations from micrograms per square centimeter to micrograms per cubic meter using the equation: ug/m3 = (Concentration (ug/cm^2^) x Filter area (cm^2^))/Sample volume (m3). Where, the filter area is 0.7853 cm^2^ and the sample volume is 0.43 m^3^. Also, I need to input the BC concentrations from the `hapin_filters` dataframe.

```{r}
#Converting the concentrations from (ug/cm2) to (ug/m3)
concentration <- concentration %>%
  mutate(across(Mg:Zn, ~ round(.x * 0.7853 / 0.43, 4)))

#Now adding the BC concentrations value from the `hapin_filters` df
hapin_filters <- hapin_filters %>% arrange(filter_id) #First arrange in ascending order by filter_id
concentration <- bind_cols(concentration, hapin_filters$bc) %>% rename(BC = "...13") %>% select(-Type) #Join the BC concentrations to the `concentrations` df, also renaming the column to "BC" and deleting the 'Type' column
```

I will add the fueltype and arm variables to this dataframe, and then I'll save it in RDS format.

```{r}
#Add the `arm` and `fueltype` variables to the `concentration` dataframe
concentration <- concentration %>%
  mutate(arm = hapin_filters$arm, fueltype = hapin_filters$fueltype)

#Save this dataframe as RDS file
saveRDS(concentration, file = here("data", "processed-data", "concentration.rds"))
```

## Chemical species Uncertainties data processing

Now, it's time to make some adjustments to the uncertainties from all filters across all chemical species. First, loading the uncertainties data to the `uncertainty` data frame.

```{r}
uncertainty <- read_excel(here("data", "raw-data", "conc_unc_hapingt_emollinedo.xlsx"), sheet = "Uncertainty")
```

First, I will remove the chemical species with more than 50% of their values below detection limit, also remove the blank filter observations, sort by Filter ID and round the values to numbers with four decimal places.

```{r}
uncertainty <- uncertainty %>% select(c(Filter, Mg, Al, Si, S, K, Ca, Ti, Mn, Fe, Zn)) %>% #Select only the necessary variables
  rename(filter_id = "Filter") %>% #Rename the Filter column
  mutate_at(vars(Mg:Zn), ~ round(., 4)) %>% #Round the uncertainties to 4 decimal places
  arrange(filter_id) %>% #Arrange observations based on Filter ID
  anti_join(hapin_blanks, by = "filter_id") #Remove blank filters

#Have a glimpse of the data
glimpse(uncertainty)
```

Now, the uncertainties from the concentration values that were LOD adjusted have to be reestimated to (5/6)\*LOD. To do this I created a list that inputs the LOD values for each element, and then using `mutate()`, I make the changes only to the values that follow the specified condition.

```{r}
#Create a list for the detection limits for each element
dl <- c(Mg = 0.0051, Al = 0.0059, Si = 0.0068, S = 0.0026, K = 0.0016, 
        Ca = 0.0040, Ti = 0.0086, Mn = 0.0030, Fe = 0.0043, Zn = 0.0024)

# Recalculate the values in df based on the conditions
uncertainty <- uncertainty %>%
  mutate(across(all_of(names(dl)),
                ~ round(if_else(concentrations_ugcm2[[cur_column()]] < dl[[cur_column()]], (5/6) * dl[[cur_column()]], .), 4),
                .names = "{.col}"))
```

Finally, for all the chemical species, the uncertainties have to be reestimated so they correctly reflect the uncertainty for each variable used in the calculations. In this case, input the XRF-estimated concentration, filter area and filter volume uncertainties using the law of propagation of uncertainties. Also, the purpose of this section is to convert the uncertainties from ug/cm2 to ug/m3 to correctly reflect the units of the concentrations.

```{r}
# Define the variables to be recalculated
variables <- c('Mg', 'Al', 'Si', 'S', 'K', 'Ca', 'Ti', 'Mn', 'Fe', 'Zn')

# Recalculate the values using the given equation
uncertainty <- uncertainty %>%
  mutate(across(all_of(variables), ~ round( #Here, selecting all the variables to be re-calculated
    sqrt((0.7853/0.43)^2 * .^2 + (concentrations_ugcm2[[cur_column()]]/0.43)^2 * 0.05^2 + (-concentrations_ugcm2[[cur_column()]]*0.7853/0.43^2)^2 * (0.05 * 0.43)^2), 4), .names = "{.col}")) #Input the equation
```

Since the BC concentrations will also be included, I need to estimate the BC uncertainties. These estimations require an additional file that will be assigned to the `atn` dataframe.

```{r}
#Load file to estimate BC uncertainties
atn <- read_csv(here("data", "raw-data", "ATN_summary_SA.csv"))

#Explore the structure of the file
glimpse(atn)
```

To estimate the BC concentrations, I used an instrument called Sootscan, that calculates transmissometry as attenuation before (pre) and after (post) sampling the filters. These values summarize the pre- and post- ATN values for all the filters, but I have to estimate the total attenuation for all sample filters with the equation: LogN(post-ATN/pre-ATN)\*100.

```{r}
#Estimate total attenuation by sample
atn <- atn %>% mutate(ATN = round(log(post_ATN/pre_ATN) *100, 2))
```

And finally, estimate the BC uncertainties using an equation also derived from the lag of propagation of uncertainties.

```{r}
#Estimate the BC uncertainties
uncertainty <- uncertainty %>% 
  mutate(BC = round((1370 * sqrt((atn$ATN)^2 * 0.000005^2 + 0.000079^2 * 1.16^2)), 4))
```

I will also add the arm and fueltype variables to this dataframe, and I'll save it as RDS format.

```{r}
#Add the `arm` and `fueltype` variables to the `uncertainty` dataframe
uncertainty <- uncertainty %>%
  mutate(arm = hapin_filters$arm, fueltype = hapin_filters$fueltype)

#Save this dataframe as RDS file
saveRDS(uncertainty, file = here("data", "processed-data", "uncertainty.rds"))
```

### 2024-02-23 code/data_exploring_code.qmd
Worked on Project Part 2
Created the project part 2 document, and two code files for data cleaning and exploratory analyses. Figures and other outputs are also included

---
title: "Project Data Exploring"
subtitle: "Project: Part 2"
author: Erick E. Mollinedo
date: '`r format(Sys.Date(), "%B %d, %Y")`'
format: html
editor: visual
---

## Concentrations

```{r}
#Load packages to use
library(readr)
library(here)
library(vtable)
library(corrplot)
library(tidyverse)
library(RColorBrewer)
library(naniar)
```

Load the `concentration` file.

```{r}
concentration <- read_rds(here("data", "processed-data", "concentration.rds"))
```

I will produce two boxplots, that summarize the concentrations categorized by `arm` and `fueltype`. First I will create a dataframe in longer format using `pivot_longer()`. Then creating the plots using `ggplot()`.

```{r}
#Pivot longer the `concentration` df to make the boxplots
conc_longer <- concentration %>% #First a df with the `arm` variable
  pivot_longer(cols = -c(filter_id, fueltype, arm),
               names_to = "species", values_to = "concentration")

#Creating the boxplot of concentrations by fueltype
options(scipen = 999) #to remove scientific notation for smaller values (i.e. changes from 1*10-3 to 0.001)
ggplot(conc_longer)+
  geom_boxplot(aes(x= species, y= concentration, fill= fueltype))+
  scale_fill_brewer(palette = "Paired")+
  labs(x= "Elemental species", y= "Concentration (ug/m3)")+
  scale_y_log10()+
  guides(fill= guide_legend(title= "Fuel type"))+ #Change legend from fuel_type to Fuel type
  theme_bw(base_size = 15)

#Creating the boxplot of concentrations by arm
options(scipen = 999) #to remove scientific notation for smaller values (i.e. changes from 1*10-3 to 0.001)
ggplot(conc_longer)+
  geom_boxplot(aes(x= species, y= concentration, fill= arm))+
  scale_fill_brewer(palette = "Paired")+
  labs(x= "Elemental species", y= "Concentration (ug/m3)")+
  scale_y_log10()+
  guides(fill= guide_legend(title= "Study Arm"))+ #Change legend from fuel_type to Fuel type
  theme_bw(base_size = 15)
```

The boxplots for `fueltype` and `arm` show what I was expecting in terms of how the data is distributed. The concentrations seem higher in some elemental especies such as BC, Ca, K, Mg, Mn and S. The following steps will be to evaluate if those differences in concentrations are statistically different.

Now I will create a correlation plot using the `corrplot()` function, to explore which elements could be more associated between each other.

```{r}
#First, delete unnecessary variables to conduct the correlation test
cor <- concentration %>% select(-c(filter_id, fueltype, arm))

#Perform the spearman correlation test
cor <- stats::cor(cor, method = "pearson", use = "complete.obs") #I had to input the `use=` argument since there are missing values in the BC variable. I will correct them later.

#Create the correlation plot
corrplot(cor, method = "color", type = "lower", #Color form and displays at the lower portion
         col = COL2("RdYlBu", 20), order = "hclust", #Color palette, and order or the species
         tl.col = "orangered4", tl.srt = 0, tl.cex = 0.9,  #Color of the axis, position and size
         addCoef.col = "gray10", number.font = 2, number.cex = 0.8, #Display correlation number, set the font and size
         col.lim = c(0, 1), is.corr = T) #Set the color limits at the bar
```

The correlation plot above shows that Si and Fe are strongly correlated, which might suggest that these elements represent one specific source of pollution. Other high correlations observed are between Al and Si, and Fe and Ti.

## Uncertainties

Load the `uncertainty` file.

```{r}
uncertainty <- read_rds(here("data", "processed-data", "uncertainty.rds"))
```

I will produce a summary table of the uncertainties, using the `vtable` package.

```{r}
#Remove the `arm` variable
unc_fuel <- uncertainty %>% select(-arm)

#Use the `sumtable()` function from the `vtable` package, which creates a table
sumtable(unc_fuel, group = 'fueltype', group.long = T, add.median = T, out = 'kable')

#Remove the `fueltype` variable
unc_arm <- uncertainty %>% select(-fueltype)

#Use the `sumtable()` function from the `vtable` package, which creates a table
sumtable(unc_arm, group = 'arm', group.long = T, add.median = T, out = 'kable')

```

The tables above are just a way to represent the summary statistics of the uncertainties, however, the uncertainties will only be useful for the PMF analysis after the conclusion of this project. One thing noticed also from this table, is that there are some infinite values from the BC concentrations, which also show that there are missing values that have to be corrected in earlier steps. From this tables we can also see that there are 376 observations from the 'Biomass' and 254 from the 'LPG' fueltypes. Meanwhile, there are 323 observations from the 'Control' arm and 307 from the 'Intervention' arm.

## HAPIN variables

Loading the `hapin_samples` file.

```{r}
hapin <- read_rds(here("data", "processed-data", "hapin_samples.rds"))
```

First, check if there are missing values for any of the observations using the `gg_miss_var()` function.

```{r}
gg_miss_var(hapin)
```

As seen above, there are some missing values from the observational exposures (kerosene to coil), there is just one missing observation for PM2.5 and BC concentrations and there is more than 30 missing observations for CO (carbon monoxide). The missing observation of BC might explain the missing value spotted before, so this observation might actually have to be dropped down.

Now I will create a barplot showing the distribution of responses for the categorical variables `kerosene`, `trash`, `stove_other`, `stove`, `smoky`, `incense`, `generator`, `smoke`, `crop`, and `coil`. First doing data manipulation so the df is useful for plotting.

```{r}
#Pivot longer the `hapin` df to make a barplot
hapin_longer <- hapin %>%
  select(c(`kerosene`, `trash`, `stove_other`, `stove`, `smoky`, `incense`, `generator`, `smoke`, `crop`, `coil`)) %>% #Here selecting only the necessary variables
  pivot_longer(cols = everything(),
               names_to = "exposure", values_to = "response") #Create the longer df

#Create a simple barplot
ggplot(hapin_longer, aes(x= exposure, fill= response))+
  geom_bar(position = "dodge")+
  theme_classic()
```

The bar plot shows that the majority of participants recorded to not being exposed to most of the exposures, except to `stove`, which is the stove that the participants use. Others also recorded that were exposed to other stoves, or to trash burning.

And finally plotting the PM2.5, BC and CO concentrations in different graphs using `ggplot()`

```{r}
#Histogram plot for the PM2.5 concentrations
ggplot(hapin, aes(x= pm25, fill= fueltype))+
  geom_histogram()+
  theme_classic()+
  labs(x= "PM2.5 Concentration (ug/m3)")

#Histogram plot for the BC concentrations
ggplot(hapin, aes(x= bc, fill= fueltype))+
  geom_histogram()+
  theme_classic()+
  labs(x= "Black Carbon Concentration (ug/m3)")

#Histogram plot for the CO concentrations
ggplot(hapin, aes(x= co, fill= fueltype))+
  geom_histogram()+
  theme_classic()+
  labs(x= "Carbon Monoxide Concentration (ppm)")
```

The histograms for the concentration of the three pollutants show that they don't follow a normal distribution. Most of the concentrations are low, and it is also seen that overall the concentrations in the LPG fueltype are lower than the Biomass group. If these variables will be used for modeling purposes, they should be log-transformed or using a logistic regression approach.

### 2024-02-23 products/manuscript/Manuscript.qmd
Worked on Project Part 2
Created the project part 2 document, and two code files for data cleaning and exploratory analyses. Figures and other outputs are also included

@@
 ---
-title: "Elemental characterization of personal samples from the HAPIN trial for Source Apportionment"
-subtitle: "Project: Part 1"
+title: "Elemental characterization of personal filter samples from the HAPIN trial for Source Apportionment"
+subtitle: "Project: Part 2"
 author: Erick E. Mollinedo
@@
 library(knitr)
+library(readr)
 library(tidyverse)
-library(readxl)
 ```
@@
 
-## General Background Information
+## Background
 
-Household air pollution (HAP), due to the combustion of solid fuels for cooking and heating, is one of the most prominent environmental health issues in low- and middle-income countries (LMICs).^1^ Exposure to HAP has been linked to more than 2.31 million deaths in 2019 and it is associated with adverse cardiovascular and respiratory health outcomes.^2,3^ Particulate matter with an aerodynamic diameter less than 2.5 micrometers (PM~2.5~) is one of the major air pollutants chemically and physically composed of crustal material, black carbon (BC), organic compounds, metals and others.
-Toxicity from PM~2.5~ depends, among many factors, on the chemical composition of the particles which heavily depend on the sources of emissions.^4,5^
+Household air pollution (HAP), due to the combustion of solid fuels for cooking and heating, is one of the most prominent environmental health issues in low- and middle-income countries (LMICs).^1^ HAP has been listed as the second environmental health risk factor worldwide, just after outdoor air pollution.^2^ Exposure to HAP has been linked to more than 2.31 million deaths in 2019 and it is associated with adverse cardiovascular and respiratory health outcomes.^3,4^ Particulate matter with an aerodynamic diameter less than 2.5 micrometers (PM~2.5~) is one of the major air pollutants chemically and physically composed of crustal material, black carbon (BC), organic compounds, metals and others.
+Toxicity from PM~2.5~ depends, among many factors, on the chemical composition of the particles which heavily depend on the sources of emissions.^5,6^
 
 The identification of point sources of pollution is important for the development of strategies to improve air quality.
-Source apportionment (SA) is a methodology to reconstruct the impact of emissions from different sources, a method applied to different pollutants such as particulate matter.^6^ The most used SA methods includes receptor models, an example of this is the positive matrix factorization (PMF).
+Source apportionment (SA) is a methodology to reconstruct the impact of emissions from different sources, a method applied to different pollutants such as particulate matter.^7^ The most used SA methods includes receptor models, an example of this is the positive matrix factorization (PMF).
 One advantage of PMF is that it does not require the source profiles prior to analysis and there is no limit on the number of sources, however, it is required the knowledge of potential source profiles.
 Another important piece before conducting source apportionment is to correctly process the data (which includes concentrations and uncertainties), and to explore associations to try to identify the potential sources.
-Just a few studies about SA in LMICs have been published, therefore it is important to characterize chemical composition and potential sources of air pollutants in these settings.^7-9^
+Just a few studies about SA in LMICs have been published, therefore it is important to characterize chemical composition and potential sources of air pollutants in these settings.^8-10^
 
-## Description of data and data source
+The Household Air Pollution Intervention Network (HAPIN) trial was a randomized controlled trial where a liquefied petroleum gas (LPG) stove was delivered as an intervention to half of the pregnant women participants, meanwhile the other half remained cooking using their biomass stove.
+The intervention was evaluated in terms of reduction of exposures and specific health outcomes by comparing the two study groups (Control and Intervention).
+Baseline measurements were taken to the participants, and then followed-up for about 18 months, conducting exposure measurements three times during pregnancy and three times after the birth of the child.
+A description of the HAPIN trial and the methods for exposure sampling can be found elsewhere.^11,12^ A source apportionment pilot study from HAPIN was conducted in 2022.
+This study was conducted in 64 and 59 personal exposure filter samples from the pilot phase in Guatemala and Rwanda, respectively.
+Twenty-two chemical species were analyzed in all samples using X-ray fluorescence (XRF), a non-destructive spectrometry analytical method, but only 12 species were detected at both study sites in comparable levels.
+Based on the chemical species detected, four potential sources of PM~2.5~ were identified in both sites.
+One study caveat is the small sample size and limited assessment of chemical composition and sources between study arms, so further studies to complement the findings are needed.^13^
 
-The data comes from filter samples of PM~2.5~ from personal exposure measurements conducted in pregnant women from the Household Air Pollution Intervention Network [(HAPIN)](https://www.hapintrial.org/) trial, in Guatemala.
-In brief, this is a randomized controlled trial, where a liquefied petroleum gas (LPG) stove was delivered as an intervention to half of the pregnant women participants, meanwhile the other half remained cooking using their biomass stove.
-The intervention was evaluated in terms of reduction of exposures and specific health outcomes by comparing the two study groups (Control and Intervention).
-The filter samples represent the exposures to PM~2.5~ in a 24-hour period.
-In total, six hundred and forty eight sample filters are available, where eighteen correspond to field blanks.
-For these filters, the elemental composition was determined using X-Ray Fluorescence (XRF) to determine the concentrations of 24 elements.
-
-1.  Concentrations:
-
-As seen in the following glimpse from the one of the spreadsheets that contains the concentrations data, each filter sample has its unique ID in `Filter`.
-There are other 24 variables that list the concentrations of each element.
-The concentrations are listed in micrograms per square centimeter (ug/cm^2^).
-Also, each filter is categorized in either two types (Personal sample or Blank) under `Type`.
-
-```{r}
-concentration <- read_excel(here("data", "raw-data", "conc_unc_hapingt_emollinedo.xlsx"), sheet = "Concentration")
-
-#Have a 'glimpse' of the data
-glimpse(concentration)
-
-#Check which values are under the `Type` variable
-unique(concentration$Type)
-```
-
-2.  Uncertainties:
-
-The uncertainties are stored on a different tab from the same excel file, but they match to the concentrations for each filter sample (rows) and each element (columns).
-These values are also stored in ug/cm^2^.
-
-```{r}
-uncertainty <- read_excel(here("data", "raw-data", "conc_unc_hapingt_emollinedo.xlsx"), sheet = "Uncertainty")
-
-#Produce a summary of uncertainties by each element
-summary(uncertainty)
-```
-
-3.  Other variables from the HAPIN study:
-
-Another source of data is a .csv file that contains over 1000 variables collected from the study, associated to a single filter sample or a single participant (depending on the perspective).
-I am only interested in exploring some of the variables such as "Treatment" and "Timepoint" to categorize the filters according to study arm or type of fuel.
-"PM~2.5~", "Black Carbon" and "Carbon monoxide" to have a reference of the concentrations of the three air pollutants measured.
-And other variables that record exposures such as trash burning, vehicle emissions, tobacco smoke among others.
-
-## Questions/Hypotheses to be addressed
+## Aims and Hypothesis questions
 
@@
 These outputs will be used as inputs for the EPA PMF 5.0 software to try to determine the potential sources that contributed to these samples.
-However, there are certain questions that I plan to answer:
+However, the questions I plan to answer with the data analysis of this project is:
 
@@
 
-Answering these questions will also be helpful before performing the PMF analysis using the EPA software, to check patterns and associations between elemental species.
-
 # Methods
 
-## Data processing
+## Data collection
 
-First, the concentration values will be cleaned by removing unnecessary character such as brackets, and by standardizing in the same numeric notation.
-Then, all elements where more than 50% of the samples are below the detection limit (DL) will be filtered out.
-The detection limit for each elemental species is listed in the same excel spreadsheet that contains the raw concentrations and uncertainties.
-For the remaining elemental species, the concentrations will be adjusted based on the blank filters, and the values below DL will be replaced as their corresponding DL divided by the square root of 2.
-Finally, the concentrations will be transformed to micrograms per cubic meter (ug/m^3^) operating the concentration by the filter area, divided by the sample volume.
-Also, the uncertainties will be transformed to ug/m^3^ to match the units of the concentrations.
-In addition, the black carbon concentrations will be added to add another species to the analysis.
+The main exposure data comes from filter samples of personal exposure measurements in pregnant women participants from the HAPIN trial, in Guatemala.
+In total, six hundred and forty eight sample filters are available, where eighteen correspond to field blanks.
+From the HAPIN trial study design, in total, six exposure measurements (visits) were conducted during pregnancy in around 800 participants, but for this scenario, only 648 filter samples were randomly selected for chemical species characterization.
+Other exposures where collected as categorical variables using standardized questionnaires administered in each visit.
+These exposures capture some of the sources of exposure such as tobacco smoke, trash burning, use on incense or coil, kerosene and others.
 
-The uncertainties will be processed by first, removing all the elemental species that had more than 50% of their samples below the detection limit, as seen above.
-Then, the uncertainties will be adjusted based on equations derived from the law of propagation of uncertainty.
+## Concentration estimates
+
+The average 24-hour PM~2.5~ concentrations were calculated using the following equation:
+
+$$PM_{2.5} = M/V$$ Where, M is the mass deposition of particles in the filter and V is the volume of air the sampler used to collect the particles.
+
+The black carbon 24-hour concentrations were calculated using the following equation:
+
+$$BC = \frac{ln(I_{f}/I_{0}) * A/V * 10^6} \sigma $$ Where, I~f~ is the post-attenuation and I~0~ is the pre-attenuation, A is the area of the filter, V is the volume of air the sampler used to collect the particles and $\sigma$ is the mass absorption coefficient (a constant).
+
+The concentration of the chemical species (elements) on the filters was determined using X-Ray Fluorescence (XRF), and the uncertainties were estimated also based on the XRF instrument.
+
+## Concentration and Uncertainties data processing
+
+First, all chemical species where more than 50% of the samples are below the detection limit (LOD) were filtered out, leaving a total of 10 chemical species.
+For these remaining species, the concentrations were adjusted based on the blank filters, and the values below LOD were replaced as their corresponding LOD divided by the square root of 2.
+Finally, the concentrations were transformed to micrograms per cubic meter (ug/m^3^).
+Also, the uncertainties were transformed to ug/m^3^ to match the units of the concentrations.
+In addition, the black carbon concentrations were added to add another chemical species to the analysis.
+
+The uncertainties were processed by, first removing all the chemical species that had more than 50% of their samples below the detection limit, as seen above.
+Then, the uncertainties were adjusted based on equations derived from the law of propagation of uncertainty.
 In brief, the equations consider the individual uncertainties from each one of the variables, in this case the XRF instrument uncertainty, the area of the filter, volume, and the attenuation (in case of the Black carbon measurements).
@@
 
-The correlation (r^2^) between elemental species will be determined using either individual paired correlations or by a correlation plot matrix.
+The correlation (r^2^) between chemical species will be determined using either individual paired correlations or by a correlation plot matrix.
 Additional plots, such as boxplots, to visualize and compare how the concentrations differ by type of fuel and by study arm.
@@
 
+# Preliminary Results
+
+## Description of data
+
+1.  Concentrations:
+
+As seen in the following glimpse from the cleaned `concentration` dataframe, it contains the ID of the 630 sample filters, along with the concentrations of 11 chemical species (Mg, Al, Si, S, K, Ca, Ti, Mn, Fe, Zn, and BC), and to which Arm and Fueltype they belong.
+
+```{r}
+concentration <- read_rds(here("data", "processed-data", "concentration.rds"))
+
+glimpse(concentration)
+```
+
+2.  Uncertainties:
+
+The cleaned `uncertainty` dataframe contains the uncertainty of the 11 chemical species. It contains the same name of variables as the `concentration` dataframe.
+
+```{r}
+uncertainty <- read_rds(here("data", "processed-data", "uncertainty.rds"))
+
+glimpse(uncertainty)
+```
+
+3.  Other variables from the HAPIN study:
+
+The cleaned `hapin_samples` dataframe contains information of the PM~2.5~, BC and CO (carbon monoxide) concentrations, besides other important recorded exposures, as seen from the glimpse below.
+It also contains the filter IDs, the arm and fueltype.
+
+```{r}
+hapin_samples <- read_rds(here("data", "processed-data", "hapin_samples.rds"))
+
+glimpse(hapin_samples)
+```
+
+## Exploratory data analysis
+
+### Chemical species concentrations
+
+The concentration of the chemical species are distributed in different ranges depending on their type.
+For example, the highest concentrations were determined to be for Black carbon (BC), with values from 1 to 100 ug/m^3^.
+The remaining chemical species have lower concentrations, ranging from less than 0.1 ug/m^3^ to 10 ug/m^3^.
+Other noticeable patterns is that some of the species were found in higher concentrations for personal samples from participants that used biomass stoves, compared to LPG stoves (@fig-concentrations), such as BC, K, and Ca, and similar patterns were observed when categorized by intervention arm (Supplementary material).
+The following step will be to determine which of these chemical species have statistically significant differences among both comparison groups.
+
+In terms of correlations, Spearman correlation was estimated between all the chemical species and @fig-correlations, shows the associations highlighting the R^2^ value.
+It is observed that the highest correlation is observed between Si and Fe, Ti-Fe, and Al-Si.
+These values could indicate that these species come from the same source, so it will probably be easier to identify using the PMF software after the conclusion of this project.
+
+It is important to remember that the concentrations of the chemical species were recalculated as referenced in the `data_cleaning_code.qmd` file, alongside with the uncertainties.
+A summary statistics table of the uncertainties can be found in the `data_cleaning_code.qmd` code.
+
+![Chemical species concentrations by type of fuel](../../results/figures/boxplot-fueltype.jpg){#fig-concentrations}
+
+![Correlation plot for chemical species](../../results/figures/corplot.jpg){#fig-correlations}
+
+### Other Exposure variables
+
+From the `hapin_samples` dataframe, it was observed that there are some missing values for most of the exposure variables according to the Missing variables plot (@fig-missing).
+The variable with more missing values (\>30) was the CO concentrations.
+Given that this variable had the highest number of missing concentrations, it may not be ideal to include it as part of the final analysis.
+Other missing values are from the recorded exposures such as kerosene, trash burning, stove use, external stoves ('stove_other'), and others.
+It was also observed that there is one observation without BC concentration.
+This observation could be dropped out since the PMF analysis after the conclusion of this project requires all the concentrations and uncertainties to be available (No NA/missing values).
+
+The bar plot (@fig-categorical), shows that most of the participants recorded little exposures, with the exception of stove use at home ('stove'), stoves from other homes ('stove_other') and trash burning ('trash').
+Having not so many recorded exposures other than the previously mentioned, could make the analysis difficult because of sample size or any other statistical adjustments will have to occur before applying a regression model analysis approach.
+
+@fig-pm shows the distribution of PM~2.5~ concentrations by type of fuel.
+It is observed that the concentrations do not follow a normal distribution and it seems like, as expected, the LPG stove users samples have lower concentrations than the biomass stove users.
+The similar trends are observed for the CO and BC concentrations (Supplementary material).
+
+![Plot for missing values for the `hapin_samples` dataframe](../../results/figures/missing-values.jpg){#fig-missing}
+
+![Distribution of responses by the Categorical variables](../../results/figures/cat-exposures.jpg){#fig-categorical}
+
+![PM~2.5~ Concentrations](../../results/figures/pm25.jpg){#fig-pm}
+
+{{< pagebreak >}}
+
 # References
@@
 1.  Gordon, S. B. et al. Respiratory risks from household air pollution in low and middle income countries. The Lancet Respiratory Medicine 2, 823–860 (2014).
-2.
+2.  GBD. Global Burden of Disease Study 2019 (GBD 2019) Results. http://ghdx.healthdata.org/gbd-results-tool (2020).
+3.
 Murray, C.
@@
 The Lancet 396, 1223–1249 (2020).
-3.  Pope, C. A. & Dockery, D. W. Health Effects of Fine Particulate Air Pollution: Lines that Connect. Journal of the Air & Waste Management Association 56, 709–742 (2006).
-4.
+4.  Pope, C. A. & Dockery, D. W. Health Effects of Fine Particulate Air Pollution: Lines that Connect. Journal of the Air & Waste Management Association 56, 709–742 (2006).
+5.
 Clasen, T.
@@
 Environ Health Perspect 128, 047008 (2020).
-5.  Mohammed, M. O. A. et al. Potential Toxicological and Cardiopulmonary Effects of PM2.5 Exposure and Related Mortality: Findings of Recent Studies Published during 2003-2013. Biomed Environ Sci.
-6.  Viana, M. et al. Source apportionment of particulate matter in Europe: A review of methods and results. Journal of Aerosol Science 39, 827–849 (2008).
-7.  Zhou, Z. et al. Chemical Characterization and Source Apportionment of Household Fine Particulate Matter in Rural, Peri-urban, and Urban West Africa. Environ. Sci. Technol. 48, 1343–1351 (2014).
-8.  Sharma, S. K. et al. Source Apportionment of PM2.5 in Delhi, India Using PMF Model. Bull Environ Contam Toxicol 97, 286–293 (2016).
-9.
+6.  Mohammed, M. O. A. et al. Potential Toxicological and Cardiopulmonary Effects of PM2.5 Exposure and Related Mortality: Findings of Recent Studies Published during 2003-2013. Biomed Environ Sci.
+7.  Viana, M. et al. Source apportionment of particulate matter in Europe: A review of methods and results. Journal of Aerosol Science 39, 827–849 (2008).
+8.  Zhou, Z. et al. Chemical Characterization and Source Apportionment of Household Fine Particulate Matter in Rural, Peri-urban, and Urban West Africa. Environ. Sci. Technol. 48, 1343–1351 (2014).
+9.  Sharma, S. K. et al. Source Apportionment of PM2.5 in Delhi, India Using PMF Model. Bull Environ Contam Toxicol 97, 286–293 (2016).
+10.
 Martínez-Cinco, M., Santos-Guzmán, J.
@@
 Journal of the Air & Waste Management Association 66, 631–642 (2016).
+11.
+Clasen, T., Checkley, W., Peel, J.
+L., Balakrishnan, K., McCracken, J.
+P., Rosa, G., Thompson, L.
+M., Barr, D.
+B., Clark, M.
+L., Johnson, M.
+A., Waller, L.
+A., Jaacks, L.
+M., Steenland, K., Miranda, J.
+J., Chang, H.
+H., Kim, D.-Y., McCollum, E.
+D., Davila-Roman, V.
+G., Papageorghiou, A., … HAPIN Investigators.
+Design and Rationale of the HAPIN Study: A Multicountry Randomized Controlled Trial to Assess the Effect of Liquefied Petroleum Gas Stove and Continuous Fuel Distribution.
+Environmental Health Perspectives, 128(4), 047008.
+(2020).
+12.
+Johnson, M.
+A., Steenland, K., Piedrahita, R., Clark, M.
+L., Pillarisetti, A., Balakrishnan, K., Peel, J.
+L., Naeher, L.
+P., Liao, J., Wilson, D., Sarnat, J., Underhill, L.
+J., Burrowes, V., McCracken, J.
+P., Rosa, G., Rosenthal, J., Sambandam, S., de Leon, O., Kirby, M.
+A., … HAPIN Investigators.
+Air Pollutant Exposure and Stove Use Assessment Methods for the Household Air Pollution Intervention Network (HAPIN) Trial.
+Environmental Health Perspectives, 128(4), 047009.
+(2020).
+13. Kearns, K. A. Exposures to household air pollution in women across four diverse countries as part of a multi-country cookstove intervention trial. (2022).
+
+{{< pagebreak >}}
+
+# Supplementary material
+
+![Chemical species concentrations by study arm](../../results/figures/boxplot-arm.jpg)
+
+![Carbon monoxide (CO) Concentrations](../../results/figures/co.jpg)
+
+![Black carbon (BC) Concentrations](../../results/figures/bc.jpg)

### 2024-03-15 code/data_analysis_code.qmd
Project Part 3
Created a new QMD file: data_analysis_code, and a Supplementary Material word document. Also, the manuscript file has been updated to include the preliminary results. ReadMe files updated and figures/tables edited.

---
title: "Project Part 3"
subtitle: "Data analysis code"
author: Erick E. Mollinedo
date: '`r format(Sys.Date(), "%B %d, %Y")`'
format: html
editor: visual
---

List of packages needed for the whole data processing/data analysis

```{r}
library(here)
library(readxl)
library(tidyverse)
library(naniar)
library(vtable)
library(corrplot)
library(RColorBrewer)
library(ggpubr)
library(rstatix)
```

##Data Cleaning

### HAPIN Guatemala data cleaning

```{r}
#Load the files and assign them to independent data frames
hapin <- read_csv(here("data", "raw-data", "HAPIN_EX17_Guatemala_20230728.csv"))
xrflist <- read_csv(here("data", "raw-data", "sa_filters_emollinedo.csv"))

#Select only the variables I am interested to work with
hapin <- hapin %>% select(trt_blinded, timepoint, h41_m_ecm_fid, h41_b_ecm_fid, h41b_filter1, h41b_filter2, h42_cook, h42_smoke, h42_coil, h42_trash, h42_kero, h42_incence, h42_generator, h42_smoked, h42_crop, h42_smoky_kitc, ECM_grav_neph_conc_M, ECM_bc_conc_M, CO_avg_ppm_M)

#Make sure all filter IDs are upper case, using the `mutate()` function with `toupper()`
hapin <- hapin %>% 
  mutate(h41_m_ecm_fid = toupper(h41_m_ecm_fid),
         h41b_filter1 = toupper(h41b_filter1),
         h41_b_ecm_fid = toupper(h41_b_ecm_fid),
         h41b_filter2 = toupper(h41b_filter2))

#Select all filters that match with the list of filters from XRF, using the `filter()` function.
hapin <- hapin %>%
  filter(h41_m_ecm_fid %in% xrflist$filter_id |
           h41b_filter1 %in% xrflist$filter_id |
           h41_b_ecm_fid %in% xrflist$filter_id |
           h41b_filter2 %in% xrflist$filter_id)

#Concatenate `trt_blinded` and `timepoint` into a new temporary variable
hapin$concatenated <- paste(hapin$trt_blinded, hapin$timepoint)

#Create a new variable `fueltype` that categorizes all the observations by type of fuel ('R' means the visit belongs to the 'Control' treatment and 'Q' for the 'Intervention' treatment)
hapin <- mutate(hapin, fueltype = case_when(
  endsWith(concatenated, "BL") ~ "Biomass", #BL corresponds to the baseline, where all the participants had a biomass type fuel stove
  endsWith(concatenated, "R P1") ~ "Biomass",
  endsWith(concatenated, "R P2") ~ "Biomass",
  endsWith(concatenated, "Q P1") ~ "LPG",
  endsWith(concatenated, "Q P2") ~ "LPG",
  endsWith(concatenated, "R BLP1") ~ "Biomass",
  endsWith(concatenated, "R P1P2") ~ "Biomass",
  endsWith(concatenated, "R P2B1") ~ "Biomass",
  endsWith(concatenated, "Q BLP1") ~ "LPG",
  endsWith(concatenated, "Q P1P2") ~ "LPG",
  endsWith(concatenated, "Q P2B1") ~ "LPG"))

#Rename some of the variables and delete unnecessary ones
hapin <- hapin %>% rename(filter_id = "h41_m_ecm_fid", dup1_filterid = "h41b_filter1", dup2_filterid = "h41b_filter2",
                         blank_filterid = "h41_b_ecm_fid", arm = "trt_blinded", pm25 = "ECM_grav_neph_conc_M",
                         bc = "ECM_bc_conc_M", co = "CO_avg_ppm_M", stove = "h42_cook", smoke = "h42_smoke",
                         coil = "h42_coil", trash = "h42_trash", kerosene = "h42_kero", incense = "h42_incence",                                   generator = "h42_generator", smoky = "h42_smoked", crop = "h42_crop", stove_other = "h42_smoky_kitc") %>% 
  select(-c(concatenated, timepoint)) %>% 
  mutate(arm = case_when(arm == 'R' ~ 'Control', arm == 'Q' ~ 'Intervention', TRUE ~ arm)) #Recode values from the 'arm' variable

#Creating the dataframe for personal sample filters, and deleting the other columns with unnecessary filter IDs
hapin_personal <- hapin %>% filter(filter_id %in% xrflist$filter_id) %>% 
  select(-c(blank_filterid, dup1_filterid, dup2_filterid))

#Creating the dataframe for duplicates in column 4, and deleting the other columns with unnecessary filter IDs, and renaming `dup1_filterid` to `filter_id`.
hapin_dup1 <- hapin %>% filter(dup1_filterid %in% xrflist$filter_id) %>% 
  select(-c(blank_filterid, filter_id, dup2_filterid)) %>% 
  rename(filter_id = "dup1_filterid")

#Creating the dataframe for duplicates in column 5, and deleting the other columns with unnecessary filter IDs, and renaming `dup2_filterid` to `filter_id`.
hapin_dup2 <- hapin %>% filter(dup2_filterid %in% xrflist$filter_id) %>% 
  select(-c(blank_filterid, filter_id, dup1_filterid)) %>% 
  rename(filter_id = "dup2_filterid")

#Creating the dataframe for blank filters, and deleting the other columns with unnecessary filter IDs, and renaming `blank_filterid` to `filter_id`.
hapin_blanks <- hapin %>% filter(blank_filterid %in% xrflist$filter_id) %>% 
  select(-c(filter_id, dup1_filterid, dup2_filterid)) %>% 
  rename(filter_id = "blank_filterid")

#Combining the `personal` and `dup` dataframes into a single one, using `bind_rows()`
hapin_filters <- bind_rows(hapin_personal, hapin_dup1, hapin_dup2)

#Save the samples and blanks dataframes to .RDS format files
saveRDS(hapin_filters, file = here("data", "processed-data", "hapin_samples.rds"))
saveRDS(hapin_blanks, file = here("data", "processed-data", "hapin_blanks.rds"))
```

### Chemical species Concentrations data cleaning

```{r}
#Loading the chemical species concentration data
concentration <- read_excel(here("data", "raw-data", "conc_unc_hapingt_emollinedo.xlsx"), sheet = "Concentration")

#Loading the chemical species limit of detection values
lod <- read_excel(here("data", "raw-data", "conc_unc_hapingt_emollinedo.xlsx"), sheet = "DL")

#having a glimpse of this dataframe
glimpse(concentration)

#Remove brackets, convert to numeric, and round to 4 decimal places
concentration <- concentration %>%
  mutate(across(3:26, ~ round(as.numeric(str_remove_all(., "\\[|\\]")), 4))) %>% 
  rename(filter_id = "Filter")

#Create dataframe with the concentrations for the blank filters for each element
blanks <- concentration %>% filter(filter_id %in% hapin_blanks$filter_id) %>% 
  select(-c(filter_id, Type))

#Remove blank filters from the `concentration` dataframe
concentration <- concentration %>%
  anti_join(hapin_blanks, by = "filter_id")

#Calculate the median for each element, using the blank filter values
median_blanks <- blanks %>% select(Mg: Na) %>%
  map_dbl(median) %>% round(4)

median_blanks

#Round the limit of detection values to 4 decimal places
lod <- lod %>% mutate(`Concentration (ug/cm2)` = round(`Concentration (ug/cm2)`, 4))
print(lod)

#Estimate the percentage of values below detection limit for each element
bdl <- concentration %>%
  summarise(
    percent_bdl_mg = sum(Mg < 0.0051) / n() * 100,
    percent_bdl_al = sum(Al < 0.0059) / n() * 100,
    percent_bdl_si = sum(Si < 0.0068) / n() * 100,
    percent_bdl_s = sum(S < 0.0026) / n() * 100,
    percent_bdl_k = sum(K < 0.0040) / n() * 100,
    percent_bdl_ca = sum(Ca < 0.0040) / n() * 100,
    percent_bdl_ti = sum(Ti < 0.0086) / n() * 100,
    percent_bdl_cr = sum(Cr < 0.0055) / n() * 100,
    percent_bdl_mn = sum(Mn < 0.0030) / n() * 100,
    percent_bdl_fe = sum(Fe < 0.0043) / n() * 100,
    percent_bdl_ni = sum(Ni < 0.0043) / n() * 100,
    percent_bdl_cu = sum(Cu < 0.0080) / n() * 100,
    percent_bdl_zn = sum(Zn < 0.0024) / n() * 100,
    percent_bdl_ga = sum(Ga < 0.0049) / n() * 100,
    percent_bdl_as = sum(As < 0.0062) / n() * 100,
    percent_bdl_se = sum(Se < 0.0181) / n() * 100,
    percent_bdl_cd = sum(Cd < 0.0932) / n() * 100,
    percent_bdl_in = sum(In < 0.1089) / n() * 100,
    percent_bdl_sn = sum(Sn < 0.1844) / n() * 100,
    percent_bdl_te = sum(Te < 0.2166) / n() * 100,
    percent_bdl_i = sum(I < 0.3149) / n() * 100,
    percent_bdl_pb = sum(Pb < 0.0324) / n() * 100,)

print(bdl)

#Select only the filter ID, type and the 10 elements that have more than 50% of their data not below the detection limit
concentration <- concentration %>% select(c(filter_id, Type, Mg, Al, Si, S, K, Ca, Ti, Mn, Fe, Zn)) %>% 
  arrange(filter_id) #Also I decided to arrange them ascending based on the filter_id

#Substract the median blank value for the elements that need blank adjustment
concentration <- concentration %>%
  mutate(Mg = Mg - 0.0112, S = S - 0.0044, K = K - 0.0014, Ca = Ca - 0.0048, Fe = Fe - 0.0038, Zn = Zn - 0.0045)

#Replace all values below the detection limit with DL/sqrt(2)
concentration <- concentration %>% 
    mutate(Mg = round(if_else(Mg < 0.0051, 0.0051 / sqrt(2), Mg), 4), Al = round(if_else(Al < 0.0059, 0.0059 / sqrt(2), Al), 4),
           Si = round(if_else(Si < 0.0068, 0.0068 / sqrt(2), Si), 4), S = round(if_else(S < 0.0026, 0.0026 / sqrt(2), S), 4),
           K = round(if_else(K < 0.0016, 0.0016 / sqrt(2), K), 4), Ca = round(if_else(Ca < 0.0040, 0.0040 / sqrt(2), Ca), 4),
           Ti = round(if_else(Ti < 0.0086, 0.0086 / sqrt(2), Ti), 4), Mn = round(if_else(Mn < 0.0030, 0.0030 / sqrt(2), Mn), 4),
           Fe = round(if_else(Fe < 0.0043, 0.0043 / sqrt(2), Fe), 4), Zn = round(if_else(Zn < 0.0024, 0.0024 / sqrt(2), Zn), 4))

#I will save these concentrations in a new dataframe `concentrations_ugcm2` that will be used for uncertainty estimations in future steps
concentrations_ugcm2 <- concentration %>% select(-Type)

#Converting the concentrations from (ug/cm2) to (ug/m3)
concentration <- concentration %>%
  mutate(across(Mg:Zn, ~ round(.x * 0.7853 / 0.43, 4)))

#Now adding the BC concentrations value from the `hapin_filters` df
hapin_filters <- hapin_filters %>% arrange(filter_id) #First arrange in ascending order by filter_id
concentration <- bind_cols(concentration, hapin_filters$bc) %>% rename(BC = "...13") %>% select(-Type) #Join the BC concentrations to the `concentrations` df, also renaming the column to "BC" and deleting the 'Type' column

#Add the `arm` and `fueltype` variables to the `concentration` dataframe
concentration <- concentration %>%
  mutate(arm = hapin_filters$arm, fueltype = hapin_filters$fueltype)

#Check if there are any missing values from the concentration file
gg_miss_var(concentration)

#Use na.omit() to remove the spotted observation with missing value
concentration <- na.omit(concentration)

#Save this dataframe as RDS file
saveRDS(concentration, file = here("data", "processed-data", "concentration.rds"))
```

### Chemical species Uncertainties data processing

```{r}
#Load the uncertainty data
uncertainty <- read_excel(here("data", "raw-data", "conc_unc_hapingt_emollinedo.xlsx"), sheet = "Uncertainty")

#Remove the chemical species with more than 50% of the observations with values BDL
uncertainty <- uncertainty %>% select(c(Filter, Mg, Al, Si, S, K, Ca, Ti, Mn, Fe, Zn)) %>% #Select only the necessary variables
  rename(filter_id = "Filter") %>% #Rename the Filter column
  mutate_at(vars(Mg:Zn), ~ round(., 4)) %>% #Round the uncertainties to 4 decimal places
  arrange(filter_id) %>% #Arrange observations based on Filter ID
  anti_join(hapin_blanks, by = "filter_id") #Remove blank filters

#Have a glimpse of the data
glimpse(uncertainty)

#Create a list for the detection limits for each element
dl <- c(Mg = 0.0051, Al = 0.0059, Si = 0.0068, S = 0.0026, K = 0.0016, 
        Ca = 0.0040, Ti = 0.0086, Mn = 0.0030, Fe = 0.0043, Zn = 0.0024)

# Recalculate the values in df based on the conditions
uncertainty <- uncertainty %>%
  mutate(across(all_of(names(dl)),
                ~ round(if_else(concentrations_ugcm2[[cur_column()]] < dl[[cur_column()]], (5/6) * dl[[cur_column()]], .), 4),
                .names = "{.col}"))

# Define the variables to be recalculated
variables <- c('Mg', 'Al', 'Si', 'S', 'K', 'Ca', 'Ti', 'Mn', 'Fe', 'Zn')

# Recalculate the values using the given equation
uncertainty <- uncertainty %>%
  mutate(across(all_of(variables), ~ round( #Here, selecting all the variables to be re-calculated
    sqrt((0.7853/0.43)^2 * .^2 + (concentrations_ugcm2[[cur_column()]]/0.43)^2 * 0.05^2 + (-concentrations_ugcm2[[cur_column()]]*0.7853/0.43^2)^2 * (0.05 * 0.43)^2), 4), .names = "{.col}")) #Input the equation

#Load file to estimate BC uncertainties
atn <- read_csv(here("data", "raw-data", "ATN_summary.csv"))

#Explore the structure of the file
glimpse(atn)

#Estimate total attenuation by sample
atn <- atn %>% mutate(ATN = round(log(post_ATN/pre_ATN) *100, 2))

#I will remove the observation with `filter_id` 3M53864 from the `uncertainty` df, since this observation does not have a BC concentration available
uncertainty <- uncertainty %>% filter(filter_id != "3M53864") %>% 
  mutate(BC = round((1370 * sqrt((atn$ATN)^2 * 0.000005^2 + 0.000079^2 * 1.16^2)), 4))#Estimate the BC uncertainties

#I will remove the observation with `filter_id` 3M53864 from the `hapin_filters` df
hapin_filters <- hapin_filters %>% filter(filter_id != "3M53864")

#Add the `arm` and `fueltype` variables to the `uncertainty` dataframe
uncertainty <- uncertainty %>%
  mutate(arm = hapin_filters$arm, fueltype = hapin_filters$fueltype)

#Save this dataframe as RDS file
saveRDS(uncertainty, file = here("data", "processed-data", "uncertainty.rds"))
```

## Data Exploring

First, I produced two boxplots, that summarize the concentrations categorized by `arm` and `fueltype`. First I will create a dataframe in longer format using `pivot_longer()`. Then creating the plots using `ggplot()`. In addition, I will add the significance test value to the plots to visualize which chemical species are statistically different among groups.

```{r}
#Pivot longer the `concentration` df to make the boxplots
conc_longer <- concentration %>% #First a df with the `arm` variable
  pivot_longer(cols = -c(filter_id, fueltype, arm),
               names_to = "species", values_to = "concentration")

#conduct the t-test of concentration-fueltype
stat.test <- conc_longer %>% 
  group_by(species) %>% 
  t_test(concentration ~ fueltype) %>% 
  add_significance() #to add the significance values

#Add the position coordinates
stat.test <- stat.test %>% 
  add_x_position(x= "species", dodge = 0.8) %>% #x position is species, and the length of significance bar is 0.8
  rstatix::add_y_position(y.trans = function(x){log10(x)}) #add y position, in this case specify using a function that it is in log10 scale

#Creating the boxplot of concentrations by fueltype
options(scipen = 999) #to remove scientific notation for smaller values (i.e. changes from 1*10-3 to 0.001)
ggplot(conc_longer)+
  geom_boxplot(aes(x= species, y= concentration, fill= fueltype))+
  scale_fill_brewer(palette = "Paired")+
  labs(x= "Chemical species", y= "Concentration (ug/m3)")+
  scale_y_log10()+
  guides(fill= guide_legend(title= "Fuel type"))+ #Change legend from fuel_type to Fuel type
  theme_bw(base_size = 15)+
  stat_pvalue_manual(stat.test, label = "p.signif", tip.length = 0.001) #Add the significance bars, use 'p.signif' to add *** or 'ns'

#Creating the boxplot of concentrations by arm
options(scipen = 999) #to remove scientific notation for smaller values (i.e. changes from 1*10-3 to 0.001)
ggplot(conc_longer)+
  geom_boxplot(aes(x= species, y= concentration, fill= arm))+
  scale_fill_brewer(palette = "Paired")+
  labs(x= "Chemical species", y= "Concentration (ug/m3)")+
  scale_y_log10()+
  guides(fill= guide_legend(title= "Study Arm"))+ #Change legend from fuel_type to Fuel type
  theme_bw(base_size = 15)+
  stat_pvalue_manual(stat.test, label = "p.signif", tip.length = 0.001) #Add the significance bars, use 'p.signif' to add *** or 'ns'
```

The boxplots for `fueltype` and `arm` show what I was expecting in terms of how the data is distributed. The concentrations seem higher in some elemental especies such as BC, Ca, K, Mg, Mn and S. The following steps will be to evaluate if those differences in concentrations are statistically different.

And now producing summary statistics by type of fuel, using the `sumtable()` function.

```{r}
#Produce summary statistics by type of fuel
#Remove arm and fueltype columns
conc <- concentration %>% select(-arm)

#Using the vtable package, which creates a table
sumtable(conc, group = 'fueltype', group.long = T, add.median = T)
```

Now I created a correlation plot using the `corrplot()` function, to explore which elements could be more associated between each other.

```{r}
#First, delete unnecessary variables to conduct the correlation test
cor <- concentration %>% select(-c(filter_id, fueltype, arm))

#Perform the spearman correlation test
cor <- stats::cor(cor, method = "pearson", use = "complete.obs") #I had to input the `use=` argument since there are missing values in the BC variable. I will correct them later.

#Create the correlation plot
corrplot(cor, method = "color", type = "lower", #Color form and displays at the lower portion
         col = COL2("RdYlBu", 20), order = "hclust", #Color palette, and order or the species
         tl.col = "orangered4", tl.srt = 0, tl.cex = 0.9,  #Color of the axis, position and size
         addCoef.col = "gray10", number.font = 2, number.cex = 0.8, #Display correlation number, set the font and size
         col.lim = c(0, 1), is.corr = T) #Set the color limits at the bar
```

The correlation plot above shows that Si and Fe are strongly correlated, which might suggest that these elements represent one specific source of pollution. Other high correlations observed are between Al and Si, and Fe and Ti.

I produced a summary table of the uncertainties, using the `vtable` package.

```{r}
#Remove the `arm` variable
unc_fuel <- uncertainty %>% select(-arm)

#Use the `sumtable()` function from the `vtable` package, which creates a table
sumtable(unc_fuel, group = 'fueltype', group.long = T, add.median = T, out = 'kable')

#Remove the `fueltype` variable
unc_arm <- uncertainty %>% select(-fueltype)

#Use the `sumtable()` function from the `vtable` package, which creates a table
sumtable(unc_arm, group = 'arm', group.long = T, add.median = T, out = 'kable')

```

The tables above are just a way to represent the summary statistics of the uncertainties, however, the uncertainties will only be useful for the PMF analysis after the conclusion of this project. One thing noticed also from this table, is that there are some infinite values from the BC concentrations, which also show that there are missing values that have to be corrected in earlier steps. From this tables we can also see that there are 376 observations from the 'Biomass' and 254 from the 'LPG' fueltypes. Meanwhile, there are 323 observations from the 'Control' arm and 307 from the 'Intervention' arm.

Now, checking if there are missing values in the `hapin` dataframe for any of the observations using the `gg_miss_var()` function.

```{r}
gg_miss_var(hapin)
```

As seen above, there are some missing values from the observational exposures (kerosene to coil), there is just one missing observation for PM2.5 and BC concentrations and there is more than 30 missing observations for CO (carbon monoxide). The missing observation of BC might explain the missing value spotted before, so this observation might actually have to be dropped down.

Now I created a barplot showing the distribution of responses for the categorical variables `kerosene`, `trash`, `stove_other`, `stove`, `smoky`, `incense`, `generator`, `smoke`, `crop`, and `coil`. First doing data manipulation so the df is useful for plotting.

```{r}
#Pivot longer the `hapin` df to make a barplot
hapin_longer <- hapin %>%
  select(c(`kerosene`, `trash`, `stove_other`, `stove`, `smoky`, `incense`, `generator`, `smoke`, `crop`, `coil`)) %>% #Here selecting only the necessary variables
  pivot_longer(cols = everything(),
               names_to = "exposure", values_to = "response") #Create the longer df

#Create a simple barplot
ggplot(hapin_longer, aes(x= exposure, fill= response))+
  geom_bar(position = "dodge")+
  theme_classic()
```

The bar plot shows that the majority of participants recorded to not being exposed to most of the exposures, except to `stove`, which is the stove that the participants use. Others also recorded that were exposed to other stoves, or to trash burning.

And finally plotting the PM2.5, BC and CO concentrations in different graphs using `ggplot()`

```{r}
#Histogram plot for the PM2.5 concentrations
ggplot(hapin, aes(x= pm25, fill= fueltype))+
  geom_histogram()+
  theme_classic()+
  labs(x= "PM2.5 Concentration (ug/m3)")

#Histogram plot for the BC concentrations
ggplot(hapin, aes(x= bc, fill= fueltype))+
  geom_histogram()+
  theme_classic()+
  labs(x= "Black Carbon Concentration (ug/m3)")

#Histogram plot for the CO concentrations
ggplot(hapin, aes(x= co, fill= fueltype))+
  geom_histogram()+
  theme_classic()+
  labs(x= "Carbon Monoxide Concentration (ppm)")
```

The histograms for the concentration of the three pollutants show that they don't follow a normal distribution. Most of the concentrations are low, and it is also seen that overall the concentrations in the LPG fueltype are lower than the Biomass group. If these variables will be used for modeling purposes, they should be log-transformed or using a logistic regression approach.

## Data Analysis

To answer question 1, I will plot the correlations in a correlation plot using the `corrplot()` function. I used the previous code, since it was already developed, but editing to add more detail to the plot.

```{r}
#Create the correlation plot
corrplot(cor, method = "color", type = "lower", #Color form and displays at the lower portion
         col = COL2("RdYlBu", 20), order = "hclust", #Color palette, and order or the species
         tl.col = "orangered4", tl.srt = 0, tl.cex = 1,  #Color of the axis, position and size
         addCoef.col = "gray10", number.font = 2, number.cex = 0.8, #Display correlation number, set the font and size
         col.lim = c(0, 1), is.corr = T) #Set the color limits at the bar
```

To answer question 2, I will conduct a T-test using the `tidymodels` package

```{r}
# Conduct T-tests for the chemical species based on 'fueltype'
results <- map_dfr(2:12, function(i) { #To select columns that contain the variables to analyze
  variable <- names(concentration)[i] #The dataframe where the name of the variables are
  t_test <- t.test(reformulate('fueltype', response = variable), data = concentration) #Select the fueltype variable

  tibble( #Format the tibble
    variable = variable,
    statistic = t_test$statistic,
    p_value = t_test$p.value,
    estimate = t_test$estimate,
    null_value = t_test$null.value,
    alternative = t_test$alternative,
    method = t_test$method,
    conf_low = t_test$conf.int[1],
    conf_high = t_test$conf.int[2]
  )
})

results #Print the results as a tibble
```

To answer question 3, I will conduct a regression analysis. But first I will merge the data from `hapin_filters` with the `concentration` data.

```{r}
#remove the 'arm' and 'fueltype' variables from the 'concentration' df
conc2 <- concentration %>% select(-c(arm, fueltype))

#Use 'inner_join()' to merge dataframes
hapin_merged <- inner_join(conc2, hapin_filters, by= "filter_id")

#Save new df as a .RDS file
saveRDS(hapin_merged, file = here("data", "processed-data", "hapin-final.rds"))
```

Now conducting the model selection using the `tidymodels` package. For the models, I used the gamma distribution family, since the concentrations of each chemical species could be explained by the gamma distribution. I used all the recorded exposures except incence and generator, since these had low variability in the answers.

```{r}
#First, loading the `tidymodels` package
library(tidymodels)

# Create a generalized linear regression model specification using gamma distribution (This will be used for all the models)
model_spec <- linear_reg(mode = "regression") %>%
  set_engine("glm", family = Gamma(link = "log"))

###########To predict the BC variable##############
# Create the recipe for preprocessing
recipeBC <- recipe(BC ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

# Create a workflow
workflowBC <- workflow() %>%
  add_recipe(recipeBC) %>%
  add_model(model_spec)

# Fit the model to the data
fitBC <- workflowBC %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsBC <- fitBC %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsBC)

###########To predict the Ca variable##############
# Create the recipe for preprocessing
recipeCa <- recipe(Ca ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

# Create a workflow
workflowCa <- workflow() %>%
  add_recipe(recipeCa) %>%
  add_model(model_spec)

# Fit the model to the data
fitCa <- workflowCa %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsCa <- fitCa %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsCa)

###########To predict the K variable##############
# Create the recipe for preprocessing
recipeK <- recipe(K ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

# Create a workflow
workflowK <- workflow() %>%
  add_recipe(recipeK) %>%
  add_model(model_spec)

# Fit the model to the data
fitK <- workflowK %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsK <- fitK %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsK)

###########To predict the Mg variable##############
# Create the recipe for preprocessing
recipeMg <- recipe(Mg ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

# Create a workflow
workflowMg <- workflow() %>%
  add_recipe(recipeMg) %>%
  add_model(model_spec)

# Fit the model to the data
fitMg <- workflowMg %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsMg <- fitMg %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsMg)

###########To predict the Mn variable##############
# Create the recipe for preprocessing
recipeMn <- recipe(Mn ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

# Create a workflow
workflowMn <- workflow() %>%
  add_recipe(recipeMn) %>%
  add_model(model_spec)

# Fit the model to the data
fitMn <- workflowMn %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsMn <- fitMn %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsMn)

###########To predict the S variable##############
# Create the recipe for preprocessing
recipeSul <- recipe(S ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

# Create a workflow
workflowSul <- workflow() %>%
  add_recipe(recipeSul) %>%
  add_model(model_spec)

# Fit the model to the data
fitSul <- workflowSul %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsSul <- fitSul %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsSul)

###########To predict the Al variable##############
# Create the recipe for preprocessing
recipeAl <- recipe(Al ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

# Create a workflow
workflowAl <- workflow() %>%
  add_recipe(recipeAl) %>%
  add_model(model_spec)

# Fit the model to the data
fitAl <- workflowAl %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsAl <- fitAl %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsAl)

###########To predict the Fe variable##############
# Create the recipe for preprocessing
recipeFe <- recipe(Fe ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

# Create a workflow
workflowFe <- workflow() %>%
  add_recipe(recipeFe) %>%
  add_model(model_spec)

# Fit the model to the data
fitFe <- workflowFe %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsFe <- fitFe %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsFe)

###########To predict the Si variable##############
# Create the recipe for preprocessing
recipeSi <- recipe(Si ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

# Create a workflow
workflowSi <- workflow() %>%
  add_recipe(recipeSi) %>%
  add_model(model_spec)

# Fit the model to the data
fitSi <- workflowSi %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsSi <- fitSi %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsSi)

###########To predict the Ti variable##############
# Create the recipe for preprocessing
recipeTi <- recipe(Ti ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

# Create a workflow
workflowTi <- workflow() %>%
  add_recipe(recipeTi) %>%
  add_model(model_spec)

# Fit the model to the data
fitTi <- workflowTi %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsTi <- fitTi %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsTi)

###########To predict the Zn variable##############
# Create the recipe for preprocessing
recipeZn <- recipe(Zn ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

# Create a workflow
workflowZn <- workflow() %>%
  add_recipe(recipeZn) %>%
  add_model(model_spec)

# Fit the model to the data
fitZn <- workflowZn %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsZn <- fitZn %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsZn)
```

Based on the previous models, it seems like the most important variables are `fueltype` (as expected), `kerosene`, `stove_other` and `smoke` depending on the chemical species. So I will test new regression models for all species (except Al and Zn, since they seem not to be explained by any of the exposures), using only their most significant predictors.

```{r}
###########To predict the BC variable##############
# Create the recipe for preprocessing
recipeBC_red <- recipe(BC ~ kerosene + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

# Create a workflow
workflowBC_red <- workflow() %>%
  add_recipe(recipeBC_red) %>%
  add_model(model_spec)

# Fit the model to the data
fitBC_red <- workflowBC_red %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsBC_red <- fitBC_red %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsBC_red)

###########To predict the Ca variable##############
# Create the recipe for preprocessing
recipeCa_red <- recipe(Ca ~ fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

# Create a workflow
workflowCa_red <- workflow() %>%
  add_recipe(recipeCa_red) %>%
  add_model(model_spec)

# Fit the model to the data
fitCa_red <- workflowCa_red %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsCa_red <- fitCa_red %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsCa_red)

###########To predict the K variable##############
# Create the recipe for preprocessing
recipeK_red <- recipe(K ~ smoke + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

# Create a workflow
workflowK_red <- workflow() %>%
  add_recipe(recipeK_red) %>%
  add_model(model_spec)

# Fit the model to the data
fitK_red <- workflowK_red %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsK_red <- fitK_red %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsK_red)

###########To predict the Mg variable##############
# Create the recipe for preprocessing
recipeMg_red <- recipe(Mg ~ kerosene + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

# Create a workflow
workflowMg_red <- workflow() %>%
  add_recipe(recipeMg_red) %>%
  add_model(model_spec)

# Fit the model to the data
fitMg_red <- workflowMg_red %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsMg_red <- fitMg_red %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsMg_red)

###########To predict the Mn variable##############
# Create the recipe for preprocessing
recipeMn_red <- recipe(Mn ~ fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

# Create a workflow
workflowMn_red <- workflow() %>%
  add_recipe(recipeMn_red) %>%
  add_model(model_spec)

# Fit the model to the data
fitMn_red <- workflowMn_red %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsMn_red <- fitMn_red %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsMn_red)

###########To predict the S variable##############
# Create the recipe for preprocessing
recipeS_red <- recipe(S ~ fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

# Create a workflow
workflowS_red <- workflow() %>%
  add_recipe(recipeS_red) %>%
  add_model(model_spec)

# Fit the model to the data
fitS_red <- workflowS_red %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsS_red <- fitS_red %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsS_red)

###########To predict the Fe variable##############
# Create the recipe for preprocessing
recipeFe_red <- recipe(Fe ~ stove, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

# Create a workflow
workflowFe_red <- workflow() %>%
  add_recipe(recipeFe_red) %>%
  add_model(model_spec)

# Fit the model to the data
fitFe_red <- workflowFe_red %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsFe_red <- fitFe_red %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsFe_red)

###########To predict the Si variable##############
# Create the recipe for preprocessing
recipeSi_red <- recipe(Si ~ fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

# Create a workflow
workflowSi_red <- workflow() %>%
  add_recipe(recipeSi_red) %>%
  add_model(model_spec)

# Fit the model to the data
fitSi_red <- workflowSi_red %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsSi_red <- fitSi_red %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsSi_red)

###########To predict the Ti variable##############
# Create the recipe for preprocessing
recipeTi_red <- recipe(Ti ~ fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

# Create a workflow
workflowTi_red <- workflow() %>%
  add_recipe(recipeTi_red) %>%
  add_model(model_spec)

# Fit the model to the data
fitTi_red <- workflowTi_red %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsTi_red <- fitTi_red %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsTi_red)
```

In this part I will evaluate both models from each chemical species (TBD)

```{r}

```

### 2024-03-15 code/data_cleaning_code.qmd
Project Part 3
Created a new QMD file: data_analysis_code, and a Supplementary Material word document. Also, the manuscript file has been updated to include the preliminary results. ReadMe files updated and figures/tables edited.

@@
 library(tidyverse)
+library(naniar)
 ```
@@
   mutate(arm = hapin_filters$arm, fueltype = hapin_filters$fueltype)
+```
+
+I used the `gg_miss_var()` function to check if there are any missing values from the concentrations file.
+(For the purpose of the final project and the personal project, any missing values will have to be removed).
+
+```{r}
+gg_miss_var(concentration)
+```
+
+There is one missing value from the `BC` variable.
+After doing some research, there is no record that the data information from this filter is available, so the best decision was to drop this value.
+And then, save the df to an RDS file.
+
+```{r}
+#Use na.omit() to remove the observation with missing value
+concentration <- na.omit(concentration)
 
@@
 #Load file to estimate BC uncertainties
-atn <- read_csv(here("data", "raw-data", "ATN_summary_SA.csv"))
+atn <- read_csv(here("data", "raw-data", "ATN_summary.csv"))
 
@@
 ```{r}
-#Estimate the BC uncertainties
-uncertainty <- uncertainty %>% 
-  mutate(BC = round((1370 * sqrt((atn$ATN)^2 * 0.000005^2 + 0.000079^2 * 1.16^2)), 4))
+#First I will remove the observation with `filter_id` 3M53864 from the `uncertainty` df, since this observation does not have a BC concentration available
+uncertainty <- uncertainty %>% filter(filter_id != "3M53864") %>% 
+  mutate(BC = round((1370 * sqrt((atn$ATN)^2 * 0.000005^2 + 0.000079^2 * 1.16^2)), 4))#Estimate the BC uncertainties
 ```
@@
 ```{r}
+#First I will remove the observation with `filter_id` 3M53864 from the `hapin_filters` df
+hapin_filters <- hapin_filters %>% filter(filter_id != "3M53864")
+
 #Add the `arm` and `fueltype` variables to the `uncertainty` dataframe

### 2024-03-15 code/data_exploring_code.qmd
Project Part 3
Created a new QMD file: data_analysis_code, and a Supplementary Material word document. Also, the manuscript file has been updated to include the preliminary results. ReadMe files updated and figures/tables edited.

@@
 library(naniar)
+library(ggpubr)
+library(rstatix)
 ```
@@
 Then creating the plots using `ggplot()`.
+In addition, I will add the significance test value to the plots to visualize which chemical species are statistically different among groups.
 
@@
 
+#conduct the t-test of concentration-fuel_type
+stat.test <- conc_longer %>% 
+  group_by(species) %>% 
+  t_test(concentration ~ fueltype) %>% 
+  add_significance() #to add the significance values
+
+#Add the position coordinates
+stat.test <- stat.test %>% 
+  add_x_position(x= "species", dodge = 0.8) %>% #x position is species, and the length of significance bar is 0.8
+  add_y_position(y.trans = function(x){log10(x)}) #add y position, in this case specify using a function that it is in log10 scale
+
 #Creating the boxplot of concentrations by fueltype
@@
   scale_fill_brewer(palette = "Paired")+
-  labs(x= "Elemental species", y= "Concentration (ug/m3)")+
+  labs(x= "Chemical species", y= "Concentration (ug/m3)")+
   scale_y_log10()+
   guides(fill= guide_legend(title= "Fuel type"))+ #Change legend from fuel_type to Fuel type
-  theme_bw(base_size = 15)
+  theme_bw(base_size = 15)+
+  stat_pvalue_manual(stat.test, label = "p.signif", tip.length = 0.001) #Add the significance bars, use 'p.signif' to add *** or 'ns'
 
@@
   scale_fill_brewer(palette = "Paired")+
-  labs(x= "Elemental species", y= "Concentration (ug/m3)")+
+  labs(x= "Chemical species", y= "Concentration (ug/m3)")+
   scale_y_log10()+
   guides(fill= guide_legend(title= "Study Arm"))+ #Change legend from fuel_type to Fuel type
-  theme_bw(base_size = 15)
+  theme_bw(base_size = 15)+
+  stat_pvalue_manual(stat.test, label = "p.signif", tip.length = 0.001) #Add the significance bars, use 'p.signif' to add *** or 'ns'
 ```
@@
 The following steps will be to evaluate if those differences in concentrations are statistically different.
+
+And now producing summary statistics by type of fuel, using the `sumtable()` function.
+
+```{r}
+#Produce summary statistics by type of fuel
+#Remove arm and fueltype columns
+conc <- concentration %>% select(-arm)
+
+#Using the vtable package, which creates a table
+sumtable(conc, group = 'fueltype', group.long = T, add.median = T)
+```
 

### 2024-03-15 products/manuscript/Manuscript.qmd
Project Part 3
Created a new QMD file: data_analysis_code, and a Supplementary Material word document. Also, the manuscript file has been updated to include the preliminary results. ReadMe files updated and figures/tables edited.

@@
 ---
-title: "Elemental characterization of personal filter samples from the HAPIN trial for Source Apportionment"
-subtitle: "Project: Part 2"
+title: "Elemental characterization of personal filter samples from the Household Air Pollution Intervention Network (HAPIN) trial for Source Apportionment"
+subtitle: "Project: Part 3"
 author: Erick E. Mollinedo
@@
     highlight-style: github
+bibliography: ../../assets/sourceapp-references.bib
+csl: ../../assets/diabetologia.csl
 ---
@@
 
-# Introduction
+## Introduction
 
-## Background
-
-Household air pollution (HAP), due to the combustion of solid fuels for cooking and heating, is one of the most prominent environmental health issues in low- and middle-income countries (LMICs).^1^ HAP has been listed as the second environmental health risk factor worldwide, just after outdoor air pollution.^2^ Exposure to HAP has been linked to more than 2.31 million deaths in 2019 and it is associated with adverse cardiovascular and respiratory health outcomes.^3,4^ Particulate matter with an aerodynamic diameter less than 2.5 micrometers (PM~2.5~) is one of the major air pollutants chemically and physically composed of crustal material, black carbon (BC), organic compounds, metals and others.
-Toxicity from PM~2.5~ depends, among many factors, on the chemical composition of the particles which heavily depend on the sources of emissions.^5,6^
+Household air pollution (HAP), due to the combustion of solid fuels for cooking and heating, is one of the most prominent environmental health issues in low- and middle-income countries (LMICs) [@gordon_respiratory_2014].
+HAP has been listed as the second environmental health risk factor worldwide, just after outdoor air pollution.
+Exposure to HAP has been linked to more than 2.31 million deaths in 2019 and it is associated with adverse cardiovascular and respiratory health outcomes [@murray_global_2020; @pope_health_2006].
+Particulate matter with an aerodynamic diameter less than 2.5 micrometers (PM~2.5~) is one of the major air pollutants chemically and physically composed of crustal material, black carbon (BC), organic compounds, metals and others.
+Toxicity from PM~2.5~ depends, among many factors, on the chemical composition of the particles which heavily depend on the sources of emissions [@clasen_design_2020; @mohammed_potential_nodate].
 
 The identification of point sources of pollution is important for the development of strategies to improve air quality.
-Source apportionment (SA) is a methodology to reconstruct the impact of emissions from different sources, a method applied to different pollutants such as particulate matter.^7^ The most used SA methods includes receptor models, an example of this is the positive matrix factorization (PMF).
+Source apportionment (SA) is a methodology to reconstruct the impact of emissions from different sources, a method applied to different pollutants such as particulate matter [@viana_source_2008].
+The most used SA methods includes receptor models, an example of this is the positive matrix factorization (PMF).
 One advantage of PMF is that it does not require the source profiles prior to analysis and there is no limit on the number of sources, however, it is required the knowledge of potential source profiles.
+Said requirement to know the potential sources is associated with the profiles of chemical species, given that the concentration of certain organic and inorganic species alongside their uncertainties can be a good estimate of the point sources [@reff_receptor_2007].
 Another important piece before conducting source apportionment is to correctly process the data (which includes concentrations and uncertainties), and to explore associations to try to identify the potential sources.
-Just a few studies about SA in LMICs have been published, therefore it is important to characterize chemical composition and potential sources of air pollutants in these settings.^8-10^
+Just a few studies about SA in LMICs have been published, therefore it is important to characterize chemical composition and potential sources of air pollutants in these settings [@zhou_chemical_2014; @sharma_source_2016; @martinez-cinco_source_2016].
 
-The Household Air Pollution Intervention Network (HAPIN) trial was a randomized controlled trial where a liquefied petroleum gas (LPG) stove was delivered as an intervention to half of the pregnant women participants, meanwhile the other half remained cooking using their biomass stove.
+The Household Air Pollution Intervention Network (HAPIN) trial was a randomized controlled trial conducted in four different LMICs, where a liquefied petroleum gas (LPG) stove was delivered as an intervention to half of the pregnant women participants, meanwhile the other half remained cooking using their biomass stove.
 The intervention was evaluated in terms of reduction of exposures and specific health outcomes by comparing the two study groups (Control and Intervention).
 Baseline measurements were taken to the participants, and then followed-up for about 18 months, conducting exposure measurements three times during pregnancy and three times after the birth of the child.
-A description of the HAPIN trial and the methods for exposure sampling can be found elsewhere.^11,12^ A source apportionment pilot study from HAPIN was conducted in 2022.
+A description of the HAPIN trial and the methods for exposure sampling can be found elsewhere [@clasen_design_2020; @johnson_air_2020].
+A source apportionment pilot study from HAPIN was conducted in 2022.
 This study was conducted in 64 and 59 personal exposure filter samples from the pilot phase in Guatemala and Rwanda, respectively.
@@
 Based on the chemical species detected, four potential sources of PM~2.5~ were identified in both sites.
-One study caveat is the small sample size and limited assessment of chemical composition and sources between study arms, so further studies to complement the findings are needed.^13^
+One study caveat is the small sample size and limited assessment of chemical composition and sources between study arms, so further studies to complement the findings are needed [@kearns_exposures_nodate].
 
-## Aims and Hypothesis questions
+The following study aims to produce the clean dataset from the HAPIN trial in Guatemala, needed for PMF analysis and to conduct the exploratory and main data analysis to answer some questions before the PMF modelling: 1.) What chemical species are correlated among each other from these filter samples?, 2.) What are the differences in concentrations of the chemical species if the samples are categorized by type of fuel and by study arm?
+and 3.) What exposures are associated with the variations in concentrations of the most significant chemical species?
 
-One of the objectives of this project is to create a clean output that consists in a file with adjusted concentrations for all samples and each elemental species, and another file with the adjusted uncertainties for the same samples and elemental species.
-These outputs will be used as inputs for the EPA PMF 5.0 software to try to determine the potential sources that contributed to these samples.
-However, the questions I plan to answer with the data analysis of this project is:
+## Methods
 
-1.  What elements are correlated among each other from these filter samples?
-2.  Is there a difference in concentrations if we categorize the samples by type of fuel and by study arm?
-3.  What exposures could explain the variations in the concentrations of the most significant elemental species?
+### Data collection
 
-# Methods
-
-## Data collection
-
-The main exposure data comes from filter samples of personal exposure measurements in pregnant women participants from the HAPIN trial, in Guatemala.
+The main exposure data comes from filter samples of PM~2.5~ personal exposure measurements in pregnant women participants from the HAPIN trial, in Guatemala.
+The Enhanced Children's MicroPEM (ECM) was the instrument used for the measurement of PM~2.5~ exposures.
+In summary, the instrument works as an air pump with a standard flow of 0.3 liters per minute, so the particles get into an inlet and get collected into the filter.
 In total, six hundred and forty eight sample filters are available, where eighteen correspond to field blanks.
@@
 
-## Concentration estimates
+### Concentration estimates
 
+The filters were pre-weighed (before sampling) in a Sartorius cubis microbalance, and then post-weighed in the same instrument after sampling.
+The difference in mass deposition was calculated as the post-pre weigh difference.
 The average 24-hour PM~2.5~ concentrations were calculated using the following equation:
@@
 
-The black carbon 24-hour concentrations were calculated using the following equation:
+Black carbon was estimated measuring transmissometry via a Magee Sootscan instrument. The black carbon 24-hour concentrations were calculated using the following equation:
 
@@
 
-The concentration of the chemical species (elements) on the filters was determined using X-Ray Fluorescence (XRF), and the uncertainties were estimated also based on the XRF instrument.
+The concentration of 22 chemical species (elements) on the filters were determined using X-Ray Fluorescence (XRF), and the uncertainties were estimated also based on the XRF instrument.
 
-## Concentration and Uncertainties data processing
+### Concentration and Uncertainties data processing
 
@@
 
-## Statistical analysis
+### Statistical analysis
 
 The correlation (r^2^) between chemical species will be determined using either individual paired correlations or by a correlation plot matrix.
-Additional plots, such as boxplots, to visualize and compare how the concentrations differ by type of fuel and by study arm.
-These differences will be assessed for statistical significance.
-After exploring the distribution that best fits the dataste, the most appropriate model will be implemented to explore what exposure variables are mostly associated with the concentrations of the most significant elemental species.
+Additional plots, such as boxplots, to visualize and compare how the concentrations differ by type of fuel.
+These differences will be assessed for statistical significance using a Welch two-sample T-test.
+Multivariate regression models will be computed to analyze which exposures could explain the concentrations of the key chemical species.
+Given that the concentrations can be approximated to a gamma distribution, the generalized linear model with gamma distribution will be used to compute the models.
+The root mean square error (RMSE) and R-squared will be the parameters to evaluate the most fit model for each of the chemical species.
 
-# Preliminary Results
+## Preliminary Results
 
-## Description of data
+Six hundred and twenty-nine filter samples were considered in the final analyses.
+@fig-stats1 presents a summary of the concentrations from the biomass fueltype samples.
+Meanwhile @fig-stats2 shows the summary of concentrations from the LPG observations.
+As a note, all the observations of the baseline measurements were categorized as biomass fuel type, since at this point, the LPG intervention was not yet allocated.
 
-1.  Concentrations:
+![Summary statistics of the concentrations in the biomass group](../../results/tables/table1.jpeg){#fig-stats1}
 
-As seen in the following glimpse from the cleaned `concentration` dataframe, it contains the ID of the 630 sample filters, along with the concentrations of 11 chemical species (Mg, Al, Si, S, K, Ca, Ti, Mn, Fe, Zn, and BC), and to which Arm and Fueltype they belong.
+![Summary statistics of the concentrations in the LPG group](../../results/tables/table2.jpeg){#fig-stats2}
 
-```{r}
-concentration <- read_rds(here("data", "processed-data", "concentration.rds"))
+### Chemical Species correlations
 
-glimpse(concentration)
-```
+In terms of correlations, Spearman correlation was estimated between all the chemical species and @fig-correlations shows the associations highlighting the R^2^ value.
+It is observed that the highest correlation is observed between Si and Fe, Ti-Fe, and Al-Si.
+These values could indicate that these species come from the same sources, so it will probably be easier to identify using the PMF software after the conclusion of this project.
 
-2.  Uncertainties:
-
-The cleaned `uncertainty` dataframe contains the uncertainty of the 11 chemical species. It contains the same name of variables as the `concentration` dataframe.
-
-```{r}
-uncertainty <- read_rds(here("data", "processed-data", "uncertainty.rds"))
-
-glimpse(uncertainty)
-```
-
-3.  Other variables from the HAPIN study:
-
-The cleaned `hapin_samples` dataframe contains information of the PM~2.5~, BC and CO (carbon monoxide) concentrations, besides other important recorded exposures, as seen from the glimpse below.
-It also contains the filter IDs, the arm and fueltype.
-
-```{r}
-hapin_samples <- read_rds(here("data", "processed-data", "hapin_samples.rds"))
-
-glimpse(hapin_samples)
-```
-
-## Exploratory data analysis
+![Correlation plot for chemical species](../../results/figures/corplot.jpg){#fig-correlations}
 
@@
 The remaining chemical species have lower concentrations, ranging from less than 0.1 ug/m^3^ to 10 ug/m^3^.
-Other noticeable patterns is that some of the species were found in higher concentrations for personal samples from participants that used biomass stoves, compared to LPG stoves (@fig-concentrations), such as BC, K, and Ca, and similar patterns were observed when categorized by intervention arm (Supplementary material).
-The following step will be to determine which of these chemical species have statistically significant differences among both comparison groups.
-
-In terms of correlations, Spearman correlation was estimated between all the chemical species and @fig-correlations, shows the associations highlighting the R^2^ value.
-It is observed that the highest correlation is observed between Si and Fe, Ti-Fe, and Al-Si.
-These values could indicate that these species come from the same source, so it will probably be easier to identify using the PMF software after the conclusion of this project.
-
-It is important to remember that the concentrations of the chemical species were recalculated as referenced in the `data_cleaning_code.qmd` file, alongside with the uncertainties.
-A summary statistics table of the uncertainties can be found in the `data_cleaning_code.qmd` code.
+Other noticeable patterns is that some of the species were found in higher concentrations for personal samples from participants that used biomass stoves, compared to LPG stoves (@fig-concentrations).
+When conducting the two-sided T-test, it was determined that all of the chemical species, except Fe, Al and Zn were statistically significantly different between type of fuel.
+The differences in concentrations were also analyzed by study arm, however, given that all baseline measurements belonged to biomass fueltype observations, these results were not presented but can be observed in the Supplementary material.
 
@@
 
-![Correlation plot for chemical species](../../results/figures/corplot.jpg){#fig-correlations}
+### Chemical species prediction
 
-### Other Exposure variables
+Nine of the eleven exposures were evaluated as predictors of the chemical species concentrations.
+The use of a generator machine (for corn mill grinding) and incence were left out since these were the categorical exposure variables with the less variability in answers.
+The gamma distributed generalized linear models were chosen given the concentrations of all the species do not follow a normal distribution.
 
-From the `hapin_samples` dataframe, it was observed that there are some missing values for most of the exposure variables according to the Missing variables plot (@fig-missing).
-The variable with more missing values (\>30) was the CO concentrations.
-Given that this variable had the highest number of missing concentrations, it may not be ideal to include it as part of the final analysis.
-Other missing values are from the recorded exposures such as kerosene, trash burning, stove use, external stoves ('stove_other'), and others.
-It was also observed that there is one observation without BC concentration.
-This observation could be dropped out since the PMF analysis after the conclusion of this project requires all the concentrations and uncertainties to be available (No NA/missing values).
-
-The bar plot (@fig-categorical), shows that most of the participants recorded little exposures, with the exception of stove use at home ('stove'), stoves from other homes ('stove_other') and trash burning ('trash').
-Having not so many recorded exposures other than the previously mentioned, could make the analysis difficult because of sample size or any other statistical adjustments will have to occur before applying a regression model analysis approach.
-
-@fig-pm shows the distribution of PM~2.5~ concentrations by type of fuel.
-It is observed that the concentrations do not follow a normal distribution and it seems like, as expected, the LPG stove users samples have lower concentrations than the biomass stove users.
-The similar trends are observed for the CO and BC concentrations (Supplementary material).
-
-![Plot for missing values for the `hapin_samples` dataframe](../../results/figures/missing-values.jpg){#fig-missing}
-
-![Distribution of responses by the Categorical variables](../../results/figures/cat-exposures.jpg){#fig-categorical}
-
-![PM~2.5~ Concentrations](../../results/figures/pm25.jpg){#fig-pm}
+Based on the models, type of fuel was a strong predictor of concentration for eight chemical species: Mg, Mn, BC, Ca, K, S, Si and Ti.
+Kerosene was another predictor of BC and Mg concentrations alongside type of fuel and being exposed to other sources of smoke (such as traffic emissions).
+Meanwhile, smoking and other sources of smoke were high predictors of K.
+On the other hand, stove usage, independently if it is biomass or LPG, was associated with concentration of Fe.
+However, additional analyses have to be conducted for final model selection.
 
@@
 
-# References
-
-1.  Gordon, S. B. et al. Respiratory risks from household air pollution in low and middle income countries. The Lancet Respiratory Medicine 2, 823–860 (2014).
-2.  GBD. Global Burden of Disease Study 2019 (GBD 2019) Results. http://ghdx.healthdata.org/gbd-results-tool (2020).
-3.
-Murray, C.
-J.
-L.
-et al.
-Global burden of 87 risk factors in 204 countries and territories, 1990–2019: a systematic analysis for the Global Burden of Disease Study 2019.
-The Lancet 396, 1223–1249 (2020).
-4.  Pope, C. A. & Dockery, D. W. Health Effects of Fine Particulate Air Pollution: Lines that Connect. Journal of the Air & Waste Management Association 56, 709–742 (2006).
-5.
-Clasen, T.
-et al.
-Design and Rationale of the HAPIN Study: A Multicountry Randomized Controlled Trial to Assess the Effect of Liquefied Petroleum Gas Stove and Continuous Fuel Distribution.
-Environ Health Perspect 128, 047008 (2020).
-6.  Mohammed, M. O. A. et al. Potential Toxicological and Cardiopulmonary Effects of PM2.5 Exposure and Related Mortality: Findings of Recent Studies Published during 2003-2013. Biomed Environ Sci.
-7.  Viana, M. et al. Source apportionment of particulate matter in Europe: A review of methods and results. Journal of Aerosol Science 39, 827–849 (2008).
-8.  Zhou, Z. et al. Chemical Characterization and Source Apportionment of Household Fine Particulate Matter in Rural, Peri-urban, and Urban West Africa. Environ. Sci. Technol. 48, 1343–1351 (2014).
-9.  Sharma, S. K. et al. Source Apportionment of PM2.5 in Delhi, India Using PMF Model. Bull Environ Contam Toxicol 97, 286–293 (2016).
-10.
-Martínez-Cinco, M., Santos-Guzmán, J.
-& Mejía-Velázquez, G.
-Source apportionment of PM 2.5 for supporting control strategies in the Monterrey Metropolitan Area, Mexico.
-Journal of the Air & Waste Management Association 66, 631–642 (2016).
-11.
-Clasen, T., Checkley, W., Peel, J.
-L., Balakrishnan, K., McCracken, J.
-P., Rosa, G., Thompson, L.
-M., Barr, D.
-B., Clark, M.
-L., Johnson, M.
-A., Waller, L.
-A., Jaacks, L.
-M., Steenland, K., Miranda, J.
-J., Chang, H.
-H., Kim, D.-Y., McCollum, E.
-D., Davila-Roman, V.
-G., Papageorghiou, A., … HAPIN Investigators.
-Design and Rationale of the HAPIN Study: A Multicountry Randomized Controlled Trial to Assess the Effect of Liquefied Petroleum Gas Stove and Continuous Fuel Distribution.
-Environmental Health Perspectives, 128(4), 047008.
-(2020).
-12.
-Johnson, M.
-A., Steenland, K., Piedrahita, R., Clark, M.
-L., Pillarisetti, A., Balakrishnan, K., Peel, J.
-L., Naeher, L.
-P., Liao, J., Wilson, D., Sarnat, J., Underhill, L.
-J., Burrowes, V., McCracken, J.
-P., Rosa, G., Rosenthal, J., Sambandam, S., de Leon, O., Kirby, M.
-A., … HAPIN Investigators.
-Air Pollutant Exposure and Stove Use Assessment Methods for the Household Air Pollution Intervention Network (HAPIN) Trial.
-Environmental Health Perspectives, 128(4), 047009.
-(2020).
-13. Kearns, K. A. Exposures to household air pollution in women across four diverse countries as part of a multi-country cookstove intervention trial. (2022).
-
-{{< pagebreak >}}
-
-# Supplementary material
-
-![Chemical species concentrations by study arm](../../results/figures/boxplot-arm.jpg)
-
-![Carbon monoxide (CO) Concentrations](../../results/figures/co.jpg)
-
-![Black carbon (BC) Concentrations](../../results/figures/bc.jpg)
+## References

### 2024-03-15 products/manuscript/supplement/Supplementary.qmd
Project Part 3
Created a new QMD file: data_analysis_code, and a Supplementary Material word document. Also, the manuscript file has been updated to include the preliminary results. ReadMe files updated and figures/tables edited.

--- 2024-03-15 code/data_analysis_code.qmd
@@
 ---
-title: "Project Part 3"
-subtitle: "Data analysis code"
+title: "Supplementary Material"
+subtitle: "Project: Part 3"
 author: Erick E. Mollinedo
 date: '`r format(Sys.Date(), "%B %d, %Y")`'
-format: html
-editor: visual
+format:
+  docx:
+    toc: false
+    number-sections: true
+    highlight-style: github
 ---
 
-List of packages needed for the whole data processing/data analysis
-
-```{r}
+```{r include=FALSE}
+library(tidyverse)
+library(tidymodels)
 library(here)
-library(readxl)
-library(tidyverse)
-library(naniar)
-library(vtable)
-library(corrplot)
-library(RColorBrewer)
-library(ggpubr)
-library(rstatix)
+library(readr)
 ```
 
-##Data Cleaning
+# Supplementary Material
 
-### HAPIN Guatemala data cleaning
+The following document includes some of the results from the exploratory data analysis.
 
-```{r}
-#Load the files and assign them to independent data frames
-hapin <- read_csv(here("data", "raw-data", "HAPIN_EX17_Guatemala_20230728.csv"))
-xrflist <- read_csv(here("data", "raw-data", "sa_filters_emollinedo.csv"))
+## Supplementary Results
 
-#Select only the variables I am interested to work with
-hapin <- hapin %>% select(trt_blinded, timepoint, h41_m_ecm_fid, h41_b_ecm_fid, h41b_filter1, h41b_filter2, h42_cook, h42_smoke, h42_coil, h42_trash, h42_kero, h42_incence, h42_generator, h42_smoked, h42_crop, h42_smoky_kitc, ECM_grav_neph_conc_M, ECM_bc_conc_M, CO_avg_ppm_M)
+@fig-conc shows the summary of concentrations for each chemical species categorized by study arm.
+It is observed a similar pattern observed from the categorization by type of fuel, where Mg, Mn, Ca, K, BC, Ti, Si and S have statistically significant differences in concentrations (shown as ****).
 
-#Make sure all filter IDs are upper case, using the `mutate()` function with `toupper()`
-hapin <- hapin %>% 
-  mutate(h41_m_ecm_fid = toupper(h41_m_ecm_fid),
-         h41b_filter1 = toupper(h41b_filter1),
-         h41_b_ecm_fid = toupper(h41_b_ecm_fid),
-         h41b_filter2 = toupper(h41b_filter2))
+![Chemical species concentrations by study arm](../../../results/figures/boxplot-arm.jpg){#fig-conc}
 
-#Select all filters that match with the list of filters from XRF, using the `filter()` function.
-hapin <- hapin %>%
-  filter(h41_m_ecm_fid %in% xrflist$filter_id |
-           h41b_filter1 %in% xrflist$filter_id |
-           h41_b_ecm_fid %in% xrflist$filter_id |
-           h41b_filter2 %in% xrflist$filter_id)
+In terms of the categorical exposures @fig-cat summarizes the responses from all the exposures.
+It is observed that for the incence and generator variables there were just a few or none observations that recorded being exposed to these sources.
+Meanwhile, stove usage was the exposure most recorded.
 
-#Concatenate `trt_blinded` and `timepoint` into a new temporary variable
-hapin$concatenated <- paste(hapin$trt_blinded, hapin$timepoint)
+![Summary of categorical exposures](../../../results/figures/cat-exposures.jpg){#fig-cat}
 
-#Create a new variable `fueltype` that categorizes all the observations by type of fuel ('R' means the visit belongs to the 'Control' treatment and 'Q' for the 'Intervention' treatment)
-hapin <- mutate(hapin, fueltype = case_when(
-  endsWith(concatenated, "BL") ~ "Biomass", #BL corresponds to the baseline, where all the participants had a biomass type fuel stove
-  endsWith(concatenated, "R P1") ~ "Biomass",
-  endsWith(concatenated, "R P2") ~ "Biomass",
-  endsWith(concatenated, "Q P1") ~ "LPG",
-  endsWith(concatenated, "Q P2") ~ "LPG",
-  endsWith(concatenated, "R BLP1") ~ "Biomass",
-  endsWith(concatenated, "R P1P2") ~ "Biomass",
-  endsWith(concatenated, "R P2B1") ~ "Biomass",
-  endsWith(concatenated, "Q BLP1") ~ "LPG",
-  endsWith(concatenated, "Q P1P2") ~ "LPG",
-  endsWith(concatenated, "Q P2B1") ~ "LPG"))
+The following figures show the distribution of the concentrations of Carbon monoxide (@fig-co) and PM~2.5~ (@fig-pm), which were considered as covariates (CO) or independent variables (PM~2.5~) to be included in the modelling process with the categorical exposures.
 
-#Rename some of the variables and delete unnecessary ones
-hapin <- hapin %>% rename(filter_id = "h41_m_ecm_fid", dup1_filterid = "h41b_filter1", dup2_filterid = "h41b_filter2",
-                         blank_filterid = "h41_b_ecm_fid", arm = "trt_blinded", pm25 = "ECM_grav_neph_conc_M",
-                         bc = "ECM_bc_conc_M", co = "CO_avg_ppm_M", stove = "h42_cook", smoke = "h42_smoke",
-                         coil = "h42_coil", trash = "h42_trash", kerosene = "h42_kero", incense = "h42_incence",                                   generator = "h42_generator", smoky = "h42_smoked", crop = "h42_crop", stove_other = "h42_smoky_kitc") %>%
-  select(-c(concatenated, timepoint)) %>% 
-  mutate(arm = case_when(arm == 'R' ~ 'Control', arm == 'Q' ~ 'Intervention', TRUE ~ arm)) #Recode values from the 'arm' variable
+![Carbon Monoxide concentrations by type of fuel](../../../results/figures/co.jpg){#fig-co}
 
-#Creating the dataframe for personal sample filters, and deleting the other columns with unnecessary filter IDs
-hapin_personal <- hapin %>% filter(filter_id %in% xrflist$filter_id) %>% 
-  select(-c(blank_filterid, dup1_filterid, dup2_filterid))
+![PM2.5 concentrations by type of fuel](../../../results/figures/pm25.jpg){#fig-pm}
 
-#Creating the dataframe for duplicates in column 4, and deleting the other columns with unnecessary filter IDs, and renaming `dup1_filterid` to `filter_id`.
-hapin_dup1 <- hapin %>% filter(dup1_filterid %in% xrflist$filter_id) %>% 
-  select(-c(blank_filterid, filter_id, dup2_filterid)) %>% 
-  rename(filter_id = "dup1_filterid")
+The following tables show the results when conducting the generalized linear models using a single or multiple variables for the chemical species that can be interpreted by sources of exposure.
 
-#Creating the dataframe for duplicates in column 5, and deleting the other columns with unnecessary filter IDs, and renaming `dup2_filterid` to `filter_id`.
-hapin_dup2 <- hapin %>% filter(dup2_filterid %in% xrflist$filter_id) %>% 
-  select(-c(blank_filterid, filter_id, dup1_filterid)) %>% 
-  rename(filter_id = "dup2_filterid")
+```{r echo=FALSE}
+#Open the hapin_merged df
 
-#Creating the dataframe for blank filters, and deleting the other columns with unnecessary filter IDs, and renaming `blank_filterid` to `filter_id`.
-hapin_blanks <- hapin %>% filter(blank_filterid %in% xrflist$filter_id) %>% 
-  select(-c(filter_id, dup1_filterid, dup2_filterid)) %>% 
-  rename(filter_id = "blank_filterid")
-
-#Combining the `personal` and `dup` dataframes into a single one, using `bind_rows()`
-hapin_filters <- bind_rows(hapin_personal, hapin_dup1, hapin_dup2)
-
-#Save the samples and blanks dataframes to .RDS format files
-saveRDS(hapin_filters, file = here("data", "processed-data", "hapin_samples.rds"))
-saveRDS(hapin_blanks, file = here("data", "processed-data", "hapin_blanks.rds"))
-```
-
-### Chemical species Concentrations data cleaning
-
-```{r}
-#Loading the chemical species concentration data
-concentration <- read_excel(here("data", "raw-data", "conc_unc_hapingt_emollinedo.xlsx"), sheet = "Concentration")
-
-#Loading the chemical species limit of detection values
-lod <- read_excel(here("data", "raw-data", "conc_unc_hapingt_emollinedo.xlsx"), sheet = "DL")
-
-#having a glimpse of this dataframe
-glimpse(concentration)
-
-#Remove brackets, convert to numeric, and round to 4 decimal places
-concentration <- concentration %>%
-  mutate(across(3:26, ~ round(as.numeric(str_remove_all(., "\\[|\\]")), 4))) %>% 
-  rename(filter_id = "Filter")
-
-#Create dataframe with the concentrations for the blank filters for each element
-blanks <- concentration %>% filter(filter_id %in% hapin_blanks$filter_id) %>% 
-  select(-c(filter_id, Type))
-
-#Remove blank filters from the `concentration` dataframe
-concentration <- concentration %>%
-  anti_join(hapin_blanks, by = "filter_id")
-
-#Calculate the median for each element, using the blank filter values
-median_blanks <- blanks %>% select(Mg: Na) %>%
-  map_dbl(median) %>% round(4)
-
-median_blanks
-
-#Round the limit of detection values to 4 decimal places
-lod <- lod %>% mutate(`Concentration (ug/cm2)` = round(`Concentration (ug/cm2)`, 4))
-print(lod)
-
-#Estimate the percentage of values below detection limit for each element
-bdl <- concentration %>%
-  summarise(
-    percent_bdl_mg = sum(Mg < 0.0051) / n() * 100,
-    percent_bdl_al = sum(Al < 0.0059) / n() * 100,
-    percent_bdl_si = sum(Si < 0.0068) / n() * 100,
-    percent_bdl_s = sum(S < 0.0026) / n() * 100,
-    percent_bdl_k = sum(K < 0.0040) / n() * 100,
-    percent_bdl_ca = sum(Ca < 0.0040) / n() * 100,
-    percent_bdl_ti = sum(Ti < 0.0086) / n() * 100,
-    percent_bdl_cr = sum(Cr < 0.0055) / n() * 100,
-    percent_bdl_mn = sum(Mn < 0.0030) / n() * 100,
-    percent_bdl_fe = sum(Fe < 0.0043) / n() * 100,
-    percent_bdl_ni = sum(Ni < 0.0043) / n() * 100,
-    percent_bdl_cu = sum(Cu < 0.0080) / n() * 100,
-    percent_bdl_zn = sum(Zn < 0.0024) / n() * 100,
-    percent_bdl_ga = sum(Ga < 0.0049) / n() * 100,
-    percent_bdl_as = sum(As < 0.0062) / n() * 100,
-    percent_bdl_se = sum(Se < 0.0181) / n() * 100,
-    percent_bdl_cd = sum(Cd < 0.0932) / n() * 100,
-    percent_bdl_in = sum(In < 0.1089) / n() * 100,
-    percent_bdl_sn = sum(Sn < 0.1844) / n() * 100,
-    percent_bdl_te = sum(Te < 0.2166) / n() * 100,
-    percent_bdl_i = sum(I < 0.3149) / n() * 100,
-    percent_bdl_pb = sum(Pb < 0.0324) / n() * 100,)
-
-print(bdl)
-
-#Select only the filter ID, type and the 10 elements that have more than 50% of their data not below the detection limit
-concentration <- concentration %>% select(c(filter_id, Type, Mg, Al, Si, S, K, Ca, Ti, Mn, Fe, Zn)) %>% 
-  arrange(filter_id) #Also I decided to arrange them ascending based on the filter_id
-
-#Substract the median blank value for the elements that need blank adjustment
-concentration <- concentration %>%
-  mutate(Mg = Mg - 0.0112, S = S - 0.0044, K = K - 0.0014, Ca = Ca - 0.0048, Fe = Fe - 0.0038, Zn = Zn - 0.0045)
-
-#Replace all values below the detection limit with DL/sqrt(2)
-concentration <- concentration %>% 
-    mutate(Mg = round(if_else(Mg < 0.0051, 0.0051 / sqrt(2), Mg), 4), Al = round(if_else(Al < 0.0059, 0.0059 / sqrt(2), Al), 4),
-           Si = round(if_else(Si < 0.0068, 0.0068 / sqrt(2), Si), 4), S = round(if_else(S < 0.0026, 0.0026 / sqrt(2), S), 4),
-           K = round(if_else(K < 0.0016, 0.0016 / sqrt(2), K), 4), Ca = round(if_else(Ca < 0.0040, 0.0040 / sqrt(2), Ca), 4),
-           Ti = round(if_else(Ti < 0.0086, 0.0086 / sqrt(2), Ti), 4), Mn = round(if_else(Mn < 0.0030, 0.0030 / sqrt(2), Mn), 4),
-           Fe = round(if_else(Fe < 0.0043, 0.0043 / sqrt(2), Fe), 4), Zn = round(if_else(Zn < 0.0024, 0.0024 / sqrt(2), Zn), 4))
-
-#I will save these concentrations in a new dataframe `concentrations_ugcm2` that will be used for uncertainty estimations in future steps
-concentrations_ugcm2 <- concentration %>% select(-Type)
-
-#Converting the concentrations from (ug/cm2) to (ug/m3)
-concentration <- concentration %>%
-  mutate(across(Mg:Zn, ~ round(.x * 0.7853 / 0.43, 4)))
-
-#Now adding the BC concentrations value from the `hapin_filters` df
-hapin_filters <- hapin_filters %>% arrange(filter_id) #First arrange in ascending order by filter_id
-concentration <- bind_cols(concentration, hapin_filters$bc) %>% rename(BC = "...13") %>% select(-Type) #Join the BC concentrations to the `concentrations` df, also renaming the column to "BC" and deleting the 'Type' column
-
-#Add the `arm` and `fueltype` variables to the `concentration` dataframe
-concentration <- concentration %>%
-  mutate(arm = hapin_filters$arm, fueltype = hapin_filters$fueltype)
-
-#Check if there are any missing values from the concentration file
-gg_miss_var(concentration)
-
-#Use na.omit() to remove the spotted observation with missing value
-concentration <- na.omit(concentration)
-
-#Save this dataframe as RDS file
-saveRDS(concentration, file = here("data", "processed-data", "concentration.rds"))
-```
-
-### Chemical species Uncertainties data processing
-
-```{r}
-#Load the uncertainty data
-uncertainty <- read_excel(here("data", "raw-data", "conc_unc_hapingt_emollinedo.xlsx"), sheet = "Uncertainty")
-
-#Remove the chemical species with more than 50% of the observations with values BDL
-uncertainty <- uncertainty %>% select(c(Filter, Mg, Al, Si, S, K, Ca, Ti, Mn, Fe, Zn)) %>% #Select only the necessary variables
-  rename(filter_id = "Filter") %>% #Rename the Filter column
-  mutate_at(vars(Mg:Zn), ~ round(., 4)) %>% #Round the uncertainties to 4 decimal places
-  arrange(filter_id) %>% #Arrange observations based on Filter ID
-  anti_join(hapin_blanks, by = "filter_id") #Remove blank filters
-
-#Have a glimpse of the data
-glimpse(uncertainty)
-
-#Create a list for the detection limits for each element
-dl <- c(Mg = 0.0051, Al = 0.0059, Si = 0.0068, S = 0.0026, K = 0.0016, 
-        Ca = 0.0040, Ti = 0.0086, Mn = 0.0030, Fe = 0.0043, Zn = 0.0024)
-
-# Recalculate the values in df based on the conditions
-uncertainty <- uncertainty %>%
-  mutate(across(all_of(names(dl)),
-                ~ round(if_else(concentrations_ugcm2[[cur_column()]] < dl[[cur_column()]], (5/6) * dl[[cur_column()]], .), 4),
-                .names = "{.col}"))
-
-# Define the variables to be recalculated
-variables <- c('Mg', 'Al', 'Si', 'S', 'K', 'Ca', 'Ti', 'Mn', 'Fe', 'Zn')
-
-# Recalculate the values using the given equation
-uncertainty <- uncertainty %>%
-  mutate(across(all_of(variables), ~ round( #Here, selecting all the variables to be re-calculated
-    sqrt((0.7853/0.43)^2 * .^2 + (concentrations_ugcm2[[cur_column()]]/0.43)^2 * 0.05^2 + (-concentrations_ugcm2[[cur_column()]]*0.7853/0.43^2)^2 * (0.05 * 0.43)^2), 4), .names = "{.col}")) #Input the equation
-
-#Load file to estimate BC uncertainties
-atn <- read_csv(here("data", "raw-data", "ATN_summary.csv"))
-
-#Explore the structure of the file
-glimpse(atn)
-
-#Estimate total attenuation by sample
-atn <- atn %>% mutate(ATN = round(log(post_ATN/pre_ATN) *100, 2))
-
-#I will remove the observation with `filter_id` 3M53864 from the `uncertainty` df, since this observation does not have a BC concentration available
-uncertainty <- uncertainty %>% filter(filter_id != "3M53864") %>% 
-  mutate(BC = round((1370 * sqrt((atn$ATN)^2 * 0.000005^2 + 0.000079^2 * 1.16^2)), 4))#Estimate the BC uncertainties
-
-#I will remove the observation with `filter_id` 3M53864 from the `hapin_filters` df
-hapin_filters <- hapin_filters %>% filter(filter_id != "3M53864")
-
-#Add the `arm` and `fueltype` variables to the `uncertainty` dataframe
-uncertainty <- uncertainty %>%
-  mutate(arm = hapin_filters$arm, fueltype = hapin_filters$fueltype)
-
-#Save this dataframe as RDS file
-saveRDS(uncertainty, file = here("data", "processed-data", "uncertainty.rds"))
-```
-
-## Data Exploring
-
-First, I produced two boxplots, that summarize the concentrations categorized by `arm` and `fueltype`.
-First I will create a dataframe in longer format using `pivot_longer()`.
-Then creating the plots using `ggplot()`.
-In addition, I will add the significance test value to the plots to visualize which chemical species are statistically different among groups.
-
-```{r}
-#Pivot longer the `concentration` df to make the boxplots
-conc_longer <- concentration %>% #First a df with the `arm` variable
-  pivot_longer(cols = -c(filter_id, fueltype, arm),
-               names_to = "species", values_to = "concentration")
-
-#conduct the t-test of concentration-fueltype
-stat.test <- conc_longer %>% 
-  group_by(species) %>% 
-  t_test(concentration ~ fueltype) %>% 
-  add_significance() #to add the significance values
-
-#Add the position coordinates
-stat.test <- stat.test %>% 
-  add_x_position(x= "species", dodge = 0.8) %>% #x position is species, and the length of significance bar is 0.8
-  rstatix::add_y_position(y.trans = function(x){log10(x)}) #add y position, in this case specify using a function that it is in log10 scale
-
-#Creating the boxplot of concentrations by fueltype
-options(scipen = 999) #to remove scientific notation for smaller values (i.e. changes from 1*10-3 to 0.001)
-ggplot(conc_longer)+
-  geom_boxplot(aes(x= species, y= concentration, fill= fueltype))+
-  scale_fill_brewer(palette = "Paired")+
-  labs(x= "Chemical species", y= "Concentration (ug/m3)")+
-  scale_y_log10()+
-  guides(fill= guide_legend(title= "Fuel type"))+ #Change legend from fuel_type to Fuel type
-  theme_bw(base_size = 15)+
-  stat_pvalue_manual(stat.test, label = "p.signif", tip.length = 0.001) #Add the significance bars, use 'p.signif' to add *** or 'ns'
-
-#Creating the boxplot of concentrations by arm
-options(scipen = 999) #to remove scientific notation for smaller values (i.e. changes from 1*10-3 to 0.001)
-ggplot(conc_longer)+
-  geom_boxplot(aes(x= species, y= concentration, fill= arm))+
-  scale_fill_brewer(palette = "Paired")+
-  labs(x= "Chemical species", y= "Concentration (ug/m3)")+
-  scale_y_log10()+
-  guides(fill= guide_legend(title= "Study Arm"))+ #Change legend from fuel_type to Fuel type
-  theme_bw(base_size = 15)+
-  stat_pvalue_manual(stat.test, label = "p.signif", tip.length = 0.001) #Add the significance bars, use 'p.signif' to add *** or 'ns'
-```
-
-The boxplots for `fueltype` and `arm` show what I was expecting in terms of how the data is distributed.
-The concentrations seem higher in some elemental especies such as BC, Ca, K, Mg, Mn and S.
-The following steps will be to evaluate if those differences in concentrations are statistically different.
-
-And now producing summary statistics by type of fuel, using the `sumtable()` function.
-
-```{r}
-#Produce summary statistics by type of fuel
-#Remove arm and fueltype columns
-conc <- concentration %>% select(-arm)
-
-#Using the vtable package, which creates a table
-sumtable(conc, group = 'fueltype', group.long = T, add.median = T)
-```
-
-Now I created a correlation plot using the `corrplot()` function, to explore which elements could be more associated between each other.
-
-```{r}
-#First, delete unnecessary variables to conduct the correlation test
-cor <- concentration %>% select(-c(filter_id, fueltype, arm))
-
-#Perform the spearman correlation test
-cor <- stats::cor(cor, method = "pearson", use = "complete.obs") #I had to input the `use=` argument since there are missing values in the BC variable. I will correct them later.
-
-#Create the correlation plot
-corrplot(cor, method = "color", type = "lower", #Color form and displays at the lower portion
-         col = COL2("RdYlBu", 20), order = "hclust", #Color palette, and order or the species
-         tl.col = "orangered4", tl.srt = 0, tl.cex = 0.9,  #Color of the axis, position and size
-         addCoef.col = "gray10", number.font = 2, number.cex = 0.8, #Display correlation number, set the font and size
-         col.lim = c(0, 1), is.corr = T) #Set the color limits at the bar
-```
-
-The correlation plot above shows that Si and Fe are strongly correlated, which might suggest that these elements represent one specific source of pollution.
-Other high correlations observed are between Al and Si, and Fe and Ti.
-
-I produced a summary table of the uncertainties, using the `vtable` package.
-
-```{r}
-#Remove the `arm` variable
-unc_fuel <- uncertainty %>% select(-arm)
-
-#Use the `sumtable()` function from the `vtable` package, which creates a table
-sumtable(unc_fuel, group = 'fueltype', group.long = T, add.median = T, out = 'kable')
-
-#Remove the `fueltype` variable
-unc_arm <- uncertainty %>% select(-fueltype)
-
-#Use the `sumtable()` function from the `vtable` package, which creates a table
-sumtable(unc_arm, group = 'arm', group.long = T, add.median = T, out = 'kable')
-
-```
-
-The tables above are just a way to represent the summary statistics of the uncertainties, however, the uncertainties will only be useful for the PMF analysis after the conclusion of this project.
-One thing noticed also from this table, is that there are some infinite values from the BC concentrations, which also show that there are missing values that have to be corrected in earlier steps.
-From this tables we can also see that there are 376 observations from the 'Biomass' and 254 from the 'LPG' fueltypes.
-Meanwhile, there are 323 observations from the 'Control' arm and 307 from the 'Intervention' arm.
-
-Now, checking if there are missing values in the `hapin` dataframe for any of the observations using the `gg_miss_var()` function.
-
-```{r}
-gg_miss_var(hapin)
-```
-
-As seen above, there are some missing values from the observational exposures (kerosene to coil), there is just one missing observation for PM2.5 and BC concentrations and there is more than 30 missing observations for CO (carbon monoxide).
-The missing observation of BC might explain the missing value spotted before, so this observation might actually have to be dropped down.
-
-Now I created a barplot showing the distribution of responses for the categorical variables `kerosene`, `trash`, `stove_other`, `stove`, `smoky`, `incense`, `generator`, `smoke`, `crop`, and `coil`.
-First doing data manipulation so the df is useful for plotting.
-
-```{r}
-#Pivot longer the `hapin` df to make a barplot
-hapin_longer <- hapin %>%
-  select(c(`kerosene`, `trash`, `stove_other`, `stove`, `smoky`, `incense`, `generator`, `smoke`, `crop`, `coil`)) %>% #Here selecting only the necessary variables
-  pivot_longer(cols = everything(),
-               names_to = "exposure", values_to = "response") #Create the longer df
-
-#Create a simple barplot
-ggplot(hapin_longer, aes(x= exposure, fill= response))+
-  geom_bar(position = "dodge")+
-  theme_classic()
-```
-
-The bar plot shows that the majority of participants recorded to not being exposed to most of the exposures, except to `stove`, which is the stove that the participants use.
-Others also recorded that were exposed to other stoves, or to trash burning.
-
-And finally plotting the PM2.5, BC and CO concentrations in different graphs using `ggplot()`
-
-```{r}
-#Histogram plot for the PM2.5 concentrations
-ggplot(hapin, aes(x= pm25, fill= fueltype))+
-  geom_histogram()+
-  theme_classic()+
-  labs(x= "PM2.5 Concentration (ug/m3)")
-
-#Histogram plot for the BC concentrations
-ggplot(hapin, aes(x= bc, fill= fueltype))+
-  geom_histogram()+
-  theme_classic()+
-  labs(x= "Black Carbon Concentration (ug/m3)")
-
-#Histogram plot for the CO concentrations
-ggplot(hapin, aes(x= co, fill= fueltype))+
-  geom_histogram()+
-  theme_classic()+
-  labs(x= "Carbon Monoxide Concentration (ppm)")
-```
-
-The histograms for the concentration of the three pollutants show that they don't follow a normal distribution.
-Most of the concentrations are low, and it is also seen that overall the concentrations in the LPG fueltype are lower than the Biomass group.
-If these variables will be used for modeling purposes, they should be log-transformed or using a logistic regression approach.
-
-## Data Analysis
-
-To answer question 1, I will plot the correlations in a correlation plot using the `corrplot()` function.
-I used the previous code, since it was already developed, but editing to add more detail to the plot.
-
-```{r}
-#Create the correlation plot
-corrplot(cor, method = "color", type = "lower", #Color form and displays at the lower portion
-         col = COL2("RdYlBu", 20), order = "hclust", #Color palette, and order or the species
-         tl.col = "orangered4", tl.srt = 0, tl.cex = 1,  #Color of the axis, position and size
-         addCoef.col = "gray10", number.font = 2, number.cex = 0.8, #Display correlation number, set the font and size
-         col.lim = c(0, 1), is.corr = T) #Set the color limits at the bar
-```
-
-To answer question 2, I will conduct a T-test using the `tidymodels` package
-
-```{r}
-# Conduct T-tests for the chemical species based on 'fueltype'
-results <- map_dfr(2:12, function(i) { #To select columns that contain the variables to analyze
-  variable <- names(concentration)[i] #The dataframe where the name of the variables are
-  t_test <- t.test(reformulate('fueltype', response = variable), data = concentration) #Select the fueltype variable
-
-  tibble( #Format the tibble
-    variable = variable,
-    statistic = t_test$statistic,
-    p_value = t_test$p.value,
-    estimate = t_test$estimate,
-    null_value = t_test$null.value,
-    alternative = t_test$alternative,
-    method = t_test$method,
-    conf_low = t_test$conf.int[1],
-    conf_high = t_test$conf.int[2]
-  )
-})
-
-results #Print the results as a tibble
-```
-
-To answer question 3, I will conduct a regression analysis. But first I will merge the data from `hapin_filters` with the `concentration` data.
-
-```{r}
-#remove the 'arm' and 'fueltype' variables from the 'concentration' df
-conc2 <- concentration %>% select(-c(arm, fueltype))
-
-#Use 'inner_join()' to merge dataframes
-hapin_merged <- inner_join(conc2, hapin_filters, by= "filter_id")
-
-#Save new df as a .RDS file
-saveRDS(hapin_merged, file = here("data", "processed-data", "hapin-final.rds"))
-```
-
-Now conducting the model selection using the `tidymodels` package.
-For the models, I used the gamma distribution family, since the concentrations of each chemical species could be explained by the gamma distribution.
-I used all the recorded exposures except incence and generator, since these had low variability in the answers.
-
-```{r}
-#First, loading the `tidymodels` package
-library(tidymodels)
+hapin_merged <- read_csv(here("data", "processed-data", "hapin-merged.csv"))
 
@@
 
-###########To predict the BC variable##############
-# Create the recipe for preprocessing
-recipeBC <- recipe(BC ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-# Create a workflow
-workflowBC <- workflow() %>%
-  add_recipe(recipeBC) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitBC <- workflowBC %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsBC <- fitBC %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsBC)
-
-###########To predict the Ca variable##############
-# Create the recipe for preprocessing
-recipeCa <- recipe(Ca ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-# Create a workflow
-workflowCa <- workflow() %>%
-  add_recipe(recipeCa) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitCa <- workflowCa %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsCa <- fitCa %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsCa)
-
-###########To predict the K variable##############
-# Create the recipe for preprocessing
-recipeK <- recipe(K ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-# Create a workflow
-workflowK <- workflow() %>%
-  add_recipe(recipeK) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitK <- workflowK %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsK <- fitK %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsK)
-
-###########To predict the Mg variable##############
-# Create the recipe for preprocessing
-recipeMg <- recipe(Mg ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-# Create a workflow
-workflowMg <- workflow() %>%
-  add_recipe(recipeMg) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitMg <- workflowMg %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsMg <- fitMg %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsMg)
-
-###########To predict the Mn variable##############
-# Create the recipe for preprocessing
-recipeMn <- recipe(Mn ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-# Create a workflow
-workflowMn <- workflow() %>%
-  add_recipe(recipeMn) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitMn <- workflowMn %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsMn <- fitMn %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsMn)
-
-###########To predict the S variable##############
-# Create the recipe for preprocessing
-recipeSul <- recipe(S ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-# Create a workflow
-workflowSul <- workflow() %>%
-  add_recipe(recipeSul) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitSul <- workflowSul %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsSul <- fitSul %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsSul)
-
-###########To predict the Al variable##############
-# Create the recipe for preprocessing
-recipeAl <- recipe(Al ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-# Create a workflow
-workflowAl <- workflow() %>%
-  add_recipe(recipeAl) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitAl <- workflowAl %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsAl <- fitAl %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsAl)
-
-###########To predict the Fe variable##############
-# Create the recipe for preprocessing
-recipeFe <- recipe(Fe ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-# Create a workflow
-workflowFe <- workflow() %>%
-  add_recipe(recipeFe) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitFe <- workflowFe %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsFe <- fitFe %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsFe)
-
-###########To predict the Si variable##############
-# Create the recipe for preprocessing
-recipeSi <- recipe(Si ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-# Create a workflow
-workflowSi <- workflow() %>%
-  add_recipe(recipeSi) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitSi <- workflowSi %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsSi <- fitSi %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsSi)
-
-###########To predict the Ti variable##############
-# Create the recipe for preprocessing
-recipeTi <- recipe(Ti ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-# Create a workflow
-workflowTi <- workflow() %>%
-  add_recipe(recipeTi) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitTi <- workflowTi %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsTi <- fitTi %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsTi)
-
-###########To predict the Zn variable##############
-# Create the recipe for preprocessing
-recipeZn <- recipe(Zn ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-# Create a workflow
-workflowZn <- workflow() %>%
-  add_recipe(recipeZn) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitZn <- workflowZn %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsZn <- fitZn %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsZn)
-```
-
-Based on the previous models, it seems like the most important variables are `fueltype` (as expected), `kerosene`, `stove_other` and `smoke` depending on the chemical species.
-So I will test new regression models for all species (except Al and Zn, since they seem not to be explained by any of the exposures), using only their most significant predictors.
-
-```{r}
 ###########To predict the BC variable##############
@@
 ```
-
-In this part I will evaluate both models from each chemical species (TBD)
-
-```{r}
-
-```

### 2024-04-05 code/data_analysis_code.qmd
Updated the README file for running code
Project Part 4 complete
Done

@@
 
-List of packages needed for the whole data processing/data analysis
+## Data Analysis
+
+List of packages needed for the whole data processing and data analysis
 
@@
 library(tidyverse)
-library(naniar)
-library(vtable)
 library(corrplot)
@@
 library(rstatix)
+library(tidymodels)
 ```
 
-##Data Cleaning
-
-### HAPIN Guatemala data cleaning
+Open the `concentration` and `hapin_samples` .RDS files to conduct the analyses.
 
 ```{r}
-#Load the files and assign them to independent data frames
-hapin <- read_csv(here("data", "raw-data", "HAPIN_EX17_Guatemala_20230728.csv"))
-xrflist <- read_csv(here("data", "raw-data", "sa_filters_emollinedo.csv"))
-
-#Select only the variables I am interested to work with
-hapin <- hapin %>% select(trt_blinded, timepoint, h41_m_ecm_fid, h41_b_ecm_fid, h41b_filter1, h41b_filter2, h42_cook, h42_smoke, h42_coil, h42_trash, h42_kero, h42_incence, h42_generator, h42_smoked, h42_crop, h42_smoky_kitc, ECM_grav_neph_conc_M, ECM_bc_conc_M, CO_avg_ppm_M)
-
-#Make sure all filter IDs are upper case, using the `mutate()` function with `toupper()`
-hapin <- hapin %>% 
-  mutate(h41_m_ecm_fid = toupper(h41_m_ecm_fid),
-         h41b_filter1 = toupper(h41b_filter1),
-         h41_b_ecm_fid = toupper(h41_b_ecm_fid),
-         h41b_filter2 = toupper(h41b_filter2))
-
-#Select all filters that match with the list of filters from XRF, using the `filter()` function.
-hapin <- hapin %>%
-  filter(h41_m_ecm_fid %in% xrflist$filter_id |
-           h41b_filter1 %in% xrflist$filter_id |
-           h41_b_ecm_fid %in% xrflist$filter_id |
-           h41b_filter2 %in% xrflist$filter_id)
-
-#Concatenate `trt_blinded` and `timepoint` into a new temporary variable
-hapin$concatenated <- paste(hapin$trt_blinded, hapin$timepoint)
-
-#Create a new variable `fueltype` that categorizes all the observations by type of fuel ('R' means the visit belongs to the 'Control' treatment and 'Q' for the 'Intervention' treatment)
-hapin <- mutate(hapin, fueltype = case_when(
-  endsWith(concatenated, "BL") ~ "Biomass", #BL corresponds to the baseline, where all the participants had a biomass type fuel stove
-  endsWith(concatenated, "R P1") ~ "Biomass",
-  endsWith(concatenated, "R P2") ~ "Biomass",
-  endsWith(concatenated, "Q P1") ~ "LPG",
-  endsWith(concatenated, "Q P2") ~ "LPG",
-  endsWith(concatenated, "R BLP1") ~ "Biomass",
-  endsWith(concatenated, "R P1P2") ~ "Biomass",
-  endsWith(concatenated, "R P2B1") ~ "Biomass",
-  endsWith(concatenated, "Q BLP1") ~ "LPG",
-  endsWith(concatenated, "Q P1P2") ~ "LPG",
-  endsWith(concatenated, "Q P2B1") ~ "LPG"))
-
-#Rename some of the variables and delete unnecessary ones
-hapin <- hapin %>% rename(filter_id = "h41_m_ecm_fid", dup1_filterid = "h41b_filter1", dup2_filterid = "h41b_filter2",
-                         blank_filterid = "h41_b_ecm_fid", arm = "trt_blinded", pm25 = "ECM_grav_neph_conc_M",
-                         bc = "ECM_bc_conc_M", co = "CO_avg_ppm_M", stove = "h42_cook", smoke = "h42_smoke",
-                         coil = "h42_coil", trash = "h42_trash", kerosene = "h42_kero", incense = "h42_incence",                                   generator = "h42_generator", smoky = "h42_smoked", crop = "h42_crop", stove_other = "h42_smoky_kitc") %>%
-  select(-c(concatenated, timepoint)) %>% 
-  mutate(arm = case_when(arm == 'R' ~ 'Control', arm == 'Q' ~ 'Intervention', TRUE ~ arm)) #Recode values from the 'arm' variable
-
-#Creating the dataframe for personal sample filters, and deleting the other columns with unnecessary filter IDs
-hapin_personal <- hapin %>% filter(filter_id %in% xrflist$filter_id) %>% 
-  select(-c(blank_filterid, dup1_filterid, dup2_filterid))
-
-#Creating the dataframe for duplicates in column 4, and deleting the other columns with unnecessary filter IDs, and renaming `dup1_filterid` to `filter_id`.
-hapin_dup1 <- hapin %>% filter(dup1_filterid %in% xrflist$filter_id) %>% 
-  select(-c(blank_filterid, filter_id, dup2_filterid)) %>% 
-  rename(filter_id = "dup1_filterid")
-
-#Creating the dataframe for duplicates in column 5, and deleting the other columns with unnecessary filter IDs, and renaming `dup2_filterid` to `filter_id`.
-hapin_dup2 <- hapin %>% filter(dup2_filterid %in% xrflist$filter_id) %>% 
-  select(-c(blank_filterid, filter_id, dup1_filterid)) %>% 
-  rename(filter_id = "dup2_filterid")
-
-#Creating the dataframe for blank filters, and deleting the other columns with unnecessary filter IDs, and renaming `blank_filterid` to `filter_id`.
-hapin_blanks <- hapin %>% filter(blank_filterid %in% xrflist$filter_id) %>% 
-  select(-c(filter_id, dup1_filterid, dup2_filterid)) %>% 
-  rename(filter_id = "blank_filterid")
-
-#Combining the `personal` and `dup` dataframes into a single one, using `bind_rows()`
-hapin_filters <- bind_rows(hapin_personal, hapin_dup1, hapin_dup2)
-
-#Save the samples and blanks dataframes to .RDS format files
-saveRDS(hapin_filters, file = here("data", "processed-data", "hapin_samples.rds"))
-saveRDS(hapin_blanks, file = here("data", "processed-data", "hapin_blanks.rds"))
+concentration <- read_rds(here("data", "processed-data", "concentration.rds"))
+hapin_filters <- read_rds(here("data", "processed-data", "hapin_samples.rds"))
 ```
 
-### Chemical species Concentrations data cleaning
-
-```{r}
-#Loading the chemical species concentration data
-concentration <- read_excel(here("data", "raw-data", "conc_unc_hapingt_emollinedo.xlsx"), sheet = "Concentration")
-
-#Loading the chemical species limit of detection values
-lod <- read_excel(here("data", "raw-data", "conc_unc_hapingt_emollinedo.xlsx"), sheet = "DL")
-
-#having a glimpse of this dataframe
-glimpse(concentration)
-
-#Remove brackets, convert to numeric, and round to 4 decimal places
-concentration <- concentration %>%
-  mutate(across(3:26, ~ round(as.numeric(str_remove_all(., "\\[|\\]")), 4))) %>% 
-  rename(filter_id = "Filter")
-
-#Create dataframe with the concentrations for the blank filters for each element
-blanks <- concentration %>% filter(filter_id %in% hapin_blanks$filter_id) %>% 
-  select(-c(filter_id, Type))
-
-#Remove blank filters from the `concentration` dataframe
-concentration <- concentration %>%
-  anti_join(hapin_blanks, by = "filter_id")
-
-#Calculate the median for each element, using the blank filter values
-median_blanks <- blanks %>% select(Mg: Na) %>%
-  map_dbl(median) %>% round(4)
-
-median_blanks
-
-#Round the limit of detection values to 4 decimal places
-lod <- lod %>% mutate(`Concentration (ug/cm2)` = round(`Concentration (ug/cm2)`, 4))
-print(lod)
-
-#Estimate the percentage of values below detection limit for each element
-bdl <- concentration %>%
-  summarise(
-    percent_bdl_mg = sum(Mg < 0.0051) / n() * 100,
-    percent_bdl_al = sum(Al < 0.0059) / n() * 100,
-    percent_bdl_si = sum(Si < 0.0068) / n() * 100,
-    percent_bdl_s = sum(S < 0.0026) / n() * 100,
-    percent_bdl_k = sum(K < 0.0040) / n() * 100,
-    percent_bdl_ca = sum(Ca < 0.0040) / n() * 100,
-    percent_bdl_ti = sum(Ti < 0.0086) / n() * 100,
-    percent_bdl_cr = sum(Cr < 0.0055) / n() * 100,
-    percent_bdl_mn = sum(Mn < 0.0030) / n() * 100,
-    percent_bdl_fe = sum(Fe < 0.0043) / n() * 100,
-    percent_bdl_ni = sum(Ni < 0.0043) / n() * 100,
-    percent_bdl_cu = sum(Cu < 0.0080) / n() * 100,
-    percent_bdl_zn = sum(Zn < 0.0024) / n() * 100,
-    percent_bdl_ga = sum(Ga < 0.0049) / n() * 100,
-    percent_bdl_as = sum(As < 0.0062) / n() * 100,
-    percent_bdl_se = sum(Se < 0.0181) / n() * 100,
-    percent_bdl_cd = sum(Cd < 0.0932) / n() * 100,
-    percent_bdl_in = sum(In < 0.1089) / n() * 100,
-    percent_bdl_sn = sum(Sn < 0.1844) / n() * 100,
-    percent_bdl_te = sum(Te < 0.2166) / n() * 100,
-    percent_bdl_i = sum(I < 0.3149) / n() * 100,
-    percent_bdl_pb = sum(Pb < 0.0324) / n() * 100,)
-
-print(bdl)
-
-#Select only the filter ID, type and the 10 elements that have more than 50% of their data not below the detection limit
-concentration <- concentration %>% select(c(filter_id, Type, Mg, Al, Si, S, K, Ca, Ti, Mn, Fe, Zn)) %>% 
-  arrange(filter_id) #Also I decided to arrange them ascending based on the filter_id
-
-#Substract the median blank value for the elements that need blank adjustment
-concentration <- concentration %>%
-  mutate(Mg = Mg - 0.0112, S = S - 0.0044, K = K - 0.0014, Ca = Ca - 0.0048, Fe = Fe - 0.0038, Zn = Zn - 0.0045)
-
-#Replace all values below the detection limit with DL/sqrt(2)
-concentration <- concentration %>% 
-    mutate(Mg = round(if_else(Mg < 0.0051, 0.0051 / sqrt(2), Mg), 4), Al = round(if_else(Al < 0.0059, 0.0059 / sqrt(2), Al), 4),
-           Si = round(if_else(Si < 0.0068, 0.0068 / sqrt(2), Si), 4), S = round(if_else(S < 0.0026, 0.0026 / sqrt(2), S), 4),
-           K = round(if_else(K < 0.0016, 0.0016 / sqrt(2), K), 4), Ca = round(if_else(Ca < 0.0040, 0.0040 / sqrt(2), Ca), 4),
-           Ti = round(if_else(Ti < 0.0086, 0.0086 / sqrt(2), Ti), 4), Mn = round(if_else(Mn < 0.0030, 0.0030 / sqrt(2), Mn), 4),
-           Fe = round(if_else(Fe < 0.0043, 0.0043 / sqrt(2), Fe), 4), Zn = round(if_else(Zn < 0.0024, 0.0024 / sqrt(2), Zn), 4))
-
-#I will save these concentrations in a new dataframe `concentrations_ugcm2` that will be used for uncertainty estimations in future steps
-concentrations_ugcm2 <- concentration %>% select(-Type)
-
-#Converting the concentrations from (ug/cm2) to (ug/m3)
-concentration <- concentration %>%
-  mutate(across(Mg:Zn, ~ round(.x * 0.7853 / 0.43, 4)))
-
-#Now adding the BC concentrations value from the `hapin_filters` df
-hapin_filters <- hapin_filters %>% arrange(filter_id) #First arrange in ascending order by filter_id
-concentration <- bind_cols(concentration, hapin_filters$bc) %>% rename(BC = "...13") %>% select(-Type) #Join the BC concentrations to the `concentrations` df, also renaming the column to "BC" and deleting the 'Type' column
-
-#Add the `arm` and `fueltype` variables to the `concentration` dataframe
-concentration <- concentration %>%
-  mutate(arm = hapin_filters$arm, fueltype = hapin_filters$fueltype)
-
-#Check if there are any missing values from the concentration file
-gg_miss_var(concentration)
-
-#Use na.omit() to remove the spotted observation with missing value
-concentration <- na.omit(concentration)
-
-#Save this dataframe as RDS file
-saveRDS(concentration, file = here("data", "processed-data", "concentration.rds"))
-```
-
-### Chemical species Uncertainties data processing
-
-```{r}
-#Load the uncertainty data
-uncertainty <- read_excel(here("data", "raw-data", "conc_unc_hapingt_emollinedo.xlsx"), sheet = "Uncertainty")
-
-#Remove the chemical species with more than 50% of the observations with values BDL
-uncertainty <- uncertainty %>% select(c(Filter, Mg, Al, Si, S, K, Ca, Ti, Mn, Fe, Zn)) %>% #Select only the necessary variables
-  rename(filter_id = "Filter") %>% #Rename the Filter column
-  mutate_at(vars(Mg:Zn), ~ round(., 4)) %>% #Round the uncertainties to 4 decimal places
-  arrange(filter_id) %>% #Arrange observations based on Filter ID
-  anti_join(hapin_blanks, by = "filter_id") #Remove blank filters
-
-#Have a glimpse of the data
-glimpse(uncertainty)
-
-#Create a list for the detection limits for each element
-dl <- c(Mg = 0.0051, Al = 0.0059, Si = 0.0068, S = 0.0026, K = 0.0016, 
-        Ca = 0.0040, Ti = 0.0086, Mn = 0.0030, Fe = 0.0043, Zn = 0.0024)
-
-# Recalculate the values in df based on the conditions
-uncertainty <- uncertainty %>%
-  mutate(across(all_of(names(dl)),
-                ~ round(if_else(concentrations_ugcm2[[cur_column()]] < dl[[cur_column()]], (5/6) * dl[[cur_column()]], .), 4),
-                .names = "{.col}"))
-
-# Define the variables to be recalculated
-variables <- c('Mg', 'Al', 'Si', 'S', 'K', 'Ca', 'Ti', 'Mn', 'Fe', 'Zn')
-
-# Recalculate the values using the given equation
-uncertainty <- uncertainty %>%
-  mutate(across(all_of(variables), ~ round( #Here, selecting all the variables to be re-calculated
-    sqrt((0.7853/0.43)^2 * .^2 + (concentrations_ugcm2[[cur_column()]]/0.43)^2 * 0.05^2 + (-concentrations_ugcm2[[cur_column()]]*0.7853/0.43^2)^2 * (0.05 * 0.43)^2), 4), .names = "{.col}")) #Input the equation
-
-#Load file to estimate BC uncertainties
-atn <- read_csv(here("data", "raw-data", "ATN_summary.csv"))
-
-#Explore the structure of the file
-glimpse(atn)
-
-#Estimate total attenuation by sample
-atn <- atn %>% mutate(ATN = round(log(post_ATN/pre_ATN) *100, 2))
-
-#I will remove the observation with `filter_id` 3M53864 from the `uncertainty` df, since this observation does not have a BC concentration available
-uncertainty <- uncertainty %>% filter(filter_id != "3M53864") %>% 
-  mutate(BC = round((1370 * sqrt((atn$ATN)^2 * 0.000005^2 + 0.000079^2 * 1.16^2)), 4))#Estimate the BC uncertainties
-
-#I will remove the observation with `filter_id` 3M53864 from the `hapin_filters` df
-hapin_filters <- hapin_filters %>% filter(filter_id != "3M53864")
-
-#Add the `arm` and `fueltype` variables to the `uncertainty` dataframe
-uncertainty <- uncertainty %>%
-  mutate(arm = hapin_filters$arm, fueltype = hapin_filters$fueltype)
-
-#Save this dataframe as RDS file
-saveRDS(uncertainty, file = here("data", "processed-data", "uncertainty.rds"))
-```
-
-## Data Exploring
-
-First, I produced two boxplots, that summarize the concentrations categorized by `arm` and `fueltype`.
-First I will create a dataframe in longer format using `pivot_longer()`.
-Then creating the plots using `ggplot()`.
-In addition, I will add the significance test value to the plots to visualize which chemical species are statistically different among groups.
-
-```{r}
-#Pivot longer the `concentration` df to make the boxplots
-conc_longer <- concentration %>% #First a df with the `arm` variable
-  pivot_longer(cols = -c(filter_id, fueltype, arm),
-               names_to = "species", values_to = "concentration")
-
-#conduct the t-test of concentration-fueltype
-stat.test <- conc_longer %>% 
-  group_by(species) %>% 
-  t_test(concentration ~ fueltype) %>% 
-  add_significance() #to add the significance values
-
-#Add the position coordinates
-stat.test <- stat.test %>% 
-  add_x_position(x= "species", dodge = 0.8) %>% #x position is species, and the length of significance bar is 0.8
-  rstatix::add_y_position(y.trans = function(x){log10(x)}) #add y position, in this case specify using a function that it is in log10 scale
-
-#Creating the boxplot of concentrations by fueltype
-options(scipen = 999) #to remove scientific notation for smaller values (i.e. changes from 1*10-3 to 0.001)
-ggplot(conc_longer)+
-  geom_boxplot(aes(x= species, y= concentration, fill= fueltype))+
-  scale_fill_brewer(palette = "Paired")+
-  labs(x= "Chemical species", y= "Concentration (ug/m3)")+
-  scale_y_log10()+
-  guides(fill= guide_legend(title= "Fuel type"))+ #Change legend from fuel_type to Fuel type
-  theme_bw(base_size = 15)+
-  stat_pvalue_manual(stat.test, label = "p.signif", tip.length = 0.001) #Add the significance bars, use 'p.signif' to add *** or 'ns'
-
-#Creating the boxplot of concentrations by arm
-options(scipen = 999) #to remove scientific notation for smaller values (i.e. changes from 1*10-3 to 0.001)
-ggplot(conc_longer)+
-  geom_boxplot(aes(x= species, y= concentration, fill= arm))+
-  scale_fill_brewer(palette = "Paired")+
-  labs(x= "Chemical species", y= "Concentration (ug/m3)")+
-  scale_y_log10()+
-  guides(fill= guide_legend(title= "Study Arm"))+ #Change legend from fuel_type to Fuel type
-  theme_bw(base_size = 15)+
-  stat_pvalue_manual(stat.test, label = "p.signif", tip.length = 0.001) #Add the significance bars, use 'p.signif' to add *** or 'ns'
-```
-
-The boxplots for `fueltype` and `arm` show what I was expecting in terms of how the data is distributed.
-The concentrations seem higher in some elemental especies such as BC, Ca, K, Mg, Mn and S.
-The following steps will be to evaluate if those differences in concentrations are statistically different.
-
-And now producing summary statistics by type of fuel, using the `sumtable()` function.
-
-```{r}
-#Produce summary statistics by type of fuel
-#Remove arm and fueltype columns
-conc <- concentration %>% select(-arm)
-
-#Using the vtable package, which creates a table
-sumtable(conc, group = 'fueltype', group.long = T, add.median = T)
-```
-
-Now I created a correlation plot using the `corrplot()` function, to explore which elements could be more associated between each other.
+To answer question 1, I will plot the correlations in a correlation plot using the `corrplot()` function.
+I used the previous code, since it was already developed, but editing to add more detail to the plot.
 
@@
 #Perform the spearman correlation test
-cor <- stats::cor(cor, method = "pearson", use = "complete.obs") #I had to input the `use=` argument since there are missing values in the BC variable. I will correct them later.
+cor <- stats::cor(cor, method = "pearson")
 
-#Create the correlation plot
-corrplot(cor, method = "color", type = "lower", #Color form and displays at the lower portion
-         col = COL2("RdYlBu", 20), order = "hclust", #Color palette, and order or the species
-         tl.col = "orangered4", tl.srt = 0, tl.cex = 0.9,  #Color of the axis, position and size
-         addCoef.col = "gray10", number.font = 2, number.cex = 0.8, #Display correlation number, set the font and size
-         col.lim = c(0, 1), is.corr = T) #Set the color limits at the bar
-```
-
-The correlation plot above shows that Si and Fe are strongly correlated, which might suggest that these elements represent one specific source of pollution.
-Other high correlations observed are between Al and Si, and Fe and Ti.
-
-I produced a summary table of the uncertainties, using the `vtable` package.
-
-```{r}
-#Remove the `arm` variable
-unc_fuel <- uncertainty %>% select(-arm)
-
-#Use the `sumtable()` function from the `vtable` package, which creates a table
-sumtable(unc_fuel, group = 'fueltype', group.long = T, add.median = T, out = 'kable')
-
-#Remove the `fueltype` variable
-unc_arm <- uncertainty %>% select(-fueltype)
-
-#Use the `sumtable()` function from the `vtable` package, which creates a table
-sumtable(unc_arm, group = 'arm', group.long = T, add.median = T, out = 'kable')
-
-```
-
-The tables above are just a way to represent the summary statistics of the uncertainties, however, the uncertainties will only be useful for the PMF analysis after the conclusion of this project.
-One thing noticed also from this table, is that there are some infinite values from the BC concentrations, which also show that there are missing values that have to be corrected in earlier steps.
-From this tables we can also see that there are 376 observations from the 'Biomass' and 254 from the 'LPG' fueltypes.
-Meanwhile, there are 323 observations from the 'Control' arm and 307 from the 'Intervention' arm.
-
-Now, checking if there are missing values in the `hapin` dataframe for any of the observations using the `gg_miss_var()` function.
-
-```{r}
-gg_miss_var(hapin)
-```
-
-As seen above, there are some missing values from the observational exposures (kerosene to coil), there is just one missing observation for PM2.5 and BC concentrations and there is more than 30 missing observations for CO (carbon monoxide).
-The missing observation of BC might explain the missing value spotted before, so this observation might actually have to be dropped down.
-
-Now I created a barplot showing the distribution of responses for the categorical variables `kerosene`, `trash`, `stove_other`, `stove`, `smoky`, `incense`, `generator`, `smoke`, `crop`, and `coil`.
-First doing data manipulation so the df is useful for plotting.
-
-```{r}
-#Pivot longer the `hapin` df to make a barplot
-hapin_longer <- hapin %>%
-  select(c(`kerosene`, `trash`, `stove_other`, `stove`, `smoky`, `incense`, `generator`, `smoke`, `crop`, `coil`)) %>% #Here selecting only the necessary variables
-  pivot_longer(cols = everything(),
-               names_to = "exposure", values_to = "response") #Create the longer df
-
-#Create a simple barplot
-ggplot(hapin_longer, aes(x= exposure, fill= response))+
-  geom_bar(position = "dodge")+
-  theme_classic()
-```
-
-The bar plot shows that the majority of participants recorded to not being exposed to most of the exposures, except to `stove`, which is the stove that the participants use.
-Others also recorded that were exposed to other stoves, or to trash burning.
-
-And finally plotting the PM2.5, BC and CO concentrations in different graphs using `ggplot()`
-
-```{r}
-#Histogram plot for the PM2.5 concentrations
-ggplot(hapin, aes(x= pm25, fill= fueltype))+
-  geom_histogram()+
-  theme_classic()+
-  labs(x= "PM2.5 Concentration (ug/m3)")
-
-#Histogram plot for the BC concentrations
-ggplot(hapin, aes(x= bc, fill= fueltype))+
-  geom_histogram()+
-  theme_classic()+
-  labs(x= "Black Carbon Concentration (ug/m3)")
-
-#Histogram plot for the CO concentrations
-ggplot(hapin, aes(x= co, fill= fueltype))+
-  geom_histogram()+
-  theme_classic()+
-  labs(x= "Carbon Monoxide Concentration (ppm)")
-```
-
-The histograms for the concentration of the three pollutants show that they don't follow a normal distribution.
-Most of the concentrations are low, and it is also seen that overall the concentrations in the LPG fueltype are lower than the Biomass group.
-If these variables will be used for modeling purposes, they should be log-transformed or using a logistic regression approach.
-
-## Data Analysis
-
-To answer question 1, I will plot the correlations in a correlation plot using the `corrplot()` function.
-I used the previous code, since it was already developed, but editing to add more detail to the plot.
-
-```{r}
 #Create the correlation plot
@@
 
-To answer question 3, I will conduct a regression analysis. But first I will merge the data from `hapin_filters` with the `concentration` data.
+To answer question 3, I will conduct a regression analysis.
+But first I will merge the data from `hapin_filters` with the `concentration` data.
+Modeling can be difficult if there are missing values, so I will additionally clean the `hapin_merged` df, so it removes all NAs.
 
@@
 hapin_merged <- inner_join(conc2, hapin_filters, by= "filter_id")
+
+#Remove pm25, bc and co, and additionally drop all NAs
+hapin_merged <- hapin_merged %>%
+  select(-c(pm25, bc, co)) %>%
+  drop_na()
 
@@
 ```{r}
-#First, loading the `tidymodels` package
-library(tidymodels)
+# Create a generalized linear regression model specification using gamma distribution (This will be used for all the models)
+gamma_spec <- linear_reg(mode = "regression") %>%
+  set_engine("glm", family = Gamma(link = "log"))
 
-# Create a generalized linear regression model specification using gamma distribution (This will be used for all the models)
-model_spec <- linear_reg(mode = "regression") %>%
-  set_engine("glm", family = Gamma(link = "log"))
+###########To predict the Al variable##############
+# Create the recipe for preprocessing
+recipeAl <- recipe(Al ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
+  step_dummy(all_nominal(), -all_outcomes())
+
+#Fit the model
+fitAl <- workflow() %>%
+  add_recipe(recipeAl) %>%
+  add_model(gamma_spec) %>%
+  fit(data = hapin_merged)
+
+# Extract and print the results of the model
+resultsAl <- fitAl %>%
+  pull_workflow_fit() %>%
+  tidy()
+
+print(resultsAl)
 
@@
 
-# Create a workflow
-workflowBC <- workflow() %>%
+#Fit the model
+fitBC <- workflow() %>%
   add_recipe(recipeBC) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitBC <- workflowBC %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 
-# Create a workflow
-workflowCa <- workflow() %>%
+#Fit the model
+fitCa <- workflow() %>%
   add_recipe(recipeCa) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitCa <- workflowCa %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 
+###########To predict the Fe variable##############
+# Create the recipe for preprocessing
+recipeFe <- recipe(Fe ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
+  step_dummy(all_nominal(), -all_outcomes())
+
+#Fit the model
+fitFe <- workflow() %>%
+  add_recipe(recipeFe) %>%
+  add_model(gamma_spec) %>%
+  fit(data = hapin_merged)
+
+# Extract and print the results of the model
+resultsFe <- fitFe %>%
+  pull_workflow_fit() %>%
+  tidy()
+
+print(resultsFe)
+
 ###########To predict the K variable##############
@@
 
-# Create a workflow
-workflowK <- workflow() %>%
+#Fit the model
+fitK <- workflow() %>%
   add_recipe(recipeK) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitK <- workflowK %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 
-# Create a workflow
-workflowMg <- workflow() %>%
+#Fit the model
+fitMg <- workflow() %>%
   add_recipe(recipeMg) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitMg <- workflowMg %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 
-# Create a workflow
-workflowMn <- workflow() %>%
+#Fit the model
+fitMn <- workflow() %>%
   add_recipe(recipeMn) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitMn <- workflowMn %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 
-# Create a workflow
-workflowSul <- workflow() %>%
+#Fit the model
+fitSul <- workflow() %>%
   add_recipe(recipeSul) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitSul <- workflowSul %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 
-###########To predict the Al variable##############
-# Create the recipe for preprocessing
-recipeAl <- recipe(Al ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-# Create a workflow
-workflowAl <- workflow() %>%
-  add_recipe(recipeAl) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitAl <- workflowAl %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsAl <- fitAl %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsAl)
-
-###########To predict the Fe variable##############
-# Create the recipe for preprocessing
-recipeFe <- recipe(Fe ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-# Create a workflow
-workflowFe <- workflow() %>%
-  add_recipe(recipeFe) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitFe <- workflowFe %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsFe <- fitFe %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsFe)
-
 ###########To predict the Si variable##############
@@
 
-# Create a workflow
-workflowSi <- workflow() %>%
+#Fit the model
+fitSi <- workflow() %>%
   add_recipe(recipeSi) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitSi <- workflowSi %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 
-# Create a workflow
-workflowTi <- workflow() %>%
+#Fit the model
+fitTi <- workflow() %>%
   add_recipe(recipeTi) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitTi <- workflowTi %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 
-# Create a workflow
-workflowZn <- workflow() %>%
+#Fit the model
+fitZn <- workflow() %>%
   add_recipe(recipeZn) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitZn <- workflowZn %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 Based on the previous models, it seems like the most important variables are `fueltype` (as expected), `kerosene`, `stove_other` and `smoke` depending on the chemical species.
-So I will test new regression models for all species (except Al and Zn, since they seem not to be explained by any of the exposures), using only their most significant predictors.
+I was worried of the low variability on responses for some of the variables so I am going to use only the variables `micro_exposures`, `macro_exposures`, `kerosene`, `stove_other` and `fueltype`.
 
 ```{r}
-###########To predict the BC variable##############
+###########To predict the Al variable##############
 # Create the recipe for preprocessing
-recipeBC_red <- recipe(BC ~ kerosene + stove_other + fueltype, data = hapin_merged) %>%
+recipeAl2 <- recipe(Al ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
   step_dummy(all_nominal(), -all_outcomes())
 
-# Create a workflow
-workflowBC_red <- workflow() %>%
-  add_recipe(recipeBC_red) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitBC_red <- workflowBC_red %>%
+#Fit the model
+fitAl2 <- workflow() %>%
+  add_recipe(recipeAl2) %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 # Extract and print the results of the model
-resultsBC_red <- fitBC_red %>%
+resultsAl2 <- fitAl2 %>%
   pull_workflow_fit() %>%
@@
 
-print(resultsBC_red)
+print(resultsAl2)
+
+###########To predict the BC variable##############
+# Create the recipe for preprocessing
+recipeBC2 <- recipe(BC ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
+  step_dummy(all_nominal(), -all_outcomes())
+
+#Fit the model
+fitBC2 <- workflow() %>%
+  add_recipe(recipeBC2) %>%
+  add_model(gamma_spec) %>%
+  fit(data = hapin_merged)
+
+# Extract and print the results of the model
+resultsBC2 <- fitBC2 %>%
+  pull_workflow_fit() %>%
+  tidy()
+
+print(resultsBC2)
 
@@
 # Create the recipe for preprocessing
-recipeCa_red <- recipe(Ca ~ fueltype, data = hapin_merged) %>%
+recipeCa2 <- recipe(Ca ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
   step_dummy(all_nominal(), -all_outcomes())
 
-# Create a workflow
-workflowCa_red <- workflow() %>%
-  add_recipe(recipeCa_red) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitCa_red <- workflowCa_red %>%
+#Fit the model
+fitCa2 <- workflow() %>%
+  add_recipe(recipeCa2) %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 # Extract and print the results of the model
-resultsCa_red <- fitCa_red %>%
+resultsCa2 <- fitCa2 %>%
   pull_workflow_fit() %>%
@@
 
-print(resultsCa_red)
+print(resultsCa2)
+
+###########To predict the Fe variable##############
+# Create the recipe for preprocessing
+recipeFe2 <- recipe(Fe ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
+  step_dummy(all_nominal(), -all_outcomes())
+
+#Fit the model
+fitFe2 <- workflow() %>%
+  add_recipe(recipeFe2) %>%
+  add_model(gamma_spec) %>%
+  fit(data = hapin_merged)
+
+# Extract and print the results of the model
+resultsFe2 <- fitFe2 %>%
+  pull_workflow_fit() %>%
+  tidy()
+
+print(resultsFe2)
 
@@
 # Create the recipe for preprocessing
-recipeK_red <- recipe(K ~ smoke + stove_other + fueltype, data = hapin_merged) %>%
+recipeK2 <- recipe(K ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
   step_dummy(all_nominal(), -all_outcomes())
 
-# Create a workflow
-workflowK_red <- workflow() %>%
-  add_recipe(recipeK_red) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitK_red <- workflowK_red %>%
+#Fit the model
+fitK2 <- workflow() %>%
+  add_recipe(recipeK2) %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 # Extract and print the results of the model
-resultsK_red <- fitK_red %>%
+resultsK2 <- fitK2 %>%
   pull_workflow_fit() %>%
@@
 
-print(resultsK_red)
+print(resultsK2)
 
@@
 # Create the recipe for preprocessing
-recipeMg_red <- recipe(Mg ~ kerosene + fueltype, data = hapin_merged) %>%
+recipeMg2 <- recipe(Mg ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
   step_dummy(all_nominal(), -all_outcomes())
 
-# Create a workflow
-workflowMg_red <- workflow() %>%
-  add_recipe(recipeMg_red) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitMg_red <- workflowMg_red %>%
+#Fit the model
+fitMg2 <- workflow() %>%
+  add_recipe(recipeMg2) %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 # Extract and print the results of the model
-resultsMg_red <- fitMg_red %>%
+resultsMg2 <- fitMg2 %>%
   pull_workflow_fit() %>%
@@
 
-print(resultsMg_red)
+print(resultsMg2)
 
@@
 # Create the recipe for preprocessing
-recipeMn_red <- recipe(Mn ~ fueltype, data = hapin_merged) %>%
+recipeMn2 <- recipe(Mn ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
   step_dummy(all_nominal(), -all_outcomes())
 
-# Create a workflow
-workflowMn_red <- workflow() %>%
-  add_recipe(recipeMn_red) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitMn_red <- workflowMn_red %>%
+#Fit the model
+fitMn2 <- workflow() %>%
+  add_recipe(recipeMn2) %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 # Extract and print the results of the model
-resultsMn_red <- fitMn_red %>%
+resultsMn2 <- fitMn2 %>%
   pull_workflow_fit() %>%
@@
 
-print(resultsMn_red)
+print(resultsMn2)
 
@@
 # Create the recipe for preprocessing
-recipeS_red <- recipe(S ~ fueltype, data = hapin_merged) %>%
+recipeSul2 <- recipe(S ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
   step_dummy(all_nominal(), -all_outcomes())
 
-# Create a workflow
-workflowS_red <- workflow() %>%
-  add_recipe(recipeS_red) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitS_red <- workflowS_red %>%
+#Fit the model
+fitSul2 <- workflow() %>%
+  add_recipe(recipeSul2) %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 # Extract and print the results of the model
-resultsS_red <- fitS_red %>%
+resultsSul2 <- fitSul2 %>%
   pull_workflow_fit() %>%
@@
 
-print(resultsS_red)
+print(resultsSul2)
 
-###########To predict the Fe variable##############
+###########To predict the Si variable##############
 # Create the recipe for preprocessing
-recipeFe_red <- recipe(Fe ~ stove, data = hapin_merged) %>%
+recipeSi2 <- recipe(Si ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
   step_dummy(all_nominal(), -all_outcomes())
 
-# Create a workflow
-workflowFe_red <- workflow() %>%
-  add_recipe(recipeFe_red) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitFe_red <- workflowFe_red %>%
+#Fit the model
+fitSi2 <- workflow() %>%
+  add_recipe(recipeSi2) %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 # Extract and print the results of the model
-resultsFe_red <- fitFe_red %>%
+resultsSi2 <- fitSi2 %>%
   pull_workflow_fit() %>%
@@
 
-print(resultsFe_red)
+print(resultsSi2)
 
-###########To predict the Si variable##############
+###########To predict the Ti variable##############
 # Create the recipe for preprocessing
-recipeSi_red <- recipe(Si ~ fueltype, data = hapin_merged) %>%
+recipeTi2 <- recipe(Ti ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
   step_dummy(all_nominal(), -all_outcomes())
 
-# Create a workflow
-workflowSi_red <- workflow() %>%
-  add_recipe(recipeSi_red) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitSi_red <- workflowSi_red %>%
+#Fit the model
+fitTi2 <- workflow() %>%
+  add_recipe(recipeTi2) %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 # Extract and print the results of the model
-resultsSi_red <- fitSi_red %>%
+resultsTi2 <- fitTi2 %>%
   pull_workflow_fit() %>%
@@
 
-print(resultsSi_red)
+print(resultsTi2)
 
-###########To predict the Ti variable##############
+###########To predict the Zn variable##############
 # Create the recipe for preprocessing
-recipeTi_red <- recipe(Ti ~ fueltype, data = hapin_merged) %>%
+recipeZn2 <- recipe(Zn ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
   step_dummy(all_nominal(), -all_outcomes())
 
-# Create a workflow
-workflowTi_red <- workflow() %>%
-  add_recipe(recipeTi_red) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitTi_red <- workflowTi_red %>%
+#Fit the model
+fitZn2 <- workflow() %>%
+  add_recipe(recipeZn2) %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 # Extract and print the results of the model
-resultsTi_red <- fitTi_red %>%
+resultsZn2 <- fitZn2 %>%
   pull_workflow_fit() %>%
@@
 
-print(resultsTi_red)
+print(resultsZn2)
 ```
 
-In this part I will evaluate both models from each chemical species (TBD)
+Using this approach it seems like only `fueltype` and `other_stove` are almost all of the variables important to make predictions of the chemical species.
+
+And now evaluating both models for all the species. First I will set the resampling method using cross-validation.
 
 ```{r}
+#Set the seed for reproducibility
+set.seed(1234)
 
+# Set up resampling method with 10 folds cross-validation
+samples <- vfold_cv(hapin_merged, v = 10)
 ```
+
+And now evaluating both models of `Al` using the RMSE and R-squared values for both models. The same process will be described below for the other chemical species.
+
+```{r}
+###### MODEL 1 #######
+# Fit the model using cross-validation
+Al_evaluation <- workflow() %>%
+  add_recipe(recipeAl) %>%
+  add_model(gamma_spec) %>%
+  fit_resamples(resamples = samples)
+
+# Extract the results of the model
+Al_metrics <- Al_evaluation %>%
+  collect_metrics()
+
+###### MODEL 2 ######
+# Fit the model using cross-validation
+Al2_evaluation <- workflow() %>%
+  add_recipe(recipeAl2) %>%
+  add_model(gamma_spec) %>%
+  fit_resamples(resamples = samples)
+
+# Extract the results of the model
+Al2_metrics <- Al2_evaluation %>%
+  collect_metrics()
+
+#Print the RMSE and R-square for both models
+print(Al_metrics)
+print(Al2_metrics)
+```
+
+Now for BC...
+
+```{r}
+###### MODEL 1 #######
+# Fit the model using cross-validation
+BC_evaluation <- workflow() %>%
+  add_recipe(recipeBC) %>%
+  add_model(gamma_spec) %>%
+  fit_resamples(resamples = samples)
+
+# Extract the results of the model
+BC_metrics <- BC_evaluation %>%
+  collect_metrics()
+
+###### MODEL 2 ######
+# Fit the model using cross-validation
+BC2_evaluation <- workflow() %>%
+  add_recipe(recipeBC2) %>%
+  add_model(gamma_spec) %>%
+  fit_resamples(resamples = samples)
+
+# Extract the results of the model
+BC2_metrics <- BC2_evaluation %>%
+  collect_metrics()
+
+#Print the RMSE and R-square for both models
+print(BC_metrics)
+print(BC2_metrics)
+```
+
+Now for Ca...
+
+```{r}
+###### MODEL 1 #######
+# Fit the model using cross-validation
+Ca_evaluation <- workflow() %>%
+  add_recipe(recipeCa) %>%
+  add_model(gamma_spec) %>%
+  fit_resamples(resamples = samples)
+
+# Extract the results of the model
+Ca_metrics <- Ca_evaluation %>%
+  collect_metrics()
+
+###### MODEL 2 ######
+# Fit the model using cross-validation
+Ca2_evaluation <- workflow() %>%
+  add_recipe(recipeCa2) %>%
+  add_model(gamma_spec) %>%
+  fit_resamples(resamples = samples)
+
+# Extract the results of the model
+Ca2_metrics <- Ca2_evaluation %>%
+  collect_metrics()
+
+#Print the RMSE and R-square for both models
+print(Ca_metrics)
+print(Ca2_metrics)
+```
+
+Now for Fe...
+
+```{r}
+###### MODEL 1 #######
+# Fit the model using cross-validation
+Fe_evaluation <- workflow() %>%
+  add_recipe(recipeFe) %>%
+  add_model(gamma_spec) %>%
+  fit_resamples(resamples = samples)
+
+# Extract the results of the model
+Fe_metrics <- Fe_evaluation %>%
+  collect_metrics()
+
+###### MODEL 2 ######
+# Fit the model using cross-validation
+Fe2_evaluation <- workflow() %>%
+  add_recipe(recipeFe2) %>%
+  add_model(gamma_spec) %>%
+  fit_resamples(resamples = samples)
+
+# Extract the results of the model
+Fe2_metrics <- Fe2_evaluation %>%
+  collect_metrics()
+
+#Print the RMSE and R-square for both models
+print(Fe_metrics)
+print(Fe2_metrics)
+```
+
+Now for K...
+
+```{r}
+###### MODEL 1 #######
+# Fit the model using cross-validation
+K_evaluation <- workflow() %>%
+  add_recipe(recipeK) %>%
+  add_model(gamma_spec) %>%
+  fit_resamples(resamples = samples)
+
+# Extract the results of the model
+K_metrics <- K_evaluation %>%
+  collect_metrics()
+
+###### MODEL 2 ######
+# Fit the model using cross-validation
+K2_evaluation <- workflow() %>%
+  add_recipe(recipeK2) %>%
+  add_model(gamma_spec) %>%
+  fit_resamples(resamples = samples)
+
+# Extract the results of the model
+K2_metrics <- K2_evaluation %>%
+  collect_metrics()
+
+#Print the RMSE and R-square for both models
+print(K_metrics)
+print(K2_metrics)
+```
+
+Now for Mg...
+
+```{r}
+###### MODEL 1 #######
+# Fit the model using cross-validation
+Mg_evaluation <- workflow() %>%
+  add_recipe(recipeMg) %>%
+  add_model(gamma_spec) %>%
+  fit_resamples(resamples = samples)
+
+# Extract the results of the model
+Mg_metrics <- Mg_evaluation %>%
+  collect_metrics()
+
+###### MODEL 2 ######
+# Fit the model using cross-validation
+Mg2_evaluation <- workflow() %>%
+  add_recipe(recipeMg2) %>%
+  add_model(gamma_spec) %>%
+  fit_resamples(resamples = samples)
+
+# Extract the results of the model
+Mg2_metrics <- Mg2_evaluation %>%
+  collect_metrics()
+
+#Print the RMSE and R-square for both models
+print(Mg_metrics)
+print(Mg2_metrics)
+```
+
+Now for Mn...
+
+```{r}
+###### MODEL 1 #######
+# Fit the model using cross-validation
+Mn_evaluation <- workflow() %>%
+  add_recipe(recipeMn) %>%
+  add_model(gamma_spec) %>%
+  fit_resamples(resamples = samples)
+
+# Extract the results of the model
+Mn_metrics <- Mn_evaluation %>%
+  collect_metrics()
+
+###### MODEL 2 ######
+# Fit the model using cross-validation
+Mn2_evaluation <- workflow() %>%
+  add_recipe(recipeMn2) %>%
+  add_model(gamma_spec) %>%
+  fit_resamples(resamples = samples)
+
+# Extract the results of the model
+Mn2_metrics <- Mn2_evaluation %>%
+  collect_metrics()
+
+#Print the RMSE and R-square for both models
+print(Mn_metrics)
+print(Mn2_metrics)
+```
+
+Now for S...
+
+```{r}
+###### MODEL 1 #######
+# Fit the model using cross-validation
+Sul_evaluation <- workflow() %>%
+  add_recipe(recipeSul) %>%
+  add_model(gamma_spec) %>%
+  fit_resamples(resamples = samples)
+
+# Extract the results of the model
+Sul_metrics <- Sul_evaluation %>%
+  collect_metrics()
+
+###### MODEL 2 ######
+# Fit the model using cross-validation
+Sul2_evaluation <- workflow() %>%
+  add_recipe(recipeSul2) %>%
+  add_model(gamma_spec) %>%
+  fit_resamples(resamples = samples)
+
+# Extract the results of the model
+Sul2_metrics <- Sul2_evaluation %>%
+  collect_metrics()
+
+#Print the RMSE and R-square for both models
+print(Sul_metrics)
+print(Sul2_metrics)
+```
+
+Now for Si...
+
+```{r}
+###### MODEL 1 #######
+# Fit the model using cross-validation
+Si_evaluation <- workflow() %>%
+  add_recipe(recipeSi) %>%
+  add_model(gamma_spec) %>%
+  fit_resamples(resamples = samples)
+
+# Extract the results of the model
+Si_metrics <- Si_evaluation %>%
+  collect_metrics()
+
+###### MODEL 2 ######
+# Fit the model using cross-validation
+Si2_evaluation <- workflow() %>%
+  add_recipe(recipeSi2) %>%
+  add_model(gamma_spec) %>%
+  fit_resamples(resamples = samples)
+
+# Extract the results of the model
+Si2_metrics <- Si2_evaluation %>%
+  collect_metrics()
+
+#Print the RMSE and R-square for both models
+print(Si_metrics)
+print(Si2_metrics)
+```
+
+Now for Ti...
+
+```{r}
+###### MODEL 1 #######
+# Fit the model using cross-validation
+Ti_evaluation <- workflow() %>%
+  add_recipe(recipeTi) %>%
+  add_model(gamma_spec) %>%
+  fit_resamples(resamples = samples)
+
+# Extract the results of the model
+Ti_metrics <- Ti_evaluation %>%
+  collect_metrics()
+
+###### MODEL 2 ######
+# Fit the model using cross-validation
+Ti2_evaluation <- workflow() %>%
+  add_recipe(recipeTi2) %>%
+  add_model(gamma_spec) %>%
+  fit_resamples(resamples = samples)
+
+# Extract the results of the model
+Ti2_metrics <- Ti2_evaluation %>%
+  collect_metrics()
+
+#Print the RMSE and R-square for both models
+print(Ti_metrics)
+print(Ti2_metrics)
+```
+
+Now for Zn...
+
+```{r}
+###### MODEL 1 #######
+# Fit the model using cross-validation
+Zn_evaluation <- workflow() %>%
+  add_recipe(recipeZn) %>%
+  add_model(gamma_spec) %>%
+  fit_resamples(resamples = samples)
+
+# Extract the results of the model
+Zn_metrics <- Zn_evaluation %>%
+  collect_metrics()
+
+###### MODEL 2 ######
+# Fit the model using cross-validation
+Zn2_evaluation <- workflow() %>%
+  add_recipe(recipeZn2) %>%
+  add_model(gamma_spec) %>%
+  fit_resamples(resamples = samples)
+
+# Extract the results of the model
+Zn2_metrics <- Zn2_evaluation %>%
+  collect_metrics()
+
+#Print the RMSE and R-square for both models
+print(Zn_metrics)
+print(Zn2_metrics)
+```
+
+For all of the models it seems like the model with all the predictors works best, instead of grouping the predcitors (exposures) in further categories.
+However, the R-square and RMSE values are really low, so maybe a different approach for modeling is required.
+
+I will do an attempt to save all the RMSE an R-squared values in one table:
+
+```{r}
+#Bind dfs that contain all RMSEs and R-squares from all models
+metrics_final <- bind_rows(Al_metrics, Al2_metrics, BC_metrics, BC2_metrics, Ca_metrics, Ca2_metrics, Fe_metrics, Fe2_metrics,
+                           K_metrics, K2_metrics, Mg_metrics, Mg2_metrics, Mn_metrics, Mn2_metrics, Si_metrics, Si2_metrics,
+                           Sul_metrics, Sul2_metrics, Ti_metrics, Ti2_metrics, Zn_metrics, Zn2_metrics)
+
+#Create a vector to name all chemical species
+species <- rep(c("Al", "BC", "Ca", "Fe", "K", "Mg", "Mn", "Si", "S", "Ti", "Zn"), each = 4)
+
+# Adding the 'chemical_species' variable to 'metrics_final'
+metrics_final$Species <- species
+
+# Creating a vector of observation types
+obs <- rep(c(rep("N1", 2), rep("N2", 2)), length.out = 44)
+
+# Adding the 'obs' variable to 'metrics_final'
+metrics_final$obs <- obs
+
+#Finally, delete non-necessary columns
+metrics_final <- metrics_final %>% select(-c(.estimator, n, std_err, .config))
+```
+
+Do additional adjustments to the metrics table, and then save it as an .RDS file.
+
+```{r}
+#Change dataframe from long to wide
+metrics_wide <- metrics_final %>% pivot_wider(names_from = obs, values_from = mean)
+
+#Rename columns, rearrange them and edit values for some columns
+metrics_wide <- metrics_wide %>% rename(Metric = ".metric", Model1 = "N1", Model2 = "N2") %>%
+  select(c(Species, Metric, Model1, Model2))
+
+#Save df as .RDS file
+saveRDS(metrics_wide, file = here("results", "tables", "model_metrics.rds"))
+```

### 2024-04-05 code/data_exploring_code.qmd
Project Part 4 complete
Done

@@
 
+Given that some of the categories have very low responses, I decided to combine some categories to have a greater variability in the exposures and then explore if these are good when modelling.
+
+```{r}
+#Combine some exposures to new categories
+hapin <- hapin %>%
+  mutate(micro_exposures = if_else( #The micro_exposures variable will include 'smoke', 'coil', 'incense' and 'kerosene'
+    (smoke == "Yes" | coil == "Yes" | incense == "Yes" | kerosene == "Yes"),
+    "Yes", "No", missing = "No"
+  )) %>%
+  mutate(macro_exposures = if_else( #The macro_exposures variable will include 'trash' and 'crop'
+    (trash == "Yes" | crop == "Yes"),
+    "Yes", "No", missing = "No"
+  ))
+
+#Save this as a new .RDS file
+saveRDS(hapin, file = here("data", "processed-data", "hapin_samples.rds"))
+```
+
 And finally plotting the PM2.5, BC and CO concentrations in different graphs using `ggplot()`

### 2024-04-05 products/manuscript/Manuscript.qmd
Project Part 4 complete
Done

@@
 title: "Elemental characterization of personal filter samples from the Household Air Pollution Intervention Network (HAPIN) trial for Source Apportionment"
-subtitle: "Project: Part 3"
+subtitle: "Project: Part 4"
 author: Erick E. Mollinedo
@@
 
-The correlation (r^2^) between chemical species will be determined using either individual paired correlations or by a correlation plot matrix.
-Additional plots, such as boxplots, to visualize and compare how the concentrations differ by type of fuel.
-These differences will be assessed for statistical significance using a Welch two-sample T-test.
-Multivariate regression models will be computed to analyze which exposures could explain the concentrations of the key chemical species.
-Given that the concentrations can be approximated to a gamma distribution, the generalized linear model with gamma distribution will be used to compute the models.
-The root mean square error (RMSE) and R-squared will be the parameters to evaluate the most fit model for each of the chemical species.
+The correlation (r^2^) between chemical species were determined using paired correlations in a correlation plot matrix.
+To assess differences in concentrations between samples by type of fuel, a Welch two-sample T-test was computed.
+Additional plots were presented to visualize those differences between samples.
 
-## Preliminary Results
+Multivariate regression models were computed for each one of the key chemical species to assess which exposures are likely to explain the variations in their concentrations.
+Given that there is low variability in the answers for some of the categorical exposures, in addition to computing a multivariate model with all the predictors, another model was evaluated.
+This second model considers other stoves and fueltype as predictors, alongside a new combined variable called 'micro exposures' that englobes exposures to kerosene, cigarette, coils and incense and the variable 'macro exposures' which englobes exposures to trash burning and crop residues.
+Given that the concentrations can be approximated to a gamma distribution, the generalized linear model with gamma distribution was used to compute the models.
+The root mean square error (RMSE) and R-squared were the metric parameters to evaluate which one of both models for all chemical species fits best.
+
+## Results
 
@@
 Other noticeable patterns is that some of the species were found in higher concentrations for personal samples from participants that used biomass stoves, compared to LPG stoves (@fig-concentrations).
-When conducting the two-sided T-test, it was determined that all of the chemical species, except Fe, Al and Zn were statistically significantly different between type of fuel.
+When conducting the two-sided T-test, it was determined that all of the chemical species, except Fe, Al and Zn were significantly different between type of fuel.
 The differences in concentrations were also analyzed by study arm, however, given that all baseline measurements belonged to biomass fueltype observations, these results were not presented but can be observed in the Supplementary material.
@@
 
-### Chemical species prediction
+### Chemical species models
 
-Nine of the eleven exposures were evaluated as predictors of the chemical species concentrations.
-The use of a generator machine (for corn mill grinding) and incence were left out since these were the categorical exposure variables with the less variability in answers.
-The gamma distributed generalized linear models were chosen given the concentrations of all the species do not follow a normal distribution.
+Model 1, for each one of the chemical species, was computed to determine which of the exposures could better explain the variations in the concentrations (Supplementary Material).
+Given that none of the observations recorded to have used a generator machine for corn mill grinding, this categorical variable was left out.
+When computing this model, none of the predictors were significant to predict the concentrations of Aluminim (Al), Iron (Fe) and Zinc (Zn).
+Kerosene was found a significant predictor of Black carbon (BC) and Magnesium (Mg) concentrations (p < 0.05).
+Exposure to smoke emitted from a biomass stove other than the one in their homes was also found as a significant predictor (p < 0.05) for BC and potassium (K).
+Also, exposure to cigarette smoke in the homes was associated to K concentration.
+Meanwhile, the strongest predictor was found to be type of fuel (p < 0.01), where exposure to biomass fuel was associated to an increase of the concentration of BC, Ca, K, Mg, Mn, S, Si and Ti.
+These last results are consistent to what was observed in terms of significant differences among samples by type of fuel (@fig-concentrations).
 
-Based on the models, type of fuel was a strong predictor of concentration for eight chemical species: Mg, Mn, BC, Ca, K, S, Si and Ti.
-Kerosene was another predictor of BC and Mg concentrations alongside type of fuel and being exposed to other sources of smoke (such as traffic emissions).
-Meanwhile, smoking and other sources of smoke were high predictors of K.
-On the other hand, stove usage, independently if it is biomass or LPG, was associated with concentration of Fe.
-However, additional analyses have to be conducted for final model selection.
+Model 2 was tested as another way to evaluate which predictors could explain the variations in concentrations by grouping certain exposures.
+Similar results were observed as from Model 1.
+Micro exposures, which is a category that includes smoking, incense use, kerosene and mosquito coils explained the variations in BC concentrations (p < 0.05).
+Exposure to other biomass stoves was a predictor for BC and K (p < 0.05).
+Fueltype was still the strongest predictor (p < 0.01) for the variations in concentrations for BC, Ca, K, Mg, Mn, S, Si and Ti.
+
+In terms of comparison between models, both models seem to under-perform in terms of predictions (Table 1, Supplementary Material).
+The lowest RMSE was observed in Model 2 for all the chemical species, meanwhile the highest R-square was observed in Model 1.
+In this case, Model 2 with grouping of variables could be a better model to explain the effect of certain exposures to the concentrations of the selected chemical species.
+
+## Discussion
+
+## Conclusions
 

### 2024-04-05 products/manuscript/supplement/Supplementary.qmd
Project Part 4 complete
Done

@@
 title: "Supplementary Material"
-subtitle: "Project: Part 3"
+subtitle: "Project: Part 4"
 author: Erick E. Mollinedo
@@
 library(readr)
+library(gt)
 ```
@@
 
-hapin_merged <- read_csv(here("data", "processed-data", "hapin-merged.csv"))
+hapin_merged <- read_rds(here("data", "processed-data", "hapin-final.rds"))
 
 # Create a generalized linear regression model specification using gamma distribution (This will be used for all the models)
-model_spec <- linear_reg(mode = "regression") %>%
+gamma_spec <- linear_reg(mode = "regression") %>%
   set_engine("glm", family = Gamma(link = "log"))
+
+###########To predict the Al variable##############
+# Create the recipe for preprocessing
+recipeAl <- recipe(Al ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
+  step_dummy(all_nominal(), -all_outcomes())
+
+#Fit the model
+fitAl <- workflow() %>%
+  add_recipe(recipeAl) %>%
+  add_model(gamma_spec) %>%
+  fit(data = hapin_merged)
+
+# Extract and print the results of the model
+resultsAl <- fitAl %>%
+  pull_workflow_fit() %>%
+  tidy()
+
+print(resultsAl)
 
@@
 # Create the recipe for preprocessing
-recipeBC_red <- recipe(BC ~ kerosene + stove_other + fueltype, data = hapin_merged) %>%
+recipeBC <- recipe(BC ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
   step_dummy(all_nominal(), -all_outcomes())
 
-# Create a workflow
-workflowBC_red <- workflow() %>%
-  add_recipe(recipeBC_red) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitBC_red <- workflowBC_red %>%
+#Fit the model
+fitBC <- workflow() %>%
+  add_recipe(recipeBC) %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 # Extract and print the results of the model
-resultsBC_red <- fitBC_red %>%
+resultsBC <- fitBC %>%
   pull_workflow_fit() %>%
@@
 
-print(resultsBC_red)
+print(resultsBC)
 
@@
 # Create the recipe for preprocessing
-recipeCa_red <- recipe(Ca ~ fueltype, data = hapin_merged) %>%
+recipeCa <- recipe(Ca ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
   step_dummy(all_nominal(), -all_outcomes())
 
-# Create a workflow
-workflowCa_red <- workflow() %>%
-  add_recipe(recipeCa_red) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitCa_red <- workflowCa_red %>%
+#Fit the model
+fitCa <- workflow() %>%
+  add_recipe(recipeCa) %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 # Extract and print the results of the model
-resultsCa_red <- fitCa_red %>%
+resultsCa <- fitCa %>%
   pull_workflow_fit() %>%
@@
 
-print(resultsCa_red)
+print(resultsCa)
+
+###########To predict the Fe variable##############
+# Create the recipe for preprocessing
+recipeFe <- recipe(Fe ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
+  step_dummy(all_nominal(), -all_outcomes())
+
+#Fit the model
+fitFe <- workflow() %>%
+  add_recipe(recipeFe) %>%
+  add_model(gamma_spec) %>%
+  fit(data = hapin_merged)
+
+# Extract and print the results of the model
+resultsFe <- fitFe %>%
+  pull_workflow_fit() %>%
+  tidy()
+
+print(resultsFe)
 
@@
 # Create the recipe for preprocessing
-recipeK_red <- recipe(K ~ smoke + stove_other + fueltype, data = hapin_merged) %>%
+recipeK <- recipe(K ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
   step_dummy(all_nominal(), -all_outcomes())
 
-# Create a workflow
-workflowK_red <- workflow() %>%
-  add_recipe(recipeK_red) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitK_red <- workflowK_red %>%
+#Fit the model
+fitK <- workflow() %>%
+  add_recipe(recipeK) %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 # Extract and print the results of the model
-resultsK_red <- fitK_red %>%
+resultsK <- fitK %>%
   pull_workflow_fit() %>%
@@
 
-print(resultsK_red)
+print(resultsK)
 
@@
 # Create the recipe for preprocessing
-recipeMg_red <- recipe(Mg ~ kerosene + fueltype, data = hapin_merged) %>%
+recipeMg <- recipe(Mg ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
   step_dummy(all_nominal(), -all_outcomes())
 
-# Create a workflow
-workflowMg_red <- workflow() %>%
-  add_recipe(recipeMg_red) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitMg_red <- workflowMg_red %>%
+#Fit the model
+fitMg <- workflow() %>%
+  add_recipe(recipeMg) %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 # Extract and print the results of the model
-resultsMg_red <- fitMg_red %>%
+resultsMg <- fitMg %>%
   pull_workflow_fit() %>%
@@
 
-print(resultsMg_red)
+print(resultsMg)
 
@@
 # Create the recipe for preprocessing
-recipeMn_red <- recipe(Mn ~ fueltype, data = hapin_merged) %>%
+recipeMn <- recipe(Mn ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
   step_dummy(all_nominal(), -all_outcomes())
 
-# Create a workflow
-workflowMn_red <- workflow() %>%
-  add_recipe(recipeMn_red) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitMn_red <- workflowMn_red %>%
+#Fit the model
+fitMn <- workflow() %>%
+  add_recipe(recipeMn) %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 # Extract and print the results of the model
-resultsMn_red <- fitMn_red %>%
+resultsMn <- fitMn %>%
   pull_workflow_fit() %>%
@@
 
-print(resultsMn_red)
+print(resultsMn)
 
@@
 # Create the recipe for preprocessing
-recipeS_red <- recipe(S ~ fueltype, data = hapin_merged) %>%
+recipeSul <- recipe(S ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
   step_dummy(all_nominal(), -all_outcomes())
 
-# Create a workflow
-workflowS_red <- workflow() %>%
-  add_recipe(recipeS_red) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitS_red <- workflowS_red %>%
+#Fit the model
+fitSul <- workflow() %>%
+  add_recipe(recipeSul) %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 # Extract and print the results of the model
-resultsS_red <- fitS_red %>%
+resultsSul <- fitSul %>%
   pull_workflow_fit() %>%
@@
 
-print(resultsS_red)
+print(resultsSul)
 
-###########To predict the Fe variable##############
+###########To predict the Si variable##############
 # Create the recipe for preprocessing
-recipeFe_red <- recipe(Fe ~ stove, data = hapin_merged) %>%
+recipeSi <- recipe(Si ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
   step_dummy(all_nominal(), -all_outcomes())
 
-# Create a workflow
-workflowFe_red <- workflow() %>%
-  add_recipe(recipeFe_red) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitFe_red <- workflowFe_red %>%
+#Fit the model
+fitSi <- workflow() %>%
+  add_recipe(recipeSi) %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 # Extract and print the results of the model
-resultsFe_red <- fitFe_red %>%
+resultsSi <- fitSi %>%
   pull_workflow_fit() %>%
@@
 
-print(resultsFe_red)
+print(resultsSi)
 
-###########To predict the Si variable##############
+###########To predict the Ti variable##############
 # Create the recipe for preprocessing
-recipeSi_red <- recipe(Si ~ fueltype, data = hapin_merged) %>%
+recipeTi <- recipe(Ti ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
   step_dummy(all_nominal(), -all_outcomes())
 
-# Create a workflow
-workflowSi_red <- workflow() %>%
-  add_recipe(recipeSi_red) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitSi_red <- workflowSi_red %>%
+#Fit the model
+fitTi <- workflow() %>%
+  add_recipe(recipeTi) %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 # Extract and print the results of the model
-resultsSi_red <- fitSi_red %>%
+resultsTi <- fitTi %>%
   pull_workflow_fit() %>%
@@
 
-print(resultsSi_red)
+print(resultsTi)
 
-###########To predict the Ti variable##############
+###########To predict the Zn variable##############
 # Create the recipe for preprocessing
-recipeTi_red <- recipe(Ti ~ fueltype, data = hapin_merged) %>%
+recipeZn <- recipe(Zn ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
   step_dummy(all_nominal(), -all_outcomes())
 
-# Create a workflow
-workflowTi_red <- workflow() %>%
-  add_recipe(recipeTi_red) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitTi_red <- workflowTi_red %>%
+#Fit the model
+fitZn <- workflow() %>%
+  add_recipe(recipeZn) %>%
+  add_model(gamma_spec) %>%
   fit(data = hapin_merged)
@@
 # Extract and print the results of the model
-resultsTi_red <- fitTi_red %>%
+resultsZn <- fitZn %>%
   pull_workflow_fit() %>%
@@
 
-print(resultsTi_red)
+print(resultsZn)
 ```
+
+The following table shows the values used to judge models (RMSE and R-square) for all chemical species between Model 1 and Model 2.
+
+```{r echo=FALSE}
+#Load the metrics table
+suptable <- read_rds(here("results", "tables", "model_metrics.rds"))
+
+#Show the table
+suptable %>% gt()
+```

### 2024-04-19 code/data_analysis_code.qmd
Project Part 5
Completed part 5 of the project (might review and do a few edits later)

@@
 ---
-title: "Project Part 3"
-subtitle: "Data analysis code"
+title: "Data analysis code"
 author: Erick E. Mollinedo
@@
 library(corrplot)
+library(grid)
 library(RColorBrewer)
@@
 
-To answer question 1, I will plot the correlations in a correlation plot using the `corrplot()` function.
+### Summarize Concentrations
+
+The following chunk of code is to create the descriptive statistics tables for all chemical species concentrations.
+The tables will be saves as RDS files, so later they can be loaded as tables using `gt()`.
+
+```{r}
+#Create data frames for both LPG and Biomass groups
+biomass <- concentration %>% filter(fueltype == "Biomass") %>% select(-c(arm, filter_id))
+lpg <- concentration %>% filter(fueltype == "LPG") %>% select(-c(arm, filter_id))
+
+#Format dfs longer
+biomass_long <- biomass %>% #Biomass dataframe
+  pivot_longer(cols = -fueltype,
+               names_to = "Species", values_to = "Concentration")
+
+lpg_long <- lpg %>% #LPG dataframe
+  pivot_longer(cols = -fueltype,
+               names_to = "Species", values_to = "Concentration")
+
+#Create summary tables for each group
+biomass_summary <- biomass_long %>% #Biomass fuel
+  group_by(Species) %>% 
+  summarise(N = n(),
+            Mean = round(mean(Concentration), 2),
+            SD = round(sd(Concentration), 2),
+            Median = round(median(Concentration), 2),
+            IQR = round(IQR(Concentration), 2),
+            Min = round(min(Concentration), 2),
+            Max = round(max(Concentration), 2))
+
+lpg_summary <- lpg_long %>% #LPG fuel
+  group_by(Species) %>% 
+  summarise(N = n(),
+            Mean = round(mean(Concentration), 2),
+            SD = round(sd(Concentration), 2),
+            Median = round(median(Concentration), 2),
+            IQR = round(IQR(Concentration), 2),
+            Min = round(min(Concentration), 2),
+            Max = round(max(Concentration), 2))
+
+#Save dataframes in RDS format so they can later be uploaded as tables
+write_rds(biomass_summary, here("results", "tables", "biomass-summary.rds"))
+write_rds(lpg_summary, here("results", "tables", "lpg-summary.rds"))
+```
+
+### Correlations
+
+To answer the first question (correlations by species), I will plot the correlations in a correlation plot using the `corrplot()` function.
 I used the previous code, since it was already developed, but editing to add more detail to the plot.
@@
 
-To answer question 2, I will conduct a T-test using the `tidymodels` package
+### Fuel type contrasts
+
+To answer question 2 (species differences among type of fuel), I will conduct a T-test using the `tidymodels` package
 
@@
     conf_low = t_test$conf.int[1],
-    conf_high = t_test$conf.int[2]
-  )
+    conf_high = t_test$conf.int[2])
 })
@@
 
-To answer question 3, I will conduct a regression analysis.
+To see a graphical representation of this test, check the boxplots generated in the `data_exploring_code.qmd` file.
+
+### PM2.5 and BC Modeling
+
+To answer questions 3 and 4 (modeling between PM2.5, BC and other chemical species), I will conduct a regression analysis.
 But first I will merge the data from `hapin_filters` with the `concentration` data.
@@
 
-#Remove pm25, bc and co, and additionally drop all NAs
+#Remove bc and co, and additionally drop all NAs
 hapin_merged <- hapin_merged %>%
-  select(-c(pm25, bc, co)) %>%
+  select(-c(bc, co)) %>%
   drop_na()
@@
 Now conducting the model selection using the `tidymodels` package.
-For the models, I used the gamma distribution family, since the concentrations of each chemical species could be explained by the gamma distribution.
-I used all the recorded exposures except incence and generator, since these had low variability in the answers.
+For the models, I tried two multivariate models, first, using the gamma distribution family, since the concentrations of each chemical species could be explained by the gamma distribution.
+And the second model, using a simple linear regression.
+Also, the data was split into training (80%) and test data (20%).
 
 ```{r}
-# Create a generalized linear regression model specification using gamma distribution (This will be used for all the models)
-gamma_spec <- linear_reg(mode = "regression") %>%
+# Split the data into training and testing sets
+rngseed = 1234
+set.seed(rngseed) # for reproducibility
+data_split <- initial_split(hapin_merged, prop = 0.8)
+train_data <- training(data_split)
+test_data <- testing(data_split)
+```
+
+Now, creating the formulas for each question. Also, computing the model specifications, the workflow and fitting the models
+
+```{r}
+#Create formulas
+pm25formula <- pm25 ~ Mg + Al + Si + S + K + Ca + Ti + Mn + Fe + Zn + BC
+bcformula <- BC ~ smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype
+
+#Set model specifications
+lin_mod <- linear_reg() %>% set_engine("lm") #Linear
+glm_mod <- linear_reg(mode = "regression") %>% #GLM gamma
   set_engine("glm", family = Gamma(link = "log"))
 
-###########To predict the Al variable##############
-# Create the recipe for preprocessing
-recipeAl <- recipe(Al ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
+## ----Model 1: PM2.5 Linear---- ##
+linpm_wflow <- workflow() %>% #Workflow
+	add_model(lin_mod) %>% 
+	add_formula(pm25formula)
 
-#Fit the model
-fitAl <- workflow() %>%
-  add_recipe(recipeAl) %>%
-  add_model(gamma_spec) %>%
-  fit(data = hapin_merged)
+linpm_fit <- linpm_wflow %>% fit(data = train_data) #Fit the model
 
-# Extract and print the results of the model
-resultsAl <- fitAl %>%
-  pull_workflow_fit() %>%
+## ----Model 2: PM2.5 GLM Gamma---- ##
+glmpm_wflow <- workflow() %>% #Workflow
+  add_model(glm_mod) %>% 
+  add_formula(pm25formula)
+
+glmpm_fit <- glmpm_wflow %>% fit(data = train_data) #Fit the model
+
+## ----Model 3: BC Linear---- ##
+linbc_wflow <- workflow() %>% #Workflow
+	add_model(lin_mod) %>% 
+	add_formula(bcformula)
+
+linbc_fit <- linbc_wflow %>% fit(data = train_data) #Fit the model
+
+## ----Model 4: BC GLM Gamma---- ##
+glmbc_wflow <- workflow() %>% #Workflow
+  add_model(glm_mod) %>% 
+  add_formula(bcformula)
+
+glmbc_fit <- glmbc_wflow %>% fit(data = train_data) #Fit the model
+```
+
+Extracting the results from each model.
+
+```{r}
+# Extract and print the results of the models
+results_pmlin <- linpm_fit %>% #PM2.5 Linear
+  extract_fit_parsnip() %>%
   tidy()
 
-print(resultsAl)
-
-###########To predict the BC variable##############
-# Create the recipe for preprocessing
-recipeBC <- recipe(BC ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-#Fit the model
-fitBC <- workflow() %>%
-  add_recipe(recipeBC) %>%
-  add_model(gamma_spec) %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsBC <- fitBC %>%
-  pull_workflow_fit() %>%
+results_pmglm <- glmpm_fit %>% #PM2.5 GLM
+  extract_fit_parsnip() %>%
   tidy()
 
-print(resultsBC)
-
-###########To predict the Ca variable##############
-# Create the recipe for preprocessing
-recipeCa <- recipe(Ca ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-#Fit the model
-fitCa <- workflow() %>%
-  add_recipe(recipeCa) %>%
-  add_model(gamma_spec) %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsCa <- fitCa %>%
-  pull_workflow_fit() %>%
+results_bclin <- linbc_fit %>% #BC Linear
+  extract_fit_parsnip() %>%
   tidy()
 
-print(resultsCa)
-
-###########To predict the Fe variable##############
-# Create the recipe for preprocessing
-recipeFe <- recipe(Fe ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-#Fit the model
-fitFe <- workflow() %>%
-  add_recipe(recipeFe) %>%
-  add_model(gamma_spec) %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsFe <- fitFe %>%
-  pull_workflow_fit() %>%
+results_bcglm <- glmbc_fit %>% #BC GLM
+  extract_fit_parsnip() %>%
   tidy()
 
-print(resultsFe)
-
-###########To predict the K variable##############
-# Create the recipe for preprocessing
-recipeK <- recipe(K ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-#Fit the model
-fitK <- workflow() %>%
-  add_recipe(recipeK) %>%
-  add_model(gamma_spec) %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsK <- fitK %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsK)
-
-###########To predict the Mg variable##############
-# Create the recipe for preprocessing
-recipeMg <- recipe(Mg ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-#Fit the model
-fitMg <- workflow() %>%
-  add_recipe(recipeMg) %>%
-  add_model(gamma_spec) %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsMg <- fitMg %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsMg)
-
-###########To predict the Mn variable##############
-# Create the recipe for preprocessing
-recipeMn <- recipe(Mn ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-#Fit the model
-fitMn <- workflow() %>%
-  add_recipe(recipeMn) %>%
-  add_model(gamma_spec) %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsMn <- fitMn %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsMn)
-
-###########To predict the S variable##############
-# Create the recipe for preprocessing
-recipeSul <- recipe(S ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-#Fit the model
-fitSul <- workflow() %>%
-  add_recipe(recipeSul) %>%
-  add_model(gamma_spec) %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsSul <- fitSul %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsSul)
-
-###########To predict the Si variable##############
-# Create the recipe for preprocessing
-recipeSi <- recipe(Si ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-#Fit the model
-fitSi <- workflow() %>%
-  add_recipe(recipeSi) %>%
-  add_model(gamma_spec) %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsSi <- fitSi %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsSi)
-
-###########To predict the Ti variable##############
-# Create the recipe for preprocessing
-recipeTi <- recipe(Ti ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-#Fit the model
-fitTi <- workflow() %>%
-  add_recipe(recipeTi) %>%
-  add_model(gamma_spec) %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsTi <- fitTi %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsTi)
-
-###########To predict the Zn variable##############
-# Create the recipe for preprocessing
-recipeZn <- recipe(Zn ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-#Fit the model
-fitZn <- workflow() %>%
-  add_recipe(recipeZn) %>%
-  add_model(gamma_spec) %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsZn <- fitZn %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsZn)
+#Print the results
+print(results_pmlin)
+print(results_pmglm)
+print(results_bclin)
+print(results_bcglm)
 ```
 
-Based on the previous models, it seems like the most important variables are `fueltype` (as expected), `kerosene`, `stove_other` and `smoke` depending on the chemical species.
-I was worried of the low variability on responses for some of the variables so I am going to use only the variables `micro_exposures`, `macro_exposures`, `kerosene`, `stove_other` and `fueltype`.
+Now, doing an evaluation of each model First, by estimating the metrics (In this case I will use RMSE, R-squared and MAE). First computing the predictions to make comparisons with the observed values.
 
 ```{r}
-###########To predict the Al variable##############
-# Create the recipe for preprocessing
-recipeAl2 <- recipe(Al ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
+#Compute the predictions
+#PM2.5 models
+lin_pm_pred <- linpm_fit %>% predict(train_data)
+glm_pm_pred <- glmpm_fit %>% predict(train_data)
+#BC models
+lin_bc_pred <- linbc_fit %>% predict(train_data)
+glm_bc_pred <- glmbc_fit %>% predict(train_data)
 
-#Fit the model
-fitAl2 <- workflow() %>%
-  add_recipe(recipeAl2) %>%
-  add_model(gamma_spec) %>%
-  fit(data = hapin_merged)
+#Compute the metrics
+#PM2.5 models
+lin_pm_metrics <-  bind_cols(train_data, lin_pm_pred) %>% metrics(truth = pm25, estimate = .pred) 
+glm_pm_metrics <- bind_cols(train_data, glm_pm_pred) %>% metrics(truth = pm25, estimate = .pred)
+#BC Models
+lin_bc_metrics <-  bind_cols(train_data, lin_bc_pred) %>% metrics(truth = BC, estimate = .pred) 
+glm_bc_metrics <- bind_cols(train_data, glm_bc_pred) %>% metrics(truth = BC, estimate = .pred)
 
-# Extract and print the results of the model
-resultsAl2 <- fitAl2 %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsAl2)
-
-###########To predict the BC variable##############
-# Create the recipe for preprocessing
-recipeBC2 <- recipe(BC ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-#Fit the model
-fitBC2 <- workflow() %>%
-  add_recipe(recipeBC2) %>%
-  add_model(gamma_spec) %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsBC2 <- fitBC2 %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsBC2)
-
-###########To predict the Ca variable##############
-# Create the recipe for preprocessing
-recipeCa2 <- recipe(Ca ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-#Fit the model
-fitCa2 <- workflow() %>%
-  add_recipe(recipeCa2) %>%
-  add_model(gamma_spec) %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsCa2 <- fitCa2 %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsCa2)
-
-###########To predict the Fe variable##############
-# Create the recipe for preprocessing
-recipeFe2 <- recipe(Fe ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-#Fit the model
-fitFe2 <- workflow() %>%
-  add_recipe(recipeFe2) %>%
-  add_model(gamma_spec) %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsFe2 <- fitFe2 %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsFe2)
-
-###########To predict the K variable##############
-# Create the recipe for preprocessing
-recipeK2 <- recipe(K ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-#Fit the model
-fitK2 <- workflow() %>%
-  add_recipe(recipeK2) %>%
-  add_model(gamma_spec) %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsK2 <- fitK2 %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsK2)
-
-###########To predict the Mg variable##############
-# Create the recipe for preprocessing
-recipeMg2 <- recipe(Mg ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-#Fit the model
-fitMg2 <- workflow() %>%
-  add_recipe(recipeMg2) %>%
-  add_model(gamma_spec) %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsMg2 <- fitMg2 %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsMg2)
-
-###########To predict the Mn variable##############
-# Create the recipe for preprocessing
-recipeMn2 <- recipe(Mn ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-#Fit the model
-fitMn2 <- workflow() %>%
-  add_recipe(recipeMn2) %>%
-  add_model(gamma_spec) %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsMn2 <- fitMn2 %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsMn2)
-
-###########To predict the S variable##############
-# Create the recipe for preprocessing
-recipeSul2 <- recipe(S ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-#Fit the model
-fitSul2 <- workflow() %>%
-  add_recipe(recipeSul2) %>%
-  add_model(gamma_spec) %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsSul2 <- fitSul2 %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsSul2)
-
-###########To predict the Si variable##############
-# Create the recipe for preprocessing
-recipeSi2 <- recipe(Si ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-#Fit the model
-fitSi2 <- workflow() %>%
-  add_recipe(recipeSi2) %>%
-  add_model(gamma_spec) %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsSi2 <- fitSi2 %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsSi2)
-
-###########To predict the Ti variable##############
-# Create the recipe for preprocessing
-recipeTi2 <- recipe(Ti ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-#Fit the model
-fitTi2 <- workflow() %>%
-  add_recipe(recipeTi2) %>%
-  add_model(gamma_spec) %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsTi2 <- fitTi2 %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsTi2)
-
-###########To predict the Zn variable##############
-# Create the recipe for preprocessing
-recipeZn2 <- recipe(Zn ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-#Fit the model
-fitZn2 <- workflow() %>%
-  add_recipe(recipeZn2) %>%
-  add_model(gamma_spec) %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsZn2 <- fitZn2 %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsZn2)
+#Print the metrics
+print(lin_pm_metrics)
+print(glm_pm_metrics)
+print(lin_bc_metrics)
+print(glm_bc_metrics)
 ```
 
-Using this approach it seems like only `fueltype` and `other_stove` are almost all of the variables important to make predictions of the chemical species.
-
-And now evaluating both models for all the species. First I will set the resampling method using cross-validation.
+Continue the model evaluation, plotting the observed vs predicted values for each model.
 
 ```{r}
-#Set the seed for reproducibility
-set.seed(1234)
+#Create dataframes to compute the plots
+#PM2.5
+pred_pm1 <- data.frame(predicted = lin_pm_pred$.pred, model = "linear")
+pred_pm2 <- data.frame(predicted = glm_pm_pred$.pred, model = "GLM")
+#BC
+pred_bc1 <- data.frame(predicted = lin_bc_pred$.pred, model = "linear")
+pred_bc2 <- data.frame(predicted = glm_bc_pred$.pred, model = "GLM")
 
-# Set up resampling method with 10 folds cross-validation
-samples <- vfold_cv(hapin_merged, v = 10)
+#Merge data frames
+#PM2.5
+pmdata_plot <- bind_rows(pred_pm1, pred_pm2) %>% 
+  mutate(observed = rep(train_data$pm25, 2))
+#BC
+bcdata_plot <- bind_rows(pred_bc1, pred_bc2) %>% 
+  mutate(observed = rep(train_data$BC, 2))
+
+#Create pred-obs plot
+
+#Define colors
+my_colors <- c("GLM" = "#E1BE6A", "linear" = "#40B0A6")
+
+#PM2.5
+pmpreds <- ggplot(pmdata_plot) +
+  geom_point(aes(x = observed, y = predicted, color= model), size= 2) +
+  facet_wrap(~ model)+
+  scale_color_manual(values = my_colors)+
+  labs(x = "Observed", y = "Predicted") +
+  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "black") +
+  theme_bw()
+
+#BC
+bcpreds <- ggplot(bcdata_plot) +
+  geom_point(aes(x = observed, y = predicted, color= model), size= 2) +
+  facet_wrap(~ model)+
+  scale_color_manual(values = my_colors)+
+  labs(x = "Observed", y = "Predicted") +
+  theme_bw()
+
+#Visualize plots
+pmpreds
+bcpreds
+
+#Save plots
+ggsave(here("results", "figures", "obs-pred_pm25.png"), plot = pmpreds, width = 10, height = 6, dpi = 300)
+ggsave(here("results", "figures", "obs-pred_bc.png"), plot = bcpreds, width = 10, height = 6, dpi = 300)
 ```
 
-And now evaluating both models of `Al` using the RMSE and R-squared values for both models. The same process will be described below for the other chemical species.
+Also, computing the residuals for each model and plotting them.
 
 ```{r}
-###### MODEL 1 #######
-# Fit the model using cross-validation
-Al_evaluation <- workflow() %>%
-  add_recipe(recipeAl) %>%
-  add_model(gamma_spec) %>%
-  fit_resamples(resamples = samples)
+#Calculate residuals and create data frame
+#PM2.5
+pmdata_plot <- pmdata_plot %>% mutate(observation = 1:988, residuals = observed - predicted)
+#BC
+bcdata_plot <- bcdata_plot %>% mutate(observation = 1:988, residuals = observed - predicted)
 
-# Extract the results of the model
-Al_metrics <- Al_evaluation %>%
-  collect_metrics()
+# Plot residuals using ggplot
+#PM2.5
+pmresiduals <- ggplot(pmdata_plot) +
+  geom_point(aes(x = observation, y = residuals, color= model), size= 2) +
+  geom_abline(intercept = 0, slope = 0, linetype = "dashed", color = "black") +
+  scale_color_manual(values = my_colors)+
+  labs(x = "Data point",
+       y = "Residuals")+
+  theme_bw()
 
-###### MODEL 2 ######
-# Fit the model using cross-validation
-Al2_evaluation <- workflow() %>%
-  add_recipe(recipeAl2) %>%
-  add_model(gamma_spec) %>%
-  fit_resamples(resamples = samples)
+#BC
+bcresiduals <- ggplot(bcdata_plot) +
+  geom_point(aes(x = observation, y = residuals, color= model), size= 2) +
+  geom_abline(intercept = 0, slope = 0, linetype = "dashed", color = "black") +
+  scale_color_manual(values = my_colors)+
+  labs(x = "Data point",
+       y = "Residuals")+
+  theme_bw()
 
-# Extract the results of the model
-Al2_metrics <- Al2_evaluation %>%
-  collect_metrics()
+#Visualize plots
+pmresiduals
+bcresiduals
 
-#Print the RMSE and R-square for both models
-print(Al_metrics)
-print(Al2_metrics)
+#Save plots
+ggsave(here("results", "figures", "residuals-pm25.png"), plot = pmresiduals, width = 10, height = 6, dpi = 300)
+ggsave(here("results", "figures", "residuals-bc.png"), plot = bcresiduals, width = 10, height = 6, dpi = 300)
 ```
 
-Now for BC...
+And then computing the 90% confidence intervals for the predicted values for each model.
 
 ```{r}
-###### MODEL 1 #######
-# Fit the model using cross-validation
-BC_evaluation <- workflow() %>%
-  add_recipe(recipeBC) %>%
-  add_model(gamma_spec) %>%
-  fit_resamples(resamples = samples)
+## ----Model 1: PM2.5 Linear---- ##
+# Compute prediction intervals
+pred_int_linpm <- predict(linpm_fit, train_data, type = "conf_int", level = 0.90) %>%
+  mutate(observation = 1:494)
 
-# Extract the results of the model
-BC_metrics <- BC_evaluation %>%
-  collect_metrics()
-
-###### MODEL 2 ######
-# Fit the model using cross-validation
-BC2_evaluation <- workflow() %>%
-  add_recipe(recipeBC2) %>%
-  add_model(gamma_spec) %>%
-  fit_resamples(resamples = samples)
-
-# Extract the results of the model
-BC2_metrics <- BC2_evaluation %>%
-  collect_metrics()
-
-#Print the RMSE and R-square for both models
-print(BC_metrics)
-print(BC2_metrics)
+# Combine the lower and upper bounds with the observed values
+pred_int_linpm <- cbind(lin_pm_pred, pred_int_linpm, train_data$pm25)
+pred_int_linpm <- pred_int_linpm %>% rename(observed = "train_data$pm25")
 ```
 
-Now for Ca...
+Finally, plotting the predictions and the prediction intervals.
 
 ```{r}
-###### MODEL 1 #######
-# Fit the model using cross-validation
-Ca_evaluation <- workflow() %>%
-  add_recipe(recipeCa) %>%
-  add_model(gamma_spec) %>%
-  fit_resamples(resamples = samples)
-
-# Extract the results of the model
-Ca_metrics <- Ca_evaluation %>%
-  collect_metrics()
-
-###### MODEL 2 ######
-# Fit the model using cross-validation
-Ca2_evaluation <- workflow() %>%
-  add_recipe(recipeCa2) %>%
-  add_model(gamma_spec) %>%
-  fit_resamples(resamples = samples)
-
-# Extract the results of the model
-Ca2_metrics <- Ca2_evaluation %>%
-  collect_metrics()
-
-#Print the RMSE and R-square for both models
-print(Ca_metrics)
-print(Ca2_metrics)
+#Prediction interval plot for the linear model of PM2.5
+ggplot(pred_int_linpm) +
+	geom_errorbar(aes(x = observed, ymin = .pred_lower, ymax = .pred_upper), width = 25) +
+	geom_point(aes(x = observed, y = .pred, color = ".pred"),	shape = 5) +
+	labs(x = "Observed", y = "Predicted", title = "Predicted observation with Prediction Intervals") +
+	geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "black") +
+	scale_color_manual(values = my_colors) +
+	theme_minimal()
 ```
 
-Now for Fe...
+This plot looks crowded for the linear model, so I won't use it as part of the final assessment of this and the other models. However I will keep it for future references.
+
+Then, doing a final evaluation of the model using the test data. I decided to select the linear multivariate models for both models (PM2.5 and BC).
 
 ```{r}
-###### MODEL 1 #######
-# Fit the model using cross-validation
-Fe_evaluation <- workflow() %>%
-  add_recipe(recipeFe) %>%
-  add_model(gamma_spec) %>%
-  fit_resamples(resamples = samples)
+## ----PM2.5 Linear---- ##
+#Fit the model using the test data
+linpm_fit_test <- linpm_wflow %>% fit(data = test_data)
 
-# Extract the results of the model
-Fe_metrics <- Fe_evaluation %>%
-  collect_metrics()
+#Make predictions in the test data
+pm_preds2 <- linpm_fit_test %>% predict(test_data)
 
-###### MODEL 2 ######
-# Fit the model using cross-validation
-Fe2_evaluation <- workflow() %>%
-  add_recipe(recipeFe2) %>%
-  add_model(gamma_spec) %>%
-  fit_resamples(resamples = samples)
+#Compute the metrics
+linpm_metrics_test <- bind_cols(test_data, pm_preds2) %>% metrics(truth = pm25, estimate = .pred)
 
-# Extract the results of the model
-Fe2_metrics <- Fe2_evaluation %>%
-  collect_metrics()
+print(linpm_metrics_test)
 
-#Print the RMSE and R-square for both models
-print(Fe_metrics)
-print(Fe2_metrics)
+## ----BC Linear---- ##
+#Fit the model using the test data
+linbc_fit_test <- linbc_wflow %>% fit(data = test_data)
+
+#Make predictions in the test data
+bc_preds2 <- linbc_fit_test %>% predict(test_data)
+
+#Compute the metrics
+linbc_metrics_test <- bind_cols(test_data, bc_preds2) %>% metrics(truth = BC, estimate = .pred)
+
+print(linbc_metrics_test)
 ```
 
-Now for K...
+### Save outputs as tables
+
+Creating a table to save all the metric values in one table. First creating new dataframes that includes all metrics for the PM2.5 and BC models.
 
 ```{r}
-###### MODEL 1 #######
-# Fit the model using cross-validation
-K_evaluation <- workflow() %>%
-  add_recipe(recipeK) %>%
-  add_model(gamma_spec) %>%
-  fit_resamples(resamples = samples)
+#Bind dfs that contain all RMSEs and R-squares from the PM2.5 and BC models
+metricspm_final <- bind_rows(lin_pm_metrics, glm_pm_metrics)
+metricsbc_final <- bind_rows(lin_bc_metrics, glm_bc_metrics)
 
-# Extract the results of the model
-K_metrics <- K_evaluation %>%
-  collect_metrics()
-
-###### MODEL 2 ######
-# Fit the model using cross-validation
-K2_evaluation <- workflow() %>%
-  add_recipe(recipeK2) %>%
-  add_model(gamma_spec) %>%
-  fit_resamples(resamples = samples)
-
-# Extract the results of the model
-K2_metrics <- K2_evaluation %>%
-  collect_metrics()
-
-#Print the RMSE and R-square for both models
-print(K_metrics)
-print(K2_metrics)
-```
-
-Now for Mg...
-
-```{r}
-###### MODEL 1 #######
-# Fit the model using cross-validation
-Mg_evaluation <- workflow() %>%
-  add_recipe(recipeMg) %>%
-  add_model(gamma_spec) %>%
-  fit_resamples(resamples = samples)
-
-# Extract the results of the model
-Mg_metrics <- Mg_evaluation %>%
-  collect_metrics()
-
-###### MODEL 2 ######
-# Fit the model using cross-validation
-Mg2_evaluation <- workflow() %>%
-  add_recipe(recipeMg2) %>%
-  add_model(gamma_spec) %>%
-  fit_resamples(resamples = samples)
-
-# Extract the results of the model
-Mg2_metrics <- Mg2_evaluation %>%
-  collect_metrics()
-
-#Print the RMSE and R-square for both models
-print(Mg_metrics)
-print(Mg2_metrics)
-```
-
-Now for Mn...
-
-```{r}
-###### MODEL 1 #######
-# Fit the model using cross-validation
-Mn_evaluation <- workflow() %>%
-  add_recipe(recipeMn) %>%
-  add_model(gamma_spec) %>%
-  fit_resamples(resamples = samples)
-
-# Extract the results of the model
-Mn_metrics <- Mn_evaluation %>%
-  collect_metrics()
-
-###### MODEL 2 ######
-# Fit the model using cross-validation
-Mn2_evaluation <- workflow() %>%
-  add_recipe(recipeMn2) %>%
-  add_model(gamma_spec) %>%
-  fit_resamples(resamples = samples)
-
-# Extract the results of the model
-Mn2_metrics <- Mn2_evaluation %>%
-  collect_metrics()
-
-#Print the RMSE and R-square for both models
-print(Mn_metrics)
-print(Mn2_metrics)
-```
-
-Now for S...
-
-```{r}
-###### MODEL 1 #######
-# Fit the model using cross-validation
-Sul_evaluation <- workflow() %>%
-  add_recipe(recipeSul) %>%
-  add_model(gamma_spec) %>%
-  fit_resamples(resamples = samples)
-
-# Extract the results of the model
-Sul_metrics <- Sul_evaluation %>%
-  collect_metrics()
-
-###### MODEL 2 ######
-# Fit the model using cross-validation
-Sul2_evaluation <- workflow() %>%
-  add_recipe(recipeSul2) %>%
-  add_model(gamma_spec) %>%
-  fit_resamples(resamples = samples)
-
-# Extract the results of the model
-Sul2_metrics <- Sul2_evaluation %>%
-  collect_metrics()
-
-#Print the RMSE and R-square for both models
-print(Sul_metrics)
-print(Sul2_metrics)
-```
-
-Now for Si...
-
-```{r}
-###### MODEL 1 #######
-# Fit the model using cross-validation
-Si_evaluation <- workflow() %>%
-  add_recipe(recipeSi) %>%
-  add_model(gamma_spec) %>%
-  fit_resamples(resamples = samples)
-
-# Extract the results of the model
-Si_metrics <- Si_evaluation %>%
-  collect_metrics()
-
-###### MODEL 2 ######
-# Fit the model using cross-validation
-Si2_evaluation <- workflow() %>%
-  add_recipe(recipeSi2) %>%
-  add_model(gamma_spec) %>%
-  fit_resamples(resamples = samples)
-
-# Extract the results of the model
-Si2_metrics <- Si2_evaluation %>%
-  collect_metrics()
-
-#Print the RMSE and R-square for both models
-print(Si_metrics)
-print(Si2_metrics)
-```
-
-Now for Ti...
-
-```{r}
-###### MODEL 1 #######
-# Fit the model using cross-validation
-Ti_evaluation <- workflow() %>%
-  add_recipe(recipeTi) %>%
-  add_model(gamma_spec) %>%
-  fit_resamples(resamples = samples)
-
-# Extract the results of the model
-Ti_metrics <- Ti_evaluation %>%
-  collect_metrics()
-
-###### MODEL 2 ######
-# Fit the model using cross-validation
-Ti2_evaluation <- workflow() %>%
-  add_recipe(recipeTi2) %>%
-  add_model(gamma_spec) %>%
-  fit_resamples(resamples = samples)
-
-# Extract the results of the model
-Ti2_metrics <- Ti2_evaluation %>%
-  collect_metrics()
-
-#Print the RMSE and R-square for both models
-print(Ti_metrics)
-print(Ti2_metrics)
-```
-
-Now for Zn...
-
-```{r}
-###### MODEL 1 #######
-# Fit the model using cross-validation
-Zn_evaluation <- workflow() %>%
-  add_recipe(recipeZn) %>%
-  add_model(gamma_spec) %>%
-  fit_resamples(resamples = samples)
-
-# Extract the results of the model
-Zn_metrics <- Zn_evaluation %>%
-  collect_metrics()
-
-###### MODEL 2 ######
-# Fit the model using cross-validation
-Zn2_evaluation <- workflow() %>%
-  add_recipe(recipeZn2) %>%
-  add_model(gamma_spec) %>%
-  fit_resamples(resamples = samples)
-
-# Extract the results of the model
-Zn2_metrics <- Zn2_evaluation %>%
-  collect_metrics()
-
-#Print the RMSE and R-square for both models
-print(Zn_metrics)
-print(Zn2_metrics)
-```
-
-For all of the models it seems like the model with all the predictors works best, instead of grouping the predcitors (exposures) in further categories.
-However, the R-square and RMSE values are really low, so maybe a different approach for modeling is required.
-
-I will do an attempt to save all the RMSE an R-squared values in one table:
-
-```{r}
-#Bind dfs that contain all RMSEs and R-squares from all models
-metrics_final <- bind_rows(Al_metrics, Al2_metrics, BC_metrics, BC2_metrics, Ca_metrics, Ca2_metrics, Fe_metrics, Fe2_metrics,
-                           K_metrics, K2_metrics, Mg_metrics, Mg2_metrics, Mn_metrics, Mn2_metrics, Si_metrics, Si2_metrics,
-                           Sul_metrics, Sul2_metrics, Ti_metrics, Ti2_metrics, Zn_metrics, Zn2_metrics)
-
-#Create a vector to name all chemical species
-species <- rep(c("Al", "BC", "Ca", "Fe", "K", "Mg", "Mn", "Si", "S", "Ti", "Zn"), each = 4)
-
-# Adding the 'chemical_species' variable to 'metrics_final'
-metrics_final$Species <- species
+#Delete non-necessary columns
+metricspm_final <- metricspm_final %>% select(-.estimator)
+metricsbc_final <- metricsbc_final %>% select(-.estimator)
 
 # Creating a vector of observation types
-obs <- rep(c(rep("N1", 2), rep("N2", 2)), length.out = 44)
+obs <- rep(c(rep("M1", 3), rep("M2", 3)), length.out = 6)
 
 # Adding the 'obs' variable to 'metrics_final'
-metrics_final$obs <- obs
-
-#Finally, delete non-necessary columns
-metrics_final <- metrics_final %>% select(-c(.estimator, n, std_err, .config))
+metricspm_final$obs <- obs
+metricsbc_final$obs <- obs
 ```
@@
 #Change dataframe from long to wide
-metrics_wide <- metrics_final %>% pivot_wider(names_from = obs, values_from = mean)
+metricspm_wide <- metricspm_final %>% pivot_wider(names_from = obs, values_from = .estimate)
+metricsbc_wide <- metricsbc_final %>% pivot_wider(names_from = obs, values_from = .estimate)
 
 #Rename columns, rearrange them and edit values for some columns
-metrics_wide <- metrics_wide %>% rename(Metric = ".metric", Model1 = "N1", Model2 = "N2") %>%
-  select(c(Species, Metric, Model1, Model2))
+metricspm_wide <- metricspm_wide %>% 
+  rename(Metric = ".metric", Linear = "M1", GLM = "M2") %>% 
+  mutate(Metric = case_when(Metric == "rmse" ~ "RMSE",
+                            Metric == "rsq" ~ "R-squared",
+                            Metric == "mae" ~ "MAE", TRUE ~ Metric))
+
+metricsbc_wide <- metricsbc_wide %>% 
+  rename(Metric = ".metric", Linear = "M1", GLM = "M2") %>% 
+  mutate(Metric = case_when(Metric == "rmse" ~ "RMSE",
+                            Metric == "rsq" ~ "R-squared",
+                            Metric == "mae" ~ "MAE", TRUE ~ Metric))
 
 #Save df as .RDS file
-saveRDS(metrics_wide, file = here("results", "tables", "model_metrics.rds"))
+saveRDS(metricspm_wide, file = here("results", "tables", "pmmodel_metrics.rds"))
+saveRDS(metricsbc_wide, file = here("results", "tables", "bcmodel_metrics.rds"))
 ```

### 2024-04-19 code/data_exploring_code.qmd
Project Part 5
Completed part 5 of the project (might review and do a few edits later)

--- 2024-03-15 code/data_exploring_code.qmd
@@
 ---
-title: "Project Data Exploring"
-subtitle: "Project: Part 2"
+title: "Data Exploring"
 author: Erick E. Mollinedo
@@
 options(scipen = 999) #to remove scientific notation for smaller values (i.e. changes from 1*10-3 to 0.001)
-ggplot(conc_longer)+
+fueltype <- ggplot(conc_longer)+
   geom_boxplot(aes(x= species, y= concentration, fill= fueltype))+
@@
 
+fueltype
+
 #Creating the boxplot of concentrations by arm
 options(scipen = 999) #to remove scientific notation for smaller values (i.e. changes from 1*10-3 to 0.001)
-ggplot(conc_longer)+
+arm <- ggplot(conc_longer)+
   geom_boxplot(aes(x= species, y= concentration, fill= arm))+
@@
   stat_pvalue_manual(stat.test, label = "p.signif", tip.length = 0.001) #Add the significance bars, use 'p.signif' to add *** or 'ns'
+
+arm
+
+#Save plots
+ggsave(here("results", "figures", "boxplot-fueltype.png"), plot = fueltype, width = 10, height = 6, dpi = 300)
+ggsave(here("results", "figures", "boxplot-arm.png"), plot = arm, width = 10, height = 6, dpi = 300)
 ```
@@
 #Create the correlation plot
-corrplot(cor, method = "color", type = "lower", #Color form and displays at the lower portion
+corplot <- corrplot(cor, method = "color", type = "lower", #Color form and displays at the lower portion
          col = COL2("RdYlBu", 20), order = "hclust", #Color palette, and order or the species
@@
 #Create a simple barplot
-ggplot(hapin_longer, aes(x= exposure, fill= response))+
+cat_expo <- ggplot(hapin_longer, aes(x= exposure, fill= response))+
   geom_bar(position = "dodge")+
   theme_classic()
+
+#Visualize plot
+cat_expo
+
+#Save plot
+ggsave(here("results", "figures", "cat-exposures.png"), plot = cat_expo, width = 10, height = 6, dpi = 300)
 ```
@@
 #Histogram plot for the PM2.5 concentrations
-ggplot(hapin, aes(x= pm25, fill= fueltype))+
+pm25 <- ggplot(hapin, aes(x= pm25, fill= fueltype))+
   geom_histogram()+
@@
 #Histogram plot for the BC concentrations
-ggplot(hapin, aes(x= bc, fill= fueltype))+
+bc <- ggplot(hapin, aes(x= bc, fill= fueltype))+
   geom_histogram()+
@@
 #Histogram plot for the CO concentrations
-ggplot(hapin, aes(x= co, fill= fueltype))+
+co <- ggplot(hapin, aes(x= co, fill= fueltype))+
   geom_histogram()+
@@
   labs(x= "Carbon Monoxide Concentration (ppm)")
+
+#Visualize plots
+pm25
+bc
+co
+
+#Save plots
+ggsave(here("results", "figures", "pm25.png"), plot = pm25, width = 10, height = 6, dpi = 300)
+ggsave(here("results", "figures", "bc.png"), plot = bc, width = 10, height = 6, dpi = 300)
+ggsave(here("results", "figures", "co.png"), plot = co, width = 10, height = 6, dpi = 300)
 ```

### 2024-04-19 products/manuscript/Manuscript.qmd
Project Part 5
Completed part 5 of the project (might review and do a few edits later)

--- 2024-03-15 products/manuscript/Manuscript.qmd
@@
 title: "Elemental characterization of personal filter samples from the Household Air Pollution Intervention Network (HAPIN) trial for Source Apportionment"
-subtitle: "Project: Part 3"
 author: Erick E. Mollinedo
@@
 library(tidyverse)
+library(gt)
 ```
+
+## Abstract
+
+Household air pollution is one the most prominent risks in low- and middle-income countries.
+Fine particulate matter (PM~2.5~) is a major air pollutant from household air pollution.
+The chemical and physical composition of PM~2.5~ depends heavily on its sources.
+Source apportionment is a methodology that aims to identify sources of pollution to reconstruct their impacts, using methods such as positive matrix factorization (PMF).
+The current project aims to conduct the initial analysis before conducting PMF modeling, to describe the most important chemical species in personal exposure filter samples from rural Guatemala.
+Six-hundred and twenty-nine PTFE filter samples from the HAPIN trial, a randomized fuel intervention controlled trial in Guatemala, were analyzed using X-Ray fluorescence to determine the presence of inorganic elements.
+Transmissometry was used to determine the black carbon (BC) concentrations and gravimetry to determine the PM~2.5~ concentrations.
+Multivariate linear regression models and gamma GLM models were used to predict which elements can be used to determine total PM~2.5~ mass concentrations and which questionnaire based exposures can infer BC mass concentrations.
+It was determined that the concentrations were significantly lower in samples from users that cooked using liquefied petroleum gas (LPG) fuel, compared to users that cooked with biomass fuels.
+The linear regression model was the most appropriate to determine the apportionment of each chemical species to total PM~2.5~ concentrations, and BC, magnesium and potassium seemed the most important predictors.
+The gamma GLM model was the most fit to predict the impact of categorical exposures on BC concentrations, where type of fuel, exposure to kerosene and a neighboring biomass stove were the most significantly associated exposures.
+However, caution should be taken into using questionnaire exposures to infer predictions.
+In addition, the ground for future source apportionment analyses was founded through these statistical analyses.
+The inclusion of additional chemical species could better inform the total composition of PM~2.5~ so PMF can be used for source apportionment.
+
+{{< pagebreak >}}
 
@@
 
-The following study aims to produce the clean dataset from the HAPIN trial in Guatemala, needed for PMF analysis and to conduct the exploratory and main data analysis to answer some questions before the PMF modelling: 1.) What chemical species are correlated among each other from these filter samples?, 2.) What are the differences in concentrations of the chemical species if the samples are categorized by type of fuel and by study arm?
-and 3.) What exposures are associated with the variations in concentrations of the most significant chemical species?
+The following study aims to produce the clean dataset from the HAPIN trial in Guatemala, needed for PMF analysis.
+This study also aims to conduct main and exploratory analysis of this chemical species dataset to answer some questions before the PMF modelling: 1.) What chemical species are correlated among each other from these filter samples?, 2.) What are the differences in concentrations of the chemical species if the samples are categorized by type of fuel and by study arm?
+3.) What chemical species are associated with the variations in concentrations of total PM~2.5~?
+and 4.) What exposures can be modeled as predictors of BC concentration?
 
@@
 In summary, the instrument works as an air pump with a standard flow of 0.3 liters per minute, so the particles get into an inlet and get collected into the filter.
-In total, six hundred and forty eight sample filters are available, where eighteen correspond to field blanks.
+In total, six hundred and forty eight sample filters were available, where eighteen correspond to field blanks.
 From the HAPIN trial study design, in total, six exposure measurements (visits) were conducted during pregnancy in around 800 participants, but for this scenario, only 648 filter samples were randomly selected for chemical species characterization.
 Other exposures where collected as categorical variables using standardized questionnaires administered in each visit.
-These exposures capture some of the sources of exposure such as tobacco smoke, trash burning, use on incense or coil, kerosene and others.
+These exposures capture some of the sources of exposure such as tobacco smoke, trash burning, use of incense or coil, kerosene, exposure to other biomass stoves and others.
 
@@
 
-The concentration of 22 chemical species (elements) on the filters were determined using X-Ray Fluorescence (XRF), and the uncertainties were estimated also based on the XRF instrument.
+The concentration of 22 chemical species (elements) on the filters were determined using a non-destructive spectrometer technique: X-Ray Fluorescence (XRF), and the uncertainties were estimated also based on the XRF instrument.
 
@@
 
+All data cleaning and processing were performed on R (version 4.3.2).
 First, all chemical species where more than 50% of the samples are below the detection limit (LOD) were filtered out, leaving a total of 10 chemical species.
@@
 
-### Statistical analysis
+### Statistical analyses
 
-The correlation (r^2^) between chemical species will be determined using either individual paired correlations or by a correlation plot matrix.
-Additional plots, such as boxplots, to visualize and compare how the concentrations differ by type of fuel.
-These differences will be assessed for statistical significance using a Welch two-sample T-test.
-Multivariate regression models will be computed to analyze which exposures could explain the concentrations of the key chemical species.
-Given that the concentrations can be approximated to a gamma distribution, the generalized linear model with gamma distribution will be used to compute the models.
-The root mean square error (RMSE) and R-squared will be the parameters to evaluate the most fit model for each of the chemical species.
+All statistical analyses were performed on R (version 4.3.2).
+An initial exploration of the data was performed, to determining missing values for all possible predictors and outcomes and trends for all chemical species.
+These explorations served in order to skim down which variables to use as predictors and outcomes and to compute their summary statistics.
 
-## Preliminary Results
+The correlation (r^2^) between chemical species were determined using paired correlations in a correlation plot matrix.
+To assess differences in concentrations between samples by type of fuel and by study arm, a Welch two-sample T-test was computed.
+Multivariate regression models were computed to assess which elements are likely to explain the variations in concentrations of total PM~2.5~ and to determine which exposures are most associated with the BC concentrations.
+For each one of the pollutants as outcomes (PM~2.5~ and BC), two different models were computed and assessed.
+Model 1 was defined as a linear multivariate regression model, meanwhile Model 2 was a Generalized linear model with gamma distribution, due to the approximation of the distributions observed for the concentrations of each chemical species.
+The PM~2.5~ model used all the chemical species to be included for PMF as predictors.
+For the BC model, the predictors were all the categorical exposures with two levels (Yes or No) registered from the questionnaires.
+The final model for each pollutant was reported and selected based on certain parameters such as the metrics, uncertainties and residuals.
 
-Six hundred and twenty-nine filter samples were considered in the final analyses.
-@fig-stats1 presents a summary of the concentrations from the biomass fueltype samples.
-Meanwhile @fig-stats2 shows the summary of concentrations from the LPG observations.
+## Results
+
+Six hundred and twenty-nine filter samples were considered in the final analyses, due to one filter sample that was damaged and the concentration was not available.
+Among the 22 chemical species, only 11 were included to be used for future PMF modeling, which means only 11 species had more than 50% of their observations above the limit of detection.
+Black carbon was also included into the chemical species, due to be a representative of a good portion of total PM~2.5~.
+Table 1 and Table 2 show a summary of the descriptive statistics for each chemical species by type of fuel.
+All the statistics are listed in ug/m^3^ units.
 As a note, all the observations of the baseline measurements were categorized as biomass fuel type, since at this point, the LPG intervention was not yet allocated.
+This explains the larger number of samples belonging to the biomass fuel type.
 
-![Summary statistics of the concentrations in the biomass group](../../results/tables/table1.jpeg){#fig-stats1}
+```{r echo=FALSE}
+#Load the summary statistics table
+biomass_table <- read_rds(here("results", "tables", "biomass-summary.rds"))
 
-![Summary statistics of the concentrations in the LPG group](../../results/tables/table2.jpeg){#fig-stats2}
+#Show the table (Biomass fueltype)
+biomass_table %>% gt() %>% tab_header(title = md("**Summary statistics: Biomass fuel samples**")) %>%
+  tab_style(style = list(cell_text(weight = "bold")),
+    locations = list(cells_title(groups = "title"),
+      cells_column_labels(columns = everything())))
+```
+
+```{r echo=FALSE}
+#Load the summary statistics table
+lpg_table <- read_rds(here("results", "tables", "lpg-summary.rds"))
+
+#Show the table (LPG fueltype)
+lpg_table %>% gt() %>% tab_header(title = md("**Summary statistics: LPG fuel samples**")) %>%
+  tab_style(style = list(cell_text(weight = "bold")),
+    locations = list(cells_title(groups = "title"),
+      cells_column_labels(columns = everything())))
+```
 
@@
 It is observed that the highest correlation is observed between Si and Fe, Ti-Fe, and Al-Si.
-These values could indicate that these species come from the same sources, so it will probably be easier to identify using the PMF software after the conclusion of this project.
+These values could indicate that these species come from the same point sources, something to account when conducting the source apportionment analysis using PMF modeling.
 
@@
 The concentration of the chemical species are distributed in different ranges depending on their type.
-For example, the highest concentrations were determined to be for Black carbon (BC), with values from 1 to 100 ug/m^3^.
+For example, the highest concentrations were determined to be for Black carbon (BC), with values from 1 to 96 ug/m^3^.
 The remaining chemical species have lower concentrations, ranging from less than 0.1 ug/m^3^ to 10 ug/m^3^.
 Other noticeable patterns is that some of the species were found in higher concentrations for personal samples from participants that used biomass stoves, compared to LPG stoves (@fig-concentrations).
-When conducting the two-sided T-test, it was determined that all of the chemical species, except Fe, Al and Zn were statistically significantly different between type of fuel.
-The differences in concentrations were also analyzed by study arm, however, given that all baseline measurements belonged to biomass fueltype observations, these results were not presented but can be observed in the Supplementary material.
+When conducting the two-sided T-test, it was determined that all of the chemical species, except Fe, Al and Zn were significantly different between type of fuel.
+The differences in concentrations were also analyzed by study arm, however, given that all baseline measurements belonged to biomass fueltype observations, these results were not presented but can be observed in the Supplementary material (Figure 1).
 
-![Chemical species concentrations by type of fuel](../../results/figures/boxplot-fueltype.jpg){#fig-concentrations}
+![Chemical species concentrations by type of fuel](../../results/figures/boxplot-fueltype.png){#fig-concentrations}
 
-### Chemical species prediction
+### Chemical species modeling
 
-Nine of the eleven exposures were evaluated as predictors of the chemical species concentrations.
-The use of a generator machine (for corn mill grinding) and incence were left out since these were the categorical exposure variables with the less variability in answers.
-The gamma distributed generalized linear models were chosen given the concentrations of all the species do not follow a normal distribution.
+#### Modeling PM~2.5~
 
-Based on the models, type of fuel was a strong predictor of concentration for eight chemical species: Mg, Mn, BC, Ca, K, S, Si and Ti.
-Kerosene was another predictor of BC and Mg concentrations alongside type of fuel and being exposed to other sources of smoke (such as traffic emissions).
-Meanwhile, smoking and other sources of smoke were high predictors of K.
-On the other hand, stove usage, independently if it is biomass or LPG, was associated with concentration of Fe.
-However, additional analyses have to be conducted for final model selection.
+The PM~2.5~ model included aluminum, calcium, iron, potassium, magnesium, manganese, sulfur, silicon, titanium, zinc and black carbon as predictors in the final model.
+Between both computed models, the linear regression model was selected as the most appropriate.
+Table 3 summarizes three metrics used for the comparisons between both models: RMSE, R-squared and MAE.
+As seen from these metrics, there was a drastic change when comparing both types of models, showing an improvement in the linear model.
+This was confirmed due to the under-prediction of certain values as seen in the observed-predicted plots and residuals (Supplementary material, Figures 5 and 6) for the GLM gamma model.
+
+Based on the coefficients and the statistical significance (p-value), it was determined that only magnesium, potassium and black carbon had a significant effect on total mass concentration of PM~2.5~.
+Black carbon and potassium had an increasing effect, meanwhile magnesium had a decreasing effect for predicting total PM~2.5~ mass concentration.
+
+```{r echo=FALSE}
+#Load the PM2.5 summary metrics table
+pmmetrics <- read_rds(here("results", "tables", "pmmodel_metrics.rds"))
+
+#Show the table (LPG fueltype)
+pmmetrics %>% gt() %>% tab_header(title = "Table 3", subtitle = md("**Metrics summary: PM2.5 Models**")) %>%
+  tab_style(style = list(cell_text(weight = "bold")),
+    locations = list(cells_title(groups = "title"),
+      cells_column_labels(columns = everything())))
+```
+
+#### Modeling Black Carbon
+
+For the model to predict black carbon concentrations, the predictors considered in the final model were: smoking, use of mosquito coils, trash burning, use of kerosene, cooking smoked foods, crop burning, exposed to a biomass stove (or a neighboring biomass stove) and type of fuel.
+Use of stove, incense and a generator for corn grinding were not included due to the small variability of answers (most or all of them coded as Yes or No).
+When comparing the linear and GLM regression models for BC, both seem to perform similarly based on the metrics (Table 4) and the residuals and observed vs predicted values plots (Supplementary: Figures 7 and 8).
+
+The gamma GLM model was selected as the most appropriate, given its ability to more accurately discriminate total mass BC concentrations from the mentioned categorical variables.
+Based on the coefficients and the p-values it was determined that type of fuel, exposure to a neighboring biomass stove and use of kerosene could be predictors of BC concentrations.
+The use of LPG as a source of fuel has a significant reduction of BC concentrations, meanwhile the use of kerosene and exposure to a neighboring biomass stove increases BC total mass concentration.
+
+```{r echo=FALSE}
+#Load the PM2.5 summary metrics table
+bcmetrics <- read_rds(here("results", "tables", "bcmodel_metrics.rds"))
+
+#Show the table (LPG fueltype)
+bcmetrics %>% gt() %>% tab_header(title = "Table 4", subtitle = md("**Metrics summary: BC Models**")) %>%
+  tab_style(style = list(cell_text(weight = "bold")),
+    locations = list(cells_title(groups = "title"),
+      cells_column_labels(columns = everything())))
+```
+
+## Discussion
+
+Although the scope of this study was not to conduct the source apportionment methodology, the findings reported here are helpful for the early stages of the overall positive matrix factorization (PMF) modeling of the HAPIN Guatemala personal exposure samples.
+Based on the limit of detection, only 10 of the 22 chemical species (elements) were deemed important to be included in the PMF model.
+The model is sensitive to the mass concentrations of all species used, that is why the remaining 11 species were left out of the analyses.
+Some important species such as lead, copper, chromium, iodine, arsenic and others could provide a better profile of the chemical composition of total PM~2.5~.
+However, since these were in trace concentrations, maybe they are not as important as others, in terms of determining the sources of PM~2.5~ in this location.
+For the chemicals that were incorporated in the final analyis, it was observed that BC had overall the highest concentrations.
+This could indicate that BC is a chemical component that comes from many different sources besides just biomass burning, which is one of the most expected exposures/sources that contribute to these samples.
+The following element in highest concentration appeared to be potassium, which again, based on literature [@rai_real-time_2020; @pervez_impact_2012], it is an indicator of biomass burning.
+
+When comparing the mass concentrations between samples from users that cooked using biomass fuels against cooking with LPG, it was observed that this provided a more accurate picture of the differences by the LPG intervention only.
+As seen in the analyses, dividing the samples in control and intervention arms produced similar results that when dividing by type of fuel.
+However there were higher concentration values observed for some species in the intervention samples, which is explained by the inclusion of all baseline measurements into this division.
+As explained in the HAPIN study description [@clasen_design_2020; @johnson_air_2020], the intervention was allocated after baseline, and during baseline it was observed that the concentrations of the main pollutants was virtually the same between both arms.
+In conclusion, it is considered that if looking at differences, categorizing by type of fuel, provides a clearer picture of the exposure contrasts.
+The correlations explored between the selected chemical species provide a glimpse of what elements could be more associated in terms of a common source.
+The more notables are iron, silicon and aluminum, that, based on the literature, are mostly associated by a crustal source of pollution [@zhou_chemical_2014].
+Crustal means ground or dust generated particles.
+Other patterns are not so easily visualized, but there is a possible association between magnesium and calcium, which could be an indicator of burning of some sort of plant-based material [@zhou_chemical_2014; @rai_real-time_2020; @pervez_impact_2012].
+Black carbon was moderately associated with potassium, which is a possible indicator of a biomass burning source, but this moderate to low correlation indicates that it could come from many other sources.
+
+In terms of model selection for predicting PM~2.5~ and BC concentrations, multivariate models were deemed the most appropriate due to the presence of many predictors.
+In the BC prediction model, the gamma GLM model seemed more appropriate compared to the linear regression model, however this may not be the most accurate and other types of model could be a better fit.
+It would be interesting to fit a LASSO regression model, in order to better discriminate which of the categorical exposures could be used to explain the variations in concentrations of BC.
+Another explanation why this model under-performed could be to the data collection process, there were just a few records of certain exposures such as trash burning, smoking, incense, such that this could have driven the model to the wrong conclusions.
+This is why it is believed that modelling exposures is more accurate when considering actual exposure measurements (pollutants or chemicals), instead of relying on questionnaire based exposures that could provide biased answers.
+On another hand, by considering only kerosene use, exposure to a neighboring biomass stove and type of fuel as accurate predictors of BC, we could infer that the model under-performed.
+However, it is also important to note that these are good representations of use of biomass fuel source, which again, could be the most significant source of pollution, instead of considering other such as vehicle emissions, crustal or others.
+
+For the PM~2.5~ models, the multivariate linear regression model seems the most appropriate as seen on the evaluation criteria (Table 3 and Supplementary material).
+However, it is important to point out that the moderate R-squared value implies that not all of the chemical species were included in the model to explain the total mass concentrations of PM~2.5~.
+Besides BC, no other organic compounds were analyzed and included in the model, and it is known that some compounds like ammonia, elemental carbon, levoglucosan and many others [@gadi_source_2019] have very important roles in particulate matter composition.
+It was surprising that the GLM model did not perform as good as expected, usually under-predicting extreme values.
+One possible way to overcome this could have been to test a multivariate linear model log-transforming the outcomes and predictors, this is due to them not following a normal distribution.
+In terms of the most significant species, the significant contribution of potassium and black carbon only add to the evidence that biomass burning could be the most important contributor of particulate matter in this location.
+
+One of the limitations for this analyses is that, as pointed before, some chemical species were not included in the model.
+This is partly because of the source of the chemical species data, in this case it was performed only transmissometry, gravimetry and x-ray fluorescence to determine some of the species.
+It would be impossible to determine all the chemical species that conform the PM~2.5~ filter samples, however, adding a chromatography analysis could improve the resolution of the model by detecting additional compounds.
+Another goal accomplished by the analyses was to compute the final concentrations and uncertainties for all the chemical species and filter samples that will be analyzed using PMF.
+Currently, the PMF algorithm is licensed by the Environmental Protection Agency, under the EPA PMF 5.0 software [@norris_positive_nodate].
+There was an attempt to compute the whole PMF analysis for source apportionment in R, however, this algorithm is not available publicly due to the patent.
+However, there is one specific packages 'pmfr' found online in a Github repository (https://github.com/skgrange/pmfr) that allows working with the outputs of the EPA PMF 5.0 software to do additional analysis and plotting.
+
+## Conclusions
+
+Twelve chemical species from the filter samples in rural Guatemala were found important for PMF modeling: Mg, Mn, Al, Si, S, Ca, K, Zn, Ti, Fe and BC.
+Of these species, all of them except Al, Zn and Fe had significantly smaller mass concentrations in LPG fuel samples compared to biomass fuel samples.
+BC and K are two of the most available chemical species found in the samples, which can be associated to the heavy influence of biomass burning in this location.
+When modelling which exposures were associated to BC concentrations, kerosene use, exposure to a neighboring biomass stove and type of fuel were significant predictors according to a GLM gamma model.
+However, the low diversity in answers for these categorical variables could have biased the results.
+On the other hand, a linear regression model was found more accurate to define total PM~2.5~ mass concentrations with all the chemical species.
+BC, potassium and magnesium were the most significant predictors in the PM~2.5~ model.
+The current study also sets the final concentration and uncertainty values needed to perform source apportionment using the PMF algorithm developed by the Environmental Protection Agency.
 

### 2024-04-19 products/manuscript/supplement/Supplementary.qmd
Project Part 5
Completed part 5 of the project (might review and do a few edits later)

--- 2024-03-15 products/manuscript/supplement/Supplementary.qmd
@@
 title: "Supplementary Material"
-subtitle: "Project: Part 3"
 author: Erick E. Mollinedo
@@
 @fig-conc shows the summary of concentrations for each chemical species categorized by study arm.
-It is observed a similar pattern observed from the categorization by type of fuel, where Mg, Mn, Ca, K, BC, Ti, Si and S have statistically significant differences in concentrations (shown as ****).
+It is observed a similar pattern observed from the categorization by type of fuel, where Mg, Mn, Ca, K, BC, Ti, Si and S have statistically significant differences in concentrations (shown as \*\*\*\*).
 
-![Chemical species concentrations by study arm](../../../results/figures/boxplot-arm.jpg){#fig-conc}
+![Chemical species concentrations by study arm](../../../results/figures/boxplot-arm.png){#fig-conc}
 
@@
 
-![Carbon Monoxide concentrations by type of fuel](../../../results/figures/co.jpg){#fig-co}
+![Carbon Monoxide concentrations by type of fuel](../../../results/figures/co.png){#fig-co}
 
-![PM2.5 concentrations by type of fuel](../../../results/figures/pm25.jpg){#fig-pm}
+![PM2.5 concentrations by type of fuel](../../../results/figures/pm25.png){#fig-pm}
 
-The following tables show the results when conducting the generalized linear models using a single or multiple variables for the chemical species that can be interpreted by sources of exposure.
+The following tables show the model summaries for the final models selected.
+In this case the linear regression model for the PM~2.5~ mass concentrations and the gamma GLM model for the BC mass concentrations.
 
 ```{r echo=FALSE}
-#Open the hapin_merged df
+#Load the `hapin_merged` dataframe
+hapin_merged <- read_rds(here("data", "processed-data", "hapin-final.rds"))
 
-hapin_merged <- read_csv(here("data", "processed-data", "hapin-merged.csv"))
+# Split the data into training and testing sets
+rngseed = 1234
+set.seed(rngseed) # for reproducibility
+data_split <- initial_split(hapin_merged, prop = 0.8)
+train_data <- training(data_split)
+test_data <- testing(data_split)
 
-# Create a generalized linear regression model specification using gamma distribution (This will be used for all the models)
-model_spec <- linear_reg(mode = "regression") %>%
+#Create formulas
+pm25formula <- pm25 ~ Mg + Al + Si + S + K + Ca + Ti + Mn + Fe + Zn + BC
+bcformula <- BC ~ smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype
+
+#Set model specifications
+lin_mod <- linear_reg() %>% set_engine("lm") #Linear
+glm_mod <- linear_reg(mode = "regression") %>% #GLM gamma
   set_engine("glm", family = Gamma(link = "log"))
 
-###########To predict the BC variable##############
-# Create the recipe for preprocessing
-recipeBC_red <- recipe(BC ~ kerosene + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
+## ----Final model PM2.5: Linear regression---- ##
+linpm_wflow <- workflow() %>% #Workflow
+	add_model(lin_mod) %>% 
+	add_formula(pm25formula)
 
-# Create a workflow
-workflowBC_red <- workflow() %>%
-  add_recipe(recipeBC_red) %>%
-  add_model(model_spec)
+linpm_fit <- linpm_wflow %>% fit(data = train_data) #Fit the model
 
-# Fit the model to the data
-fitBC_red <- workflowBC_red %>%
-  fit(data = hapin_merged)
+## ----Final model BC: GLM Gamma---- ##
+glmbc_wflow <- workflow() %>% #Workflow
+  add_model(glm_mod) %>% 
+  add_formula(bcformula)
 
-# Extract and print the results of the model
-resultsBC_red <- fitBC_red %>%
-  pull_workflow_fit() %>%
+glmbc_fit <- glmbc_wflow %>% fit(data = train_data) #Fit the model
+
+# Extract and print the results of the models
+results_pmlin <- linpm_fit %>% #PM2.5 Linear
+  extract_fit_parsnip() %>%
   tidy()
 
-print(resultsBC_red)
-
-###########To predict the Ca variable##############
-# Create the recipe for preprocessing
-recipeCa_red <- recipe(Ca ~ fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-# Create a workflow
-workflowCa_red <- workflow() %>%
-  add_recipe(recipeCa_red) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitCa_red <- workflowCa_red %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsCa_red <- fitCa_red %>%
-  pull_workflow_fit() %>%
+results_bcglm <- glmbc_fit %>% #BC GLM
+  extract_fit_parsnip() %>%
   tidy()
 
-print(resultsCa_red)
+#Print the results
+print(results_pmlin)
+print(results_bcglm)
+```
 
-###########To predict the K variable##############
-# Create the recipe for preprocessing
-recipeK_red <- recipe(K ~ smoke + stove_other + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
+The following figures illustrate the observed vs predicted values plots for both PM~2.5~ and BC models, and the residual plots.
 
-# Create a workflow
-workflowK_red <- workflow() %>%
-  add_recipe(recipeK_red) %>%
-  add_model(model_spec)
+![Observed vs Predicted values of the PM2.5 models](../../../results/figures/obs-pred_pm25.png)
 
-# Fit the model to the data
-fitK_red <- workflowK_red %>%
-  fit(data = hapin_merged)
+![PM2.5 Models residuals](../../../results/figures/residuals-pm25.png)
 
-# Extract and print the results of the model
-resultsK_red <- fitK_red %>%
-  pull_workflow_fit() %>%
-  tidy()
+![Observed vs Predicted values of the BC models](../../../results/figures/obs-pred_bc.png)
 
-print(resultsK_red)
-
-###########To predict the Mg variable##############
-# Create the recipe for preprocessing
-recipeMg_red <- recipe(Mg ~ kerosene + fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-# Create a workflow
-workflowMg_red <- workflow() %>%
-  add_recipe(recipeMg_red) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitMg_red <- workflowMg_red %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsMg_red <- fitMg_red %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsMg_red)
-
-###########To predict the Mn variable##############
-# Create the recipe for preprocessing
-recipeMn_red <- recipe(Mn ~ fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-# Create a workflow
-workflowMn_red <- workflow() %>%
-  add_recipe(recipeMn_red) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitMn_red <- workflowMn_red %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsMn_red <- fitMn_red %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsMn_red)
-
-###########To predict the S variable##############
-# Create the recipe for preprocessing
-recipeS_red <- recipe(S ~ fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-# Create a workflow
-workflowS_red <- workflow() %>%
-  add_recipe(recipeS_red) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitS_red <- workflowS_red %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsS_red <- fitS_red %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsS_red)
-
-###########To predict the Fe variable##############
-# Create the recipe for preprocessing
-recipeFe_red <- recipe(Fe ~ stove, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-# Create a workflow
-workflowFe_red <- workflow() %>%
-  add_recipe(recipeFe_red) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitFe_red <- workflowFe_red %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsFe_red <- fitFe_red %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsFe_red)
-
-###########To predict the Si variable##############
-# Create the recipe for preprocessing
-recipeSi_red <- recipe(Si ~ fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-# Create a workflow
-workflowSi_red <- workflow() %>%
-  add_recipe(recipeSi_red) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitSi_red <- workflowSi_red %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsSi_red <- fitSi_red %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsSi_red)
-
-###########To predict the Ti variable##############
-# Create the recipe for preprocessing
-recipeTi_red <- recipe(Ti ~ fueltype, data = hapin_merged) %>%
-  step_dummy(all_nominal(), -all_outcomes())
-
-# Create a workflow
-workflowTi_red <- workflow() %>%
-  add_recipe(recipeTi_red) %>%
-  add_model(model_spec)
-
-# Fit the model to the data
-fitTi_red <- workflowTi_red %>%
-  fit(data = hapin_merged)
-
-# Extract and print the results of the model
-resultsTi_red <- fitTi_red %>%
-  pull_workflow_fit() %>%
-  tidy()
-
-print(resultsTi_red)
-```
+![BC models residuals](../../../results/figures/residuals-bc.png)

### 2024-04-25 project-review-Hu.qmd

---
title: Project Review for Mollinedo
author: Xueyan Hu
date: 04-24-2024
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: Elemental characterization of personal filter samples from the Household Air Pollution Intervention Network (HAPIN) trial for Source Apportionment

Name of project author(s): Erick E. Mollinedo

Name of project reviewer: Xueyan Hu

# Instructions

Write your comments and feedback below for each section/component of the project. The goal should be to help the author improve their project. Make comments as constructive and actionable as possible. You can provide both criticism and praise.

For each component, pick one summary statement by deleting the ones that do not apply and keeping only the one that you think most closely summarizes a given component.

Make sure your final document compiles/knits into a readable, well-formatted html document.

Delete any sections/text of this template that are not part of your final review document. (Including these instructions.)

# Specific project content evaluation
Evaluate the different parts of the project by filling in the sections below.

## Background, Context and Motivation
How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments

There is an detailed background description and it is easy to understand the current condition of household air pollution. And the transition from introduction to the hypothesis/question is natural and they are highly related.

### Summary assessment (PICK ONE, DELETE THE OTHERS)

* strong contextualization and motivation

## Question description
How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments

The questions raised are clear enough and they provide me a rough frame that how the dataset would be organized, what variables it might include, and what kind of analysis the author would apply to the dataset.

### Summary assessment

* question/hypotheses fully clear

## Data description
How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

### Feedback and Comments

The source is provided and the calculating equation is also listed. Although I could have a rough understanding for the raw data, it will be a little better if a codebook is created (I am sorry if I miss this file.)

### Summary assessment

* source and overall structure of data somewhat explained

## Data wrangling and exploratory analysis
How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments

The data cleaning process is well organized, all of the cleaning steps are well illustrated, and the processed data looks good to me.

### Summary assessment

* essentially no weaknesses in wrangling and exploratory component

## Appropriateness of Analysis
Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

The analysis methods are appropriate for the data in this project according to the type of outcome. And there is a specific model related to environmental analysis involved in this project, different from what models we usually learn during the course. I think it is suitable for data analysis of this project.

### Summary assessment

* strong and reasonable analysis

## Presentation
How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments

Tables are used for presenting the result. A summary is written firstly and then it is divided into several parts for further description. Figure 1 is impressive to me because tt allows me to intuitively find the relationship between variables, especially in the condition that many variables are included it seems really important. The metircs are proper and supplement file is well used.

### Summary assessment

* results are very well presented

## Discussion/Conclusions
Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments

The study findings are properly discussed with strengths and limitations mentioned, such as linear model can fit the data well and low diversity in answers for categorical variables could lead to bias. And I think both are reasonable.

### Summary assessment

* strong, complete and clear discussion

## Further comments

_Overall, this is a mature, highly completed and meaningful project. From data collection to data cleaning, to the use and supplement of all interpreted figures in the model, every part shows that the project is well organized, and we can also see how much thinking and time the author spent on the project.But there is only one small thing that I would like to mention. There is a folder called R which is created in the template by Dr. Handel. The author haven't used any file in this folder, so I think it could be deleted._

# Overall project content evaluation
Evaluate overall features of the project  by filling in the sections below.

## Structure
Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments

The project is well structured and files are well labeled in the folders. So it is quite easy for me to find all of the relative materials. All of the files have reasonable name. But there is only one folder named 'R' left which should be deleted. And under the result file, maybe 'large file' folder and 'output' folder could also be deleted.

### Summary assessment

* mostly clear, but some confusing parts (e.g. useless files, things in the wrong folders)

## Documentation 
How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

Yes, I think reading through data analysis qmd file, coupled with the contents of the manuscript, I can understand the whole process and don't have much difficulty. But the process of data cleaning is kind of complicated, and it needs to be reviewed back and forth for better understanding, but this is only my own situation.

### Summary assessment

* fully and well documented

## Reproducibility
Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

I have tried all the codes involved in this project and they are fully reproducible. But I haven't found the instruction for reproducing this whole project in the supplement file.
I think it is "fully reproducible without issues" but it will be better if some instructions could be provided.

### Summary assessment

* fully reproducible without issues

## Thoroughness
How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments

The overall study is thorough, and both linear and logistic models are used to fit the dataset and the performances are compared, evaluated and discussed in different files. And the result mentioning predictors and the better model that can answer the questions raised above.

### Summary assessment

* strong level of thoroughness

## Further comments

_I just have a small question. Because only simple linear model and glm are fit in this project, plus the number of variables involved in the data is relatively larger, I wonder if more models could be tried to fit the data to compare the performance and maybe the overfitting problem could also be considered in the discussion part._

### 2024-04-26 project-review-Robertson.qmd

---
title: Project Review Robertson 
author: Rachel Robertson
date: 4-26-24
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: "Elemental characterization of personal filter samples from the Household Air Pollution Intervention Network (HAPIN) trial for Source Apportionment"

Name of project author(s): Erick Mollinedo

Name of project reviewer: Rachel Robertson

# Specific project content evaluation

## Background, Context and Motivation

>How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments

Great work at providing context for the existing work in the HAPIN trial and using that rationale of providing a basis for SA (or PMF)! I think that the study objectives are very clear! I just have a few comments below:

Great work establishing the importance of studying HAPs. I think that adding statistics about the impact of HAPs in Guatemala (the country of interest), might strengthen the rationale for your study.

I want to point out that you also do not define LMICs before using the acronym. This might be useful information to add.

Why was the HAPIN trial interested in studying HAP effects on pregnant women? I know that this is to measure the health outcomes of infants (e.g. birthweights), but this might be useful to specify in your background.

### Summary assessment

-   strong contextualization and motivation

## Question description

>How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments

The research questions are very clear and straight-forward. I think that question number 2) almost sounds like two questions, because you are categorizing by type of fuel and also by study arm.

Do you have any hypothesis about these questions or is your analysis mainly exploratory?

### Summary assessment

-   question/hypotheses fully clear

## Data description

>How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

### Feedback and Comments

Amazing description of where your data came from and how it was collected! However, I did not see a codebook for all of the pollutants included in the original analysis. This might be useful to add to your data file or to the readme in your data file.

### Summary assessment

-   source and overall structure of data well explained

## Data wrangling and exploratory analysis

>How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments

Great work explaining each step of your eda and wrangling with rationale. I have a few questions for you below:

Why were the LOD values that were too low imputed with LOD/ sqrt(2)?
I think that it is cool that you added your uncertainties to the data for the remaining predictors! You provide good rationale for this as well.

I think you did a great job at including the necessary EDA without providing an overwhelming amount of figures. What is the significance of correlation between certain metal pollutants?

I am also curious about whether you determined the pollutant's importance just off concentration or also on potency. Should each pollutant have a certain weight if they're more potent or should the cut-off be at a significant level of toxicity (which might mean higher concentrations for some pollutants than others)?

### Summary assessment

-   essentially no weaknesses in wrangling and exploratory component

## Appropriateness of Analysis

>Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

Overall, from looking at the code and manuscript it seems that the analysis was done properly, but I do not entirely understand What is the rationale for choosing to compare a generalized gamma linear regression and a multivariate linear regression. I am wondering what other kinds of analysis might be appropriate to analyze environmental pollutants. Do these models account for the non-normal distribution of each pollutant?

### Summary assessment

-   defensible but not optimal analysis

## Presentation

>How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments

The figures and tables are beautiful and easy to interpret using the labels, title, and caption! They are nicely formatted and publication quality. I feel like the only thing you may consider including are figures from the exploratory analysis that demonstrate the non-normal distribution of the pollutant data.

### Summary assessment

-   results are very well presented

## Discussion/Conclusions

>Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments

Thank you for explaining the potential alternative sources of some important pollutants in the discussion. I also like how you differentiated between clinical significance and importance in SA models. This actually answers my question above, about potency.

I like how you discuss the limitations of data collection through questionnaire. I also like that you are critical of your analysis method and that it might not be the ideal analysis method.

Despite these limitations, you support that biomass burning is the most significant source of pollution in your study area, which is important.

Should the limitations to the study's data collection be included in your final limitations paragraph (with your analysis limitations)? They should still be included, but I think that perhaps adding the limitations of the collection method to the paragraph that you discuss the HAPIN limitations (such as questionnaire data collection) might be more appropriate.

### Summary assessment

-   strong, complete and clear discussion

## Further comments

I think it might also be useful to add more detail about the significance of yoru findings and how they may specifically be used for PMF analysis in the future.

# Overall project content evaluation

## Structure

> Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments

There is a copy of the original 'code' file in the 'R' file in your repo (containing code templates). You may consider deleting this, as it caused me some initial confusion. It might also be useful to create separate folders for each of your code files and associated documents (e.g. cleaning, eda, and analysis folders). It may also be useful to create sub-folders in your results/figures folder to organize the figures from each step of your project (e.g. eda figures, model output, model assessment). You may also delete the unused sub-folders in the results folder.

### Summary assessment

-   mostly clear, but some confusing parts (e.g. useless files, things in the wrong folders)

## Documentation

> How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

I appreciated that the data cleaning file was explained very thoroughly, with an explanation for each chunk of cleaning steps. In addition, you added labels between lines of codes to specify the functions' purposes.
You provide rationale for why you do each step of your exploratory analysis. Most of the time, one goes into EDA without much prior knowledge of the data, but your prior knowledge helped you know what to look for during your exploration. Because of this you have well-informed steps in your EDA. It also allowed you to make informed decisions about which model you chose. Great work!
Again, amazing rationale and output for your analysis code! My only comment is that you may want to include some interpretation under your model performance sections, but this is not necessary.
I also have one question;
I was curious why you chose a linear model knowing that the concentration of each pollutant is not normally distributed. Do you think a GAM might be able to capture differences in pollutant distribution trends?

### Summary assessment

-   fully and well documented

## Reproducibility

>Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

There was no need for manual intervention at all! All of the code ran very smoothly for my computer and each step was well documented. Great work!

### Summary assessment

-   fully reproducible without issues

## Thoroughness

>How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments

As you mentioned in your discussion, another model method, such as LASSO, could be assessed. However, you did consider the pros and cons of the alternatives versus your chosen methods. I would have just liked it if you provided more explanation for why you chose the analysis methods that you used.

I also think that your research questions 1) and 2) could use some more explanation in the discussion, according to what you found in your results.

### Summary assessment

-   decent level of thoroughness

### 2024-04-26 project-review-Kaggwa.qmd_Erick.qmd
Review by Patrick
Including review by Patrick K.

--- 2024-04-25 project-review-Hu.qmd
@@
 ---
-title: Project Review for Mollinedo
-author: Xueyan Hu
+title: Project Review Template 
+author: Patrick Kaggwa
 date: 04-24-2024
@@
 
-Name of project reviewer: Xueyan Hu
+Name of project reviewer: Patrick Kaggwa
 
@@
 
-Make sure your final document compiles/knits into a readable, well-formatted html document.
+Make sure your final document compiles/renders into a readable, well-formatted html document.
 
@@
 
-# Specific project content evaluation
-Evaluate the different parts of the project by filling in the sections below.
+# Specific project content evaluation.
 
@@
 
-There is an detailed background description and it is easy to understand the current condition of household air pollution.
-And the transition from introduction to the hypothesis/question is natural and they are highly related.
+* Context of the Project: 
+The project provides a solid background on household air pollution (HAP) in low- and middle-income countries (LMICs).
+While the background offers a decent summary of previous work, it could be strengthened with more explicit referencing.
+The project is reasonably well-placed within the existing scientific and public health discourse.
+The introduction effectively outlines why the project was undertaken by highlighting the significance of HAP, the limited research on SA in LMICs.
+
+* Areas of improvement 
+*Conciseness: 
+Some parts of the introduction and background sections contain detailed explanations that could be condensed for brevity without sacrificing clarity. 
+* Citations:
+While citations are provided throughout the text, there are instances where additional citations could strengthen the claims made. 
+* Integration of Pilot Study Findings: 
+The introduction briefly mentions the pilot study conducted in Guatemala and Rwanda as part of the HAPIN trial, but it could be more explicit in integrating the findings of this pilot study into the rationale for the current research.
 
@@
 ### Feedback and Comments
-
-The questions raised are clear enough and they provide me a rough frame that how the dataset would be organized, what variables it might include, and what kind of analysis the author would apply to the dataset.
+The questions and hypotheses that the project aims to address are fairly clear and well-described.
 
@@
 ### Feedback and Comments
+Data Source is well described and the text provides a detailed description of how the data was collected and processed.
 
-The source is provided and the calculating equation is also listed.
-Although I could have a rough understanding for the raw data, it will be a little better if a codebook is created (I am sorry if I miss this file.)
+#Codebook provided: 
+The code book is loaded well
 
 ### Summary assessment
-
-* source and overall structure of data somewhat explained
+* source and overall structure of data well explained
 
@@
 
-The data cleaning process is well organized, all of the cleaning steps are well illustrated, and the processed data looks good to me.
+The exploratory data analysis appears sound and is thoroughly documented. The goals for each step were evident, and the coding and procedures were straightforward and easily understandable.
 
@@
 ### Feedback and Comments
-
-The analysis methods are appropriate for the data in this project according to the type of outcome.
-And there is a specific model related to environmental analysis involved in this project, different from what models we usually learn during the course.
-I think it is suitable for data analysis of this project.
-
+The analysis appears to be done properly, with appropriate methods used for each component of the analysis. The code is well-explained and provides clear justification for the chosen approaches.
 ### Summary assessment
@@
 
-Tables are used for presenting the result.
-A summary is written firstly and then it is divided into several parts for further description.
-Figure 1 is impressive to me because tt allows me to intuitively find the relationship between variables, especially in the condition that many variables are included it seems really important.
-The metircs are proper and supplement file is well used.
+Overall, the presentation of the results is well-executed, striking a good balance between textual explanation and visual representation, which enhances the clarity and accessibility of the findings.
 
@@
 
-The study findings are properly discussed with strengths and limitations mentioned, such as linear model can fit the data well and low diversity in answers for categorical variables could lead to bias.
-And I think both are reasonable.
+Overall, the discussion effectively addresses the study findings, acknowledges both strengths and limitations, and provides a proper interpretation of the results, contributing to a comprehensive understanding of the research outcomes.
 
@@
 
-_Overall, this is a mature, highly completed and meaningful project.
-From data collection to data cleaning, to the use and supplement of all interpreted figures in the model, every part shows that the project is well organized, and we can also see how much thinking and time the author spent on the project.But there is only one small thing that I would like to mention.
-There is a folder called R which is created in the template by Dr.
-Handel.
-The author haven't used any file in this folder, so I think it could be deleted._
+Overall, the project demonstrates a strong understanding of the subject matter and presents the findings in a coherent and thoughtful manner.
 
@@
 ### Feedback and Comments
-
-The project is well structured and files are well labeled in the folders.
-So it is quite easy for me to find all of the relative materials.
-All of the files have reasonable name.
-But there is only one folder named 'R' left which should be deleted.
-And under the result file, maybe 'large file' folder and 'output' folder could also be deleted.
+The project is well structured but it still has default files and folders which should removed.
 
 ### Summary assessment
-
-* mostly clear, but some confusing parts (e.g. useless files, things in the wrong folders)
+* well structured
 
@@
 
-Yes, I think reading through data analysis qmd file, coupled with the contents of the manuscript, I can understand the whole process and don't have much difficulty.
-But the process of data cleaning is kind of complicated, and it needs to be reviewed back and forth for better understanding, but this is only my own situation.
+Overall, while the project is reasonably well-documented and understandable.
 
@@
 
-I have tried all the codes involved in this project and they are fully reproducible. But I haven't found the instruction for reproducing this whole project in the supplement file.
-I think it is "fully reproducible without issues" but it will be better if some instructions could be provided.
+Everything run smoothly and it reproducible.
 
@@
 
-The overall study is thorough, and both linear and logistic models are used to fit the dataset and the performances are compared, evaluated and discussed in different files.
-And the result mentioning predictors and the better model that can answer the questions raised above.
+Overall the study is thorough.
 
@@
 
-* strong level of thoroughness
+* strong level of thorougness
 
@@
 
-_I just have a small question.
-Because only simple linear model and glm are fit in this project, plus the number of variables involved in the data is relatively larger, I wonder if more models could be tried to fit the data to compare the performance and maybe the overfitting problem could also be considered in the discussion part._
+_I feel justice was done but remove all unused folders and files as they still exist_

### 2024-04-29 products/manuscript/Manuscript.qmd
Project Part 6 and final
Incorporated final edits to the project (maybe)

@@
 ---
-title: "Elemental characterization of personal filter samples from the Household Air Pollution Intervention Network (HAPIN) trial for Source Apportionment"
+title: "Chemical characterization of personal filter samples from the Household Air Pollution Intervention Network (HAPIN) trial for Source Apportionment"
 author: Erick E. Mollinedo
@@
 HAP has been listed as the second environmental health risk factor worldwide, just after outdoor air pollution.
-Exposure to HAP has been linked to more than 2.31 million deaths in 2019 and it is associated with adverse cardiovascular and respiratory health outcomes [@murray_global_2020; @pope_health_2006].
+Exposure to HAP has been linked to more than 2.31 million deaths globally in 2019, and it is more prominent in LMICs.
+An example of this is Guatemala with an estimate of 29.79 deaths per 100,000 inhabitants [@murray_global_2020].
+HAP is associated with adverse cardiovascular and respiratory health outcomes [@pope_health_2006].
 Particulate matter with an aerodynamic diameter less than 2.5 micrometers (PM~2.5~) is one of the major air pollutants chemically and physically composed of crustal material, black carbon (BC), organic compounds, metals and others.
@@
 The following study aims to produce the clean dataset from the HAPIN trial in Guatemala, needed for PMF analysis.
-This study also aims to conduct main and exploratory analysis of this chemical species dataset to answer some questions before the PMF modelling: 1.) What chemical species are correlated among each other from these filter samples?, 2.) What are the differences in concentrations of the chemical species if the samples are categorized by type of fuel and by study arm?
-3.) What chemical species are associated with the variations in concentrations of total PM~2.5~?
+This study also aims to conduct main and exploratory analysis of this chemical species dataset to answer some questions before the PMF modelling: 1.) What chemical species are correlated among each other from these filter samples?, 2.) What are the differences in concentrations of the chemical species if the samples are categorized by type of fuel?, 3.) What chemical species are associated with the variations in concentrations of total PM~2.5~?
 and 4.) What exposures can be modeled as predictors of BC concentration?
@@
 
+\
+
 ```{r echo=FALSE}
@@
 When conducting the two-sided T-test, it was determined that all of the chemical species, except Fe, Al and Zn were significantly different between type of fuel.
-The differences in concentrations were also analyzed by study arm, however, given that all baseline measurements belonged to biomass fueltype observations, these results were not presented but can be observed in the Supplementary material (Figure 1).
+The differences in concentrations were also analyzed by study arm, these results can be observed in the Supplementary material (Figure 1).
 
@@
 #Show the table (LPG fueltype)
-pmmetrics %>% gt() %>% tab_header(title = "Table 3", subtitle = md("**Metrics summary: PM2.5 Models**")) %>%
+pmmetrics %>% gt() %>% tab_header(title = md("**Metrics summary: PM2.5 Models**")) %>%
   tab_style(style = list(cell_text(weight = "bold")),
@@
 
-For the model to predict black carbon concentrations, the predictors considered in the final model were: smoking, use of mosquito coils, trash burning, use of kerosene, cooking smoked foods, crop burning, exposed to a biomass stove (or a neighboring biomass stove) and type of fuel.
+For the model to predict BC concentrations, the predictors considered in the final model were: smoking, use of mosquito coils, trash burning, use of kerosene, cooking smoked foods, crop burning, exposed to a biomass stove (or a neighboring biomass stove) and type of fuel.
 Use of stove, incense and a generator for corn grinding were not included due to the small variability of answers (most or all of them coded as Yes or No).
@@
 #Show the table (LPG fueltype)
-bcmetrics %>% gt() %>% tab_header(title = "Table 4", subtitle = md("**Metrics summary: BC Models**")) %>%
+bcmetrics %>% gt() %>% tab_header(title = md("**Metrics summary: BC Models**")) %>%
   tab_style(style = list(cell_text(weight = "bold")),
@@
 
-Although the scope of this study was not to conduct the source apportionment methodology, the findings reported here are helpful for the early stages of the overall positive matrix factorization (PMF) modeling of the HAPIN Guatemala personal exposure samples.
+Although the scope of this study was not to conduct the source apportionment methodology, the findings reported here are helpful for the early stages of the positive matrix factorization (PMF) modeling of the HAPIN Guatemala personal exposure samples.
 Based on the limit of detection, only 10 of the 22 chemical species (elements) were deemed important to be included in the PMF model.
@@
 However, since these were in trace concentrations, maybe they are not as important as others, in terms of determining the sources of PM~2.5~ in this location.
-For the chemicals that were incorporated in the final analyis, it was observed that BC had overall the highest concentrations.
+For the chemicals that were incorporated in the final analysis, it was observed that BC had overall the highest concentrations.
 This could indicate that BC is a chemical component that comes from many different sources besides just biomass burning, which is one of the most expected exposures/sources that contribute to these samples.
@@
 
-When comparing the mass concentrations between samples from users that cooked using biomass fuels against cooking with LPG, it was observed that this provided a more accurate picture of the differences by the LPG intervention only.
 As seen in the analyses, dividing the samples in control and intervention arms produced similar results that when dividing by type of fuel.
-However there were higher concentration values observed for some species in the intervention samples, which is explained by the inclusion of all baseline measurements into this division.
-As explained in the HAPIN study description [@clasen_design_2020; @johnson_air_2020], the intervention was allocated after baseline, and during baseline it was observed that the concentrations of the main pollutants was virtually the same between both arms.
+The high concentrations observed for some species in the intervention samples is explained by the inclusion of all baseline measurements as part of the analyses.
+As explained in the HAPIN study design [@clasen_design_2020; @johnson_air_2020], the intervention was allocated after baseline, and during baseline it was observed that the concentrations of the main pollutants was virtually the same between both arms.
 In conclusion, it is considered that if looking at differences, categorizing by type of fuel, provides a clearer picture of the exposure contrasts.
@@
 Other patterns are not so easily visualized, but there is a possible association between magnesium and calcium, which could be an indicator of burning of some sort of plant-based material [@zhou_chemical_2014; @rai_real-time_2020; @pervez_impact_2012].
-Black carbon was moderately associated with potassium, which is a possible indicator of a biomass burning source, but this moderate to low correlation indicates that it could come from many other sources.
+BC was moderately associated with K, which is a possible indicator of a biomass burning source, but this level of correlation indicates that it could come from many other sources.
 
@@
 Another explanation why this model under-performed could be to the data collection process, there were just a few records of certain exposures such as trash burning, smoking, incense, such that this could have driven the model to the wrong conclusions.
-This is why it is believed that modelling exposures is more accurate when considering actual exposure measurements (pollutants or chemicals), instead of relying on questionnaire based exposures that could provide biased answers.
+This is why it is believed that modelling exposures is more accurate when considering actual exposure measurements (pollutants or chemicals), instead of relying on questionnaire based exposures that could bias the results.
 On another hand, by considering only kerosene use, exposure to a neighboring biomass stove and type of fuel as accurate predictors of BC, we could infer that the model under-performed.
@@
 However, it is important to point out that the moderate R-squared value implies that not all of the chemical species were included in the model to explain the total mass concentrations of PM~2.5~.
-Besides BC, no other organic compounds were analyzed and included in the model, and it is known that some compounds like ammonia, elemental carbon, levoglucosan and many others [@gadi_source_2019] have very important roles in particulate matter composition.
+Besides BC, no other organic compounds were analyzed and included in the model, and it is known that some compounds like ammonia, elemental carbon, levoglucosan and others [@gadi_source_2019] have very important roles in PM composition.
 It was surprising that the GLM model did not perform as good as expected, usually under-predicting extreme values.
-One possible way to overcome this could have been to test a multivariate linear model log-transforming the outcomes and predictors, this is due to them not following a normal distribution.
-In terms of the most significant species, the significant contribution of potassium and black carbon only add to the evidence that biomass burning could be the most important contributor of particulate matter in this location.
+One possible way to overcome this could have been to test a multivariate linear model log-transforming the outcomes and predictors due to them not following a normal distribution.
+In terms of the most significant species, the significant contribution of K and BC only add to the evidence that biomass burning could be the most important contributor of particulate matter in this location.
+In general, the results from these models suggest that biomass burning will be a major source of PM~2.5~ pollution, something to account when analyzing the PMF results.
+Unfortunately, the list of categorical exposure was not wide, as to include outdoor sources such as traffic emissions or any other environmental exposures that could provide evidence of sources when conducting the future source apportionment study.
 
 One of the limitations for this analyses is that, as pointed before, some chemical species were not included in the model.
-This is partly because of the source of the chemical species data, in this case it was performed only transmissometry, gravimetry and x-ray fluorescence to determine some of the species.
+This is partly because the chemical species were determined only by transmissometry, gravimetry and x-ray fluorescence.
 It would be impossible to determine all the chemical species that conform the PM~2.5~ filter samples, however, adding a chromatography analysis could improve the resolution of the model by detecting additional compounds.
 Another goal accomplished by the analyses was to compute the final concentrations and uncertainties for all the chemical species and filter samples that will be analyzed using PMF.
-Currently, the PMF algorithm is licensed by the Environmental Protection Agency, under the EPA PMF 5.0 software [@norris_positive_nodate].
+Currently, the PMF algorithm is licensed by the Environmental Protection Agency using the PMF 5.0 software [@norris_positive_nodate].
 There was an attempt to compute the whole PMF analysis for source apportionment in R, however, this algorithm is not available publicly due to the patent.
@@
 
-Twelve chemical species from the filter samples in rural Guatemala were found important for PMF modeling: Mg, Mn, Al, Si, S, Ca, K, Zn, Ti, Fe and BC.
-Of these species, all of them except Al, Zn and Fe had significantly smaller mass concentrations in LPG fuel samples compared to biomass fuel samples.
+Eleven chemical species from the filter samples in rural Guatemala were found important for PMF modeling: Mg, Mn, Al, Si, S, Ca, K, Zn, Ti, Fe and BC.
+Of these species, all of them except Al, Zn and Fe had significantly lower mass concentrations in LPG compared to biomass fuel samples.
 BC and K are two of the most available chemical species found in the samples, which can be associated to the heavy influence of biomass burning in this location.
 When modelling which exposures were associated to BC concentrations, kerosene use, exposure to a neighboring biomass stove and type of fuel were significant predictors according to a GLM gamma model.
-However, the low diversity in answers for these categorical variables could have biased the results.
-On the other hand, a linear regression model was found more accurate to define total PM~2.5~ mass concentrations with all the chemical species.
-BC, potassium and magnesium were the most significant predictors in the PM~2.5~ model.
-The current study also sets the final concentration and uncertainty values needed to perform source apportionment using the PMF algorithm developed by the Environmental Protection Agency.
+However, the low diversity in answers for these categorical exposures could have biased the results.
+Also, a linear regression model was found more accurate to define total PM~2.5~ mass concentrations with all the chemical species.
+BC, K and Mg were the most significant predictors in the PM~2.5~ model.
+The current study also sets the final concentration and uncertainty values needed to perform source apportionment using the PMF algorithm developed by the EPA.
 
